### 2011-09-28 paper/paper.tex
getting back into things -- started to remove caching, added paper stub

\documentclass[preprint]{sigplanconf}

\begin{document}

\authorinfo{}{}{}
\title{Reagents}
\maketitle

\begin{abstract}
\end{abstract}

\bibliographystyle{abbrvnat}

\end{document}

### 2011-09-28 src/main/scala/EliminationStack.scala

@@ -10,12 +10,14 @@ final class EliminationStack[A >: Null] {
   val tryPop: Reagent[Unit,Option[A]] = stack.tryPop <+> elimPop.map(Some(_))
   def pop: Reagent[Unit,A] = stack.pop <+> elimPop
 
+/*
   val dpush: Reagent[A,Int] = 
     stack.push.map(_ => 0) <+> 
     elimPush.map(_ => 1)
   val dtryPop: Reagent[Unit,Int] = 
     stack.tryPop.map(_ => 0) <+> 
     elimPop.map(_ => 1)
+*/
 
 /*
   val dpush: Reagent[A,Int] = 

### 2011-09-28 src/main/scala/Reagent.scala

@@ -7,25 +7,17 @@ import java.util.concurrent.locks._
 import chemistry.Util.Implicits._
 
 private[chemistry] sealed abstract class BacktrackCommand
-private[chemistry] case object Blocked extends BacktrackCommand 
-private[chemistry] abstract class Retry extends BacktrackCommand 
-private[chemistry] case object RetryUncached extends Retry
-
-private object OfferFail extends Exception
+private[chemistry] case object Block extends BacktrackCommand 
+private[chemistry] case object Retry extends BacktrackCommand 
 
 abstract class Reagent[-A, +B] {
-  private[chemistry] type Cache >: Null <: Retry
-  private[chemistry] def useCache: Boolean
   // returns either a BacktrackCommand or a B
-  private[chemistry] def tryReact(a: A, rx: Reaction, cache: Cache): Any
+  private[chemistry] def tryReact(a: A, rx: Reaction): Any
   protected def makeOfferI(a: A, offer: Offer[B]): Unit
   protected def composeI[C](next: Reagent[B,C]): Reagent[A,C]
   private[chemistry] def alwaysCommits: Boolean
   private[chemistry] def maySync: Boolean
   private[chemistry] def snoop(a: A): Boolean
-
-  @inline private[chemistry] final def convCache(r: Retry): Cache = 
-    if (r == RetryUncached) null else r.asInstanceOf[Cache]
 
   @inline private[chemistry] final def makeOffer(a: A, offer: Offer[B]) {
     // abort early if offer has already been consumed
@@ -64,7 +56,7 @@ abstract class Reagent[-A, +B] {
     throw Util.Impossible
   }
 
-  private[chemistry] final def withBackoff(a: A, cache: Cache): B = {
+  private[chemistry] final def withBackoff(a: A): B = {
     val backoff = new Backoff
     val doOffer = maySync
     var waiter: Waiter[B] = null      
@@ -82,7 +74,7 @@ abstract class Reagent[-A, +B] {
 	}
       } else backoff.once
       
-      tryReact(a, Inert, cache) match {
+      tryReact(a, Inert) match {
 	case (_: Retry) => loop
 	case Blocked    => block
 	case ans        => ans.asInstanceOf[B]
@@ -92,19 +84,19 @@ abstract class Reagent[-A, +B] {
   }
 
[…]

### 2011-09-28 src/main/scala/Ref.scala

@@ -33,27 +33,21 @@ final class Ref[A <: AnyRef](init: A) {
   // really, the type of data should belong to Reaction
 //  private[chemistry] val data = new AtomicReference[AnyRef](init)
 //  @inline private def get: A = Reaction.read(data).asInstanceOf[A]
-  private[chemistry] val data = new AtomicReference[AnyRef](init)
+  private[chemistry] val data = new AtomicReference[A](init)
   // ultimately, Reaction.read(data)
-  @inline private def get = data.get.asInstanceOf[A]
+  @inline private def get = data.get // .asInstanceOf[A]
 
   @inline def read: Reagent[Unit,A] = new AutoCont[Unit,A] {
-    def retValue(u: Unit): Any = Reaction.read(data)
+    def retValue(u: Unit): Any = get //Reaction.read(data)
   }   
 
   private final case class CAS[B](expect: A, update: A, k: Reagent[Unit,B]) 
 		extends Reagent[Unit, B] {
-    type Cache = k.Cache
-    def useCache = k.useCache
-    def tryReact(u: Unit, rx: Reaction, cache: Cache): Any = 
-      throw Util.Impossible
-/*
+    def tryReact(u: Unit, rx: Reaction): Any = 
       Ref.rxWithCAS(rx, data, expect, update, k) match {
-	case (newRx: Reaction) =>
-	  k.tryReact((), newRx, cache)
-	case ow => ow
+	case null  => Retry
+	case newRx => k.tryReact((), newRx)
       }
-*/
     def makeOfferI(u: Unit, offer: Offer[B]) =
       k.makeOffer(u, offer)
     def snoop(u: Unit) = false
@@ -101,17 +95,19 @@ final class Ref[A <: AnyRef](init: A) {
     }
 
     @inline final def tryReact(b: B, rx: Reaction, cache: Cache): Any = {
-      if (rx.casCount == 0 && k.alwaysCommits) {
+//      if (rx.casCount == 0 && k.alwaysCommits) {
 	val ov = get
 	val nv = newValue(ov, b)
 	if (data.compareAndSet(ov, nv))
 	  return k.tryReact(retValue(ov, b), rx, cache)
 	RetryUncached
+/*
       } else {
 	val ov = get
 	val nv = newValue(ov, b)
 	k.tryReact(retValue(ov, b), rx.withCAS(data, ov, nv), cache)
       }
+*/
     }
[…]

### 2011-09-28 src/main/scala/TreiberStack.scala

@@ -37,6 +37,7 @@ final class TreiberStack[A >: Null] {
   }
 */
 
+/*
   case class Node(var data: A, var next: Node) extends Retry
   val head = Ref[Node](null)
 
@@ -67,33 +68,64 @@ final class TreiberStack[A >: Null] {
   def pop: Reagent[Unit,A] = head.upd[A] {
     case Node(x,xs) => (xs, x)
   }
+*/
 
-/*
-  private val headX = new AtomicReference[List[A]](List())
+  case class Node(var data: A, var next: Node) extends Retry
+  val head = new AtomicReference[Node](null)
 
   object push extends Reagent[A,Unit] {
-    def tryReact(x:A, rx: Reaction): Any = {
-      val xs = headX.get
-      if (headX.compareAndSet(xs,x::xs)) () else ShouldRetry
+    type Cache = Node
+    def useCache = true
+    def tryReact(x:A, rx: Reaction, cache: Node): Any = {
+      val cached: Node = if (cache == null) Node(x, null) else cache
+      cached.next = head.get
+      if (head.compareAndSet(cached.next,cached)) () 
+      else cached
     }
     def makeOfferI(x:A, offer: Offer[Unit]) {}
     def composeI[B](next: Reagent[Unit,B]) = throw Util.Impossible
     def maySync = false
     def alwaysCommits = false
+    @inline def fastReact(x: A): Unit = {
+      val cached = Node(x, null)
+      while (true) {
+	cached.next = head.get
+	if (head.compareAndSet(cached.next, cached)) return ()
+      }
+      throw Util.Impossible      
+    }
+    def snoop(a: A) = false
   }
 
   object tryPop extends Reagent[Unit,Option[A]] {
-    def tryReact(u:Unit, rx: Reaction): Any = headX.get match {
-      case (ov@(x::xs)) => 
-	if (headX.compareAndSet(ov,xs)) Some(x) else ShouldRetry
-      case emp     => None
+    type Cache = Retry
+    def useCache = false
+    @inline def tryReact(u:Unit, rx: Reaction, cache: Retry): Any = 
+      head.get match {
+	case null => None
+	case n@Node(x, xs) =>
+	  if (head.compareAndSet(n,xs)) Some(x) else RetryUncached
+	case emp     => None
+      }
+    @inline def fastReact(u: Unit): Option[A] = {
+      while (true) {
+	tryReact(u, Inert, null) match {
[…]

### 2011-09-28 src/main/scala/bench/Results.scala

@@ -11,8 +11,8 @@ case class Measurement(
   trials: Int,
   cores: Int
 ) {
-  def format = " %8.2f |".format(throughput)
-  def formatR = " %8.2f |".format(rawThroughput)
+  def format = " %4.1f |".format(throughput)
+  def formatR = " %4.1f |".format(rawThroughput)
   def formatN(compTo: Double) = 
     " %8.2f |".format(rawThroughput / compTo)
 }
@@ -31,9 +31,9 @@ case class EntryResult(name: String, ms: Seq[Measurement]) {
   def formatN(compTo: Double) = 
     "%16.16s |".format(name ++ "(N)") ++
     ms.map(_.formatN(compTo)).mkString ++
-    "  max cov: %5.2f  avg cov: %5.2f\n".format(
-      ms.map(_.coeffOfVar).max,
-      ms.map(_.coeffOfVar).sum / ms.length
+    " mcov: %5.2f\n".format(
+      ms.map(_.coeffOfVar).max
+//      ms.map(_.coeffOfVar).sum / ms.length
     )
 }
 
@@ -45,13 +45,13 @@ case class BenchResult(name: String, work: Int, es: Seq[EntryResult]) {
   val columns = es(0).ms.length
 
   private def hrule {
-    println("-" * (18 + columns * 11))
+    println("-" * (18 + columns * 7))
   }
   def display {
     // header
     hrule
     print("%10.10s %5d |".format(name, work))
-    (1 to columns).map(i => print(" %8d |".format(i)))
+    (1 to columns).map(i => print(" %4d |".format(i)))
     println("")    
     hrule
 

### 2011-09-28 src/main/scala/bench/benchmarks/PushPop.scala

@@ -115,6 +115,13 @@ object PushPop extends Benchmark {
     def push(s: S, x: AnyRef) = s.push(x)
     def tryPop(s: S): Option[AnyRef]  = s.tryPop
   } 
+  private object handAloc extends Entry with Generic{
+    def name = "handAloc"
+    type S = HandStackAloc[AnyRef]
+    def setup = new HandStackAloc()
+    def push(s: S, x: AnyRef) = s.push(x)
+    def tryPop(s: S): Option[AnyRef]  = s.tryPop
+  } 
   private object rTreiber extends Entry with Generic {
     def name = "rTreiber"
     type S = TreiberStack[AnyRef]
@@ -136,6 +143,7 @@ private object rTreiberFast extends Entry with Generic {
     def push(s: S, x: AnyRef) = s.push ! x
     def tryPop(s: S): Option[AnyRef] =  s.tryPop ! ()    
   }
+/*
   private object rElimStat extends Entry with GenericCounting {
     def name = "rElimStat"
     type S = EliminationStack[AnyRef]
@@ -143,6 +151,7 @@ private object rTreiberFast extends Entry with Generic {
     def push(s: S, x: AnyRef): Int = s.dpush ! x
     def tryPop(s: S): Int =  s.dtryPop ! ()    
   }
+*/
 /*
   private object ebstack extends Entry {
     def name = "ebstack"
@@ -169,6 +178,5 @@ private object rTreiberFast extends Entry with Generic {
 //  def entries = List(rTreiber, rElim, hand)
 //  def entries = List(rElim, rTreiber, handElim, hand, handPool)
 //  def entries: List[Entry] = List(rElim, handPool, handElim)
-  def entries: List[Entry] = List(rTreiber, rTreiberFast, rElim, 
-				  hand, handPool, handElim)
+  def entries: List[Entry] = List(handAloc, hand)
 }

### 2011-09-28 src/main/scala/bench/competition/HandStackAloc.scala

@@ -1,0 +1,42 @@
+package chemistry.bench.competition
+
+import java.util.concurrent.atomic._
+import scala.annotation.tailrec
+
+// Standard lock-free stack, due to Treiber.  Doesn't yet perform
+// exponential backoff.
+class HandStackAloc[A >: Null] {
+  class Node(val data: A, var next: Node) 
+
+  // head always points to top of stack,
+  //   from which the rest of the stack is reachable
+  private val head = new AtomicReference[Node](null)
+
+  def push(x: A) {
+    while (true) {
+      val n = new Node(x, head.get)
+      if (head compareAndSet (n.next, n)) return
+    } 
+  }
+
+  def tryPop(): Option[A] = {
+    while (true) {
+      val h = head.get
+      if (h eq null) 
+	return None
+      else if (head compareAndSet (h, h.next)) 
+	return Some(h.data) 
+    }
+    throw new Exception("Impossible")
+  } 
+
+  // definitely possible to do this all at once with a CAS on head
+  def popAll(): List[A] = {
+    @tailrec def grabAll(acc: List[A]): List[A] = 
+      tryPop match {
+	case None => acc
+	case Some(x) => grabAll(x :: acc)
+      }
+    grabAll(List()).reverse
+  }
+}

### 2011-09-29 src/main/scala/Channels.scala
more work undoing caching and cleaning

@@ -24,17 +24,20 @@ final private class RMessage[A,B,C](
 ) extends Message[A,B] {
   private case class CompleteExchange[D](kk: Reagent[A,D]) 
 	       extends Reagent[C,D] {
-    type Cache = Retry
-    def useCache = false
-    def tryReact(c: C, rx: Reaction, cache: Cache): Any = {
+    def tryReact(c: C, rx: Reaction): Any = {
+      if (rx.canCASImmediate(kk)) {
+	if (
+      } else {
+      }
+
       val newRX = Ref.rxWithCAS(rx, waiter.status, 
 				Waiter.Waiting, c.asInstanceOf[AnyRef], kk)
       if (newRX == null) 
-	RetryUncached
+	Retry
       else if (waiter.blocking)
-	kk.tryReact(m, newRX.withPostCommit((_:Unit) => waiter.wake), null)
+	kk.tryReact(m, newRX.withPostCommit((_:Unit) => waiter.wake))
       else
-	kk.tryReact(m, newRX, null)
+	kk.tryReact(m, newRX)
     }
     def makeOfferI(c: C, offer: Offer[D]) = throw Util.Impossible
     def composeI[E](next: Reagent[D,E]): Reagent[C,E] =
@@ -53,17 +56,14 @@ private final case class Endpoint[A,B,C](
   incoming: Pool[Message[B,A]],
   k: Reagent[B,C]
 ) extends Reagent[A,C] {
-  type Cache = Retry
-  def useCache = false
-
-  def tryReact(a: A, rx: Reaction, cache: Cache): Any = {
+  def tryReact(a: A, rx: Reaction): Any = {
     @tailrec def tryFrom(n: incoming.Node, retry: Boolean): Any = 
       if (n == null) {
-	if (retry) RetryUncached else Blocked
-      } else n.data.exchange.compose(k).tryReact(a, rx, null) match {
-	case (_: Retry) => tryFrom(n.next, true)
-	case Blocked    => tryFrom(n.next, retry)
-	case ans        => ans
+	if (retry) Retry else Block
+      } else n.data.exchange.compose(k).tryReact(a, rx) match {
+	case Retry => tryFrom(n.next, true)
+	case Block => tryFrom(n.next, retry)
+	case ans   => ans
       }
     tryFrom(incoming.cursor, false)
   }

### 2011-09-29 src/main/scala/Reaction.scala

@@ -11,6 +11,11 @@ private abstract sealed class Reaction {
 
   def casCount: Int
 
+  // is it safe to do a CAS *while creating the reaction*?  generally, this is
+  // fine as long as the whole reaction is guaranteed to be a 1-cas.
+  def canCASImmediate[A,B](k: Reagent[A,B]): Boolean = 
+    casCount == 0 && k.alwaysCommits
+
   def withPostCommit(postCommit: Unit => Unit): Reaction =
     PostCommit(postCommit, this)
   def withCAS(ref: AtomicReference[AnyRef], ov: AnyRef, nv: AnyRef): Reaction =

### 2011-09-29 src/main/scala/Reagent.scala

@@ -75,22 +75,22 @@ abstract class Reagent[-A, +B] {
       } else backoff.once
       
       tryReact(a, Inert) match {
-	case (_: Retry) => loop
-	case Blocked    => block
-	case ans        => ans.asInstanceOf[B]
+	case Retry => loop
+	case Block => block
+	case ans   => ans.asInstanceOf[B]
       }
     }
     loop
   }
 
-  final def !(a: A): B = tryReact(a, Inert, null) match {
+  final def !(a: A): B = tryReact(a, Inert) match {
     case Retry => withBackoff(a)
     case Block => block
     case ans   => ans.asInstanceOf[B]
   }
 
   @inline final def !?(a:A) : Option[B] = {
-    tryReact(a, Inert, null) match {
+    tryReact(a, Inert) match {
       case Retry => None // should we actually retry here?  if we do, more
 			 // informative: a failed attempt entails a
 			 // linearization where no match was possible.  but
@@ -200,7 +200,7 @@ object computed {
 object lift {
   @inline def apply[A,B](f: PartialFunction[A,B]): Reagent[A,B] = 
     new AutoCont[A,B] {
-      def retValue(a: A): Any = if (f.isDefinedAt(a)) f(a) else Blocked
+      def retValue(a: A): Any = if (f.isDefinedAt(a)) f(a) else Block
     }
 }
 

### 2011-09-29 src/main/scala/Ref.scala

@@ -3,7 +3,6 @@
 package chemistry
 
 import java.util.concurrent.atomic._
-import scala.annotation.tailrec
 
 /*
 private final class PaddedAtomicReference[A](init:A) 
@@ -29,41 +28,45 @@ private final class PaddedAtomicReference[A](init:A)
 final class Ref[A <: AnyRef](init: A) {
 //  private val waiters = new MSQueue[]()
 
-  //  private[chemistry] val data = new PaddedAtomicReference[AnyRef](init)
   // really, the type of data should belong to Reaction
-//  private[chemistry] val data = new AtomicReference[AnyRef](init)
-//  @inline private def get: A = Reaction.read(data).asInstanceOf[A]
-  private[chemistry] val data = new AtomicReference[A](init)
-  // ultimately, Reaction.read(data)
-  @inline private def get = data.get // .asInstanceOf[A]
+  private[chemistry] val data = new AtomicReference[AnyRef](init)
+//  private[chemistry] val data = new PaddedAtomicReference[AnyRef](init)
+  private def get: A = Reaction.read(data).asInstanceOf[A]
+
+  @inline private def cas(ov: A, nv: A): Boolean =
+    data.compareAndSet(ov, nv)
 
   @inline def read: Reagent[Unit,A] = new AutoCont[Unit,A] {
-    def retValue(u: Unit): Any = get //Reaction.read(data)
-  }   
+    def retValue(u: Unit): Any = get
+  }
 
   private final case class CAS[B](expect: A, update: A, k: Reagent[Unit,B]) 
 		extends Reagent[Unit, B] {
     def tryReact(u: Unit, rx: Reaction): Any = 
-      Ref.rxWithCAS(rx, data, expect, update, k) match {
-	case null  => Retry
-	case newRx => k.tryReact((), newRx)
-      }
+      Ref.continueWithCAS(data, rx, (), k, expect, update)
     def makeOfferI(u: Unit, offer: Offer[B]) =
       k.makeOffer(u, offer)
-    def snoop(u: Unit) = false
     def composeI[C](next: Reagent[B,C]) = CAS(expect, update, k.compose(next))
     def maySync = k.maySync
     def alwaysCommits = false
[…]

### 2011-09-30 src/main/scala/Channels.scala
approaching compilable state

@@ -25,16 +25,17 @@ final private class RMessage[A,B,C](
   private case class CompleteExchange[D](kk: Reagent[A,D]) 
 	       extends Reagent[C,D] {
     def tryReact(c: C, rx: Reaction): Any = {
-      if (rx.canCASImmediate(kk)) {
-	if (
-      } else {
-      }
+      val ov = Waiter.Waiting
+      val nv = c.asInstanceOf[AnyRef]
 
-      val newRX = Ref.rxWithCAS(rx, waiter.status, 
-				Waiter.Waiting, c.asInstanceOf[AnyRef], kk)
-      if (newRX == null) 
-	Retry
-      else if (waiter.blocking)
+      val newRX = 
+	if (rx.canCASImmediate(kk)) {
+	  if (!waiter.status.compareAndSet(ov, nv)) // attempt early
+	    return Retry			    // escape early
+	  else rx				    
+	} else rx.withCAS(waiter.status, ov, nv)
+
+      if (waiter.blocking)
 	kk.tryReact(m, newRX.withPostCommit((_:Unit) => waiter.wake))
       else
 	kk.tryReact(m, newRX)

### 2011-09-30 src/main/scala/Ref.scala

@@ -33,9 +33,6 @@ final class Ref[A <: AnyRef](init: A) {
 //  private[chemistry] val data = new PaddedAtomicReference[AnyRef](init)
   private def get: A = Reaction.read(data).asInstanceOf[A]
 
-  @inline private def cas(ov: A, nv: A): Boolean =
-    data.compareAndSet(ov, nv)
-
   @inline def read: Reagent[Unit,A] = new AutoCont[Unit,A] {
     def retValue(u: Unit): Any = get
   }
@@ -43,7 +40,12 @@ final class Ref[A <: AnyRef](init: A) {
   private final case class CAS[B](expect: A, update: A, k: Reagent[Unit,B]) 
 		extends Reagent[Unit, B] {
     def tryReact(u: Unit, rx: Reaction): Any = 
-      Ref.continueWithCAS(data, rx, (), k, expect, update)
+      if (rx.canCASImmediate(k)) {
+	if (data.compareAndSet(expect, update))
+	  k.tryReact((), rx)
+	else Retry
+      } else k.tryReact((), rx.withCAS(data, expect, update))
+
     def makeOfferI(u: Unit, offer: Offer[B]) =
       k.makeOffer(u, offer)
     def composeI[C](next: Reagent[B,C]) = CAS(expect, update, k.compose(next))
@@ -61,7 +63,9 @@ final class Ref[A <: AnyRef](init: A) {
 	// to the read as possible, to decrease chance of interference.
 	val ov = get
 	val (nv, ret) = f(ov, b)
-	if (cas(ov, nv)) k.tryReact(ret, rx) else Retry
+	if (data.compareAndSet(ov, nv)) 
+	  k.tryReact(ret, rx) 
+	else Retry
       } else {
 	val ov = get
 	val (nv, ret) = f(ov, b)
@@ -117,17 +121,18 @@ final class Ref[A <: AnyRef](init: A) {
     def retValue(a: A, b: B): C
     def retryValue(cur: A, lastAttempt: A, b: B): A = newValue(cur, b)
   }
-  abstract class FastUpd[B,C] extends InnerFastUpd[B,C,C](Commit[C]()) 
+  abstract class FastUpd[B,C] extends InnerFastUpd[B,C,C](Commit[C]())
+
+  
+ 
 }
 object upd {
[…]

### 2011-10-01 src/main/scala/Pool.scala
back in compilable state

@@ -188,6 +188,7 @@ final class CircularPool[A <: DeletionFlag] extends Pool[A] {
 */
 }
 
+/*
 private object BoundPool {
   val capacity = 32
   val full = math.max(0, math.min(capacity, Chemistry.procs / 2) - 1)
@@ -268,3 +269,4 @@ final class BoundPool[A <: DeletionFlag] {
     return null
   }
 }
+*/

### 2011-10-01 src/main/scala/Ref.scala

@@ -55,43 +55,14 @@ final class Ref[A <: AnyRef](init: A) {
   }
   @inline def cas(ov:A,nv:A): Reagent[Unit,Unit] = CAS(ov,nv,Commit[Unit]()) 
 
-  private final case class Upd[B,C,D](f: (A,B) => (A,C), k: Reagent[C,D]) 
-		     extends Reagent[B, D] {
-    def tryReact(b: B, rx: Reaction): Any = {
-      if (rx.canCASImmediate(k)) {
-	// this bit is repeated: we want the actual CAS to be as close
-	// to the read as possible, to decrease chance of interference.
-	val ov = get
-	val (nv, ret) = f(ov, b)
-	if (data.compareAndSet(ov, nv)) 
-	  k.tryReact(ret, rx) 
-	else Retry
-      } else {
-	val ov = get
-	val (nv, ret) = f(ov, b)
-	k.tryReact(ret, rx.withCAS(data, ov, nv))
-      }
-    }
-    def makeOfferI(b: B, offer: Offer[D]): Unit = {
-      val ov = get
-      val (_, ret) = f(ov, b)
-      k.makeOffer(ret, offer)
-    }
-    def composeI[E](next: Reagent[D,E]) = Upd(f, k.compose(next))
-    def maySync = k.maySync
-    def alwaysCommits = false
-    def snoop(b: B) = false
-  }
-  @inline def upd[B,C](f: (A,B) => (A,C)): Reagent[B, C] = 
-    Upd(f, Commit[C]())
-
-  abstract class InnerFastUpd[B,C,D] private[chemistry] (k: Reagent[C,D])
+  abstract class InnerUpd[B,C,D] private[chemistry] (k: Reagent[C,D])
 	   extends Reagent[B, D] {
     def tryReact(b: B, rx: Reaction): Any = {
       if (rx.canCASImmediate(k)) {
 	var tries = 3
 	while (tries > 0) {
 	  val ov = get
+	  if (!valid(ov,b)) return Retry
 	  val nv = newValue(ov, b)
 	  if (data.compareAndSet(ov, nv))
 	    return k.tryReact(retValue(ov, b), rx)
@@ -100,6 +71,7 @@ final class Ref[A <: AnyRef](init: A) {
 	Retry
       } else {
 	val ov = get
+	if (!valid(ov,b)) return Retry
 	val nv = newValue(ov, b)
[…]

### 2011-10-01 src/main/scala/TreiberStack.scala

@@ -37,24 +37,17 @@ final class TreiberStack[A >: Null] {
   }
 */
 
-/*
-  case class Node(var data: A, var next: Node) extends Retry
-  val head = Ref[Node](null)
 
-  // Reagent[A,Unit]
-  val push = new head.CachedUpd[A,Unit] {
-    type Cache = Node
-    @inline final def initCache: Node = Node(null, null)
-    @inline final def newValue(cur: Node, cache: Node, a: A): Node = {
-      cache.data = a
-      cache.next = cur
-      cache
-    } 
+  private case class Node(var data: A, var next: Node) 
+  private val head = Ref[Node](null)
+
+  val push: Reagent[A,Unit] = new head.Upd[A,Unit] {
+    @inline final def newValue(cur: Node, a: A): Node = 
+      Node(null, null)
     @inline final def retValue(cur: Node, a: A): Unit = ()
   }
 
-  // Reagent[Unit,Option[A]]
-  val tryPop = new head.FastUpd[Unit,Option[A]] {
+  val tryPop: Reagent[Unit,Option[A]] = new head.Upd[Unit,Option[A]] {
     @inline final def newValue(cur: Node, u: Unit): Node = cur match { 
       case null   => null
       case Node(x,xs) => xs
@@ -68,8 +61,8 @@ final class TreiberStack[A >: Null] {
   def pop: Reagent[Unit,A] = head.upd[A] {
     case Node(x,xs) => (xs, x)
   }
-*/
 
+/*
   case class Node(var data: A, var next: Node) extends Retry
   val head = new AtomicReference[Node](null)
 
@@ -124,7 +117,7 @@ final class TreiberStack[A >: Null] {
   }
 
   def pop: Reagent[Unit,A] = throw Util.Impossible
-
+*/
 /*
   object pop extends Reagent[Unit,A] {
     @tailrec def tryReact(u:Unit, rx: Reaction): Any = headX.get match {

### 2011-10-01 src/main/scala/bench/benchmarks/PushPop.scala

@@ -129,6 +129,7 @@ object PushPop extends Benchmark {
     def push(s: S, x: AnyRef) = s.push ! x
     def tryPop(s: S): Option[AnyRef]  = s.tryPop ! ()
   }
+/*
 private object rTreiberFast extends Entry with Generic {
     def name = "rTreiberFast"
     type S = TreiberStack[AnyRef]
@@ -136,6 +137,7 @@ private object rTreiberFast extends Entry with Generic {
     def push(s: S, x: AnyRef) = s.push.fastReact(x)
     def tryPop(s: S): Option[AnyRef]  = s.tryPop.fastReact()
   }
+*/
   private object rElim extends Entry with Generic {
     def name = "rElim"
     type S = EliminationStack[AnyRef]

### 2011-10-11 paper/paper.tex
some outlining for the paper

@@
 \documentclass[preprint]{sigplanconf}
+
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{mathpartir}
+
+\newcommand{\ra}{\rightarrow}
+\newcommand{\gives}{\vdash}
+\newcommand{\proves}{\vdash}
+\newcommand{\GA}{\ |\ }
+\newcommand{\kw}[1]{\textsf{#1}}
+\newcommand{\ty}[1]{\textrm{#1}\ }
+\newcommand{\bind}{\mbox{\ \scriptsize$>\!\!>\!=$\ }}
+\newcommand{\letrec}[2]{\kw{letrec}\ #1\ \kw{in}\ #2}
+\newcommand{\LET}[4]{\kw{let}_{#1}\ #2 = #3\ \kw{in}\ #4}
 
@@
 \authorinfo{}{}{}
-\title{Reagents}
+% yikes!
+\title{Reagents: expressing and composing patterns of fine-grained concurrency}
 \maketitle
@@
 \begin{abstract}
+blah blah
 \end{abstract}
+
+``Programs are what happens between cache misses.''
+
+As with most ..., concurrency only exacerbates performance problems
+with the memory hierarchy.
+
+ahmdahl's law -- even if you parallelize the big chunks, the
+fine-grained stuff will get you in the end
+
+%\section{Reagent semantics}
+
+%\begin{verbatim}
+
+%\end{verbatim}
 
@@
 \end{document}
+
+# General notes
+
+Imposes no cache-coherence overhead on isolated data structures
+
+Can argue that transactional-event-style liveness properties are
+important if you want general way of composing fine-grained concurrent
+algorithms with message-passing
+
+To what extent do reagents encompass STM?  who are we competing with?
+
+multiparadim, expressive, fast -- all that's great, but sales pitch
+needs to be oriented around a customer with specific needs.  any
+reason not to use the pitch of extensible, composable j.u.c.?
+
+# Contributions
+
+  - make it easier to express fine-grained concurrent algorithms,
+  including blocking operations etc.  make it possible to build them
+  compositionally, which already happens in the literature but is
+  largely unremarked upon.
+
+  - we capture exponential backoff, elimination backoff, and other
+  such patterns once and for all
+
+  - can use flat combining as illustration of capturing very general
+  concurrency abstraction
+
+  - smooth blend of message-passing and shared-state concurrency
+  isolation for shared state, coherent communication through messages,
+  uniform blocking/synchronization
+
+  - operational semantics
+
+  - yields completely lock-free implementation of core CML
+
+  - generalizes join calculus implementation, allowing dissolution
+  even for existing channels
+
+  - generalizes transactions events, but with much better
+  implementation
+
+# To do
+
+  - benchmark against STM (possibly Akka Transactors?)
+
+  - benchmark queues, including flat combining and buckets
+
+# Outline
+
+2.5 Title, abstract, intro, library overview 
+1.5 Semantics 
+2.5 Implementation/algorithms 
+1.0 Performance results
+1.5 Related work
+1.0 Conclusion and bib
+
+How much time should we spend covering scalable concurrency
+background?  This is perhaps tied to strategies for introducing the
+library constructs: the primitives can be shown by-need for examples.
+Should consult good papers like Ryan's and Jesse's for examples here.
+
+# Semantics
+
+Need to nail down story for overlapping state access.  In particular,
+when (if ever) is an error flagged?  Unfortunately, talking about heap
+unsplittability is rather more complicated than splittability.

### 2011-10-11 src/main/scala/Channels.scala
starting to think about chains of message sync, 'multithreaded' reagents

@@ -35,6 +35,7 @@ private final case class Endpoint[A,B,C](
   k: Reagent[B,C]
 ) extends Reagent[A,C] {
   def tryReact(a: A, rx: Reaction, offer: Offer[C]): Any = {
+    // todo: avoid attempts of claiming the same message twice
     @tailrec def tryFrom(n: incoming.Node, retry: Boolean): Any = 
       if (n == null) {
 	if (retry) Retry else Block
@@ -50,7 +51,10 @@ private final case class Endpoint[A,B,C](
     if (offer != null) outgoing.put(new Message(a, rx, k, offer))
 
     // now attempt an immediate reaction
-    tryFrom(incoming.cursor, false)
+    tryFrom(incoming.cursor, false) match {
+      case Block => Block // todo: fall back on "multithreaded" reagents
+      case ow => ow
+    }
   }
   def snoop(a: A): Boolean = incoming.snoop
   @inline def composeI[D](next: Reagent[C,D]) = 
@@ -71,3 +75,6 @@ object Chan {
   @inline def apply[A](): (Reagent[Unit,A], Reagent[A,Unit]) =
     SwapChan[Unit,A]()
 }
+
+// is there a possibility of building the exchanger directly here?
+// also, can the endpoint logic be decomposed/parameterized more?

### 2011-10-11 src/main/scala/Reaction.scala

@@ -4,8 +4,10 @@
 package chemistry
 
 private sealed class Reaction private (
-  val casList: List[CAS[_]], 
-  val pcList: List[Unit => Unit]) {
+  val casList: List[CAS[_]],		// k-cas built up so far
+  val pcList: List[Unit => Unit],	// post-commit actions
+  val msgSet: HashSet[Message[_,_,_]]   // messages intended for consumption
+) {
   import Reaction._
   
   def casCount: Int = casList.size

### 2011-10-11 src/main/scala/Reagent.scala

@@ -123,7 +123,7 @@ private abstract class AutoContImpl[A,B,C](val k: Reagent[B, C])
       override def newRx(a: A, rx: Reaction): Reaction = 
 	AutoContImpl.this.newRx(a, rx)
     }
-  final def alwaysCommits = k.alwaysCommits
+  final def alwaysCommits = k.alwaysCommits // this needs to be overridable!
   final def maySync = k.maySync
 }
 private abstract class AutoCont[A,B] extends AutoContImpl[A,B,B](Commit[B]())
@@ -183,6 +183,7 @@ object computed {
 }
 
 object lift {
+  // this is WRONG -- does NOT always commit
   @inline def apply[A,B](f: PartialFunction[A,B]): Reagent[A,B] = 
     new AutoCont[A,B] {
       def retValue(a: A): Any = if (f.isDefinedAt(a)) f(a) else Block

### 2011-10-12 paper/paper.tex
work on paper intro

@@
 
-``Programs are what happens between cache misses.''
+\section{Introduction}
 
-As with most ..., concurrency only exacerbates performance problems
-with the memory hierarchy.
+\hspace{\stretch{1}}\emph{Programs are what happens between cache
+  misses}.\footnote{Folklore; origin unknown.}
 
-ahmdahl's law -- even if you parallelize the big chunks, the
-fine-grained stuff will get you in the end
+\subsection*{The problem}
 
-%\section{Reagent semantics}
+Ahmdahl's law tells us that sequential bottlenecks fundamentally limit
+profit from parallelism~\cite{?}.  In practice, the effect is made
+fatal by a silent killer: interprocessor communication, often in the
+form of cache coherence.  When one thread waits on another, the
+program pays the cost of lost parallelism \emph{and} an extra cache
+miss.  Extra misses can easily add up to parallel \emph{slowdown},
+more than negating the benefits of the remaining parallelism.  Cache,
+as ever, is king.
 
-%\begin{verbatim}
+The easy answer is: avoid communication.  In other words, parallelize
+at a coarse grain, giving threads large chunks of independent work.
+That strategy only goes so far.  Some work doesn't easily factor into
+large chunks, or equal size chunks.  Fine-grained parallelism is
+easier to find, and easier to sprinkle throughout existing sequential
+code.  With increasing pressure to parallelize, fine-grained
+parallelism will only become more common.
 
-%\end{verbatim}
+In the past two decades,\footnote{Herlihy/Wing, Mellor-Crummey/Scott}
+an enormous literature has emerged
+
+\subsection*{Our contribution}
+
+\section{Reagents}
+
+\section{Semantics}
+
+\section{Implemention}
+
+\section{Performance}
+
+\section{Related work}
 
@@
 \end{document}
+
+# Intro notes
+
+What do we want to accomplish?
+
+ - motivate libraries like j.u.c
+ - downsides of such libraries
+   - huge research & impl effort
+   - not easily extensible
+   - internal structure has repeated, but unabstracted patterns
+ - our contribution
+   - a new way of expressing scalable algorithms that
+     - doesn't lose much performance 
+     - allows composition
+     - is much nicer than direct programming
+     - also incorporates blocking behavior
 

### 2011-10-12 paper/paper.tex
continued work on introduction

@@
 
-Ahmdahl's law tells us that sequential bottlenecks fundamentally limit
-profit from parallelism~\cite{?}.  In practice, the effect is made
-fatal by a silent killer: interprocessor communication, often in the
-form of cache coherence.  When one thread waits on another, the
-program pays the cost of lost parallelism \emph{and} an extra cache
-miss.  Extra misses can easily add up to parallel \emph{slowdown},
-more than negating the benefits of the remaining parallelism.  Cache,
-as ever, is king.
+Ahmdahl's law tells us that sequential bottlenecks fundamentally limit our
+profit from parallelism~\cite{?}.  In practice, the effect is amplified by
+another factor: interprocessor communication, often in the form of cache
+coherence.  When one thread waits on another, the program pays the cost of
+lost parallelism \emph{and} an extra cache miss.  The extra misses can easily
+yield parallel \emph{slowdown}, more than negating the benefits of the
+remaining parallelism.
 
-The easy answer is: avoid communication.  In other words, parallelize
-at a coarse grain, giving threads large chunks of independent work.
-That strategy only goes so far.  Some work doesn't easily factor into
-large chunks, or equal size chunks.  Fine-grained parallelism is
-easier to find, and easier to sprinkle throughout existing sequential
-code.  With increasing pressure to parallelize, fine-grained
-parallelism will only become more common.
+Cache, as ever, is king.
 
-In the past two decades,\footnote{Herlihy/Wing, Mellor-Crummey/Scott}
-an enormous literature has emerged
+The easy answer is: avoid communication.  In other words, parallelize at a
+coarse grain, giving threads large chunks of independent work.  But some work
+doesn't easily factor into large chunks, or equal-size chunks.  Fine-grained
+parallelism is easier to find, and easier to sprinkle throughout existing
+sequential code.
+%With increasing pressure to parallelize, fine-grained parallelism will only
+%become more common.
+
+Another answer is: communicate efficiently.  The past two
+decades\footnote{Herlihy/Wing, Mellor-Crummey/Scott} have produced a sizable
+collection of algorithms for synchronization, communication, and shared
+storage which minimize the use of memory bandwidth and avoid unnecessary
+waiting.  This research effort has led to industrial-strength
+libraries---\texttt{java.util.concurrent} (JUC) the most prominent---offering
+a wide range of concurrency primitives appropriate for fine-grained
+parallelism.
+
+Such libraries are an enormous undertaking---and one that must be repeated for
+new platforms.  They tend to be conservative, implementing only those data
+structures and primitives likely to fulfill common needs.  In addition, it is
+generally not possible to safely combine the facilities of the library.  For
+example, JUC provides queues, sets and maps, but not stacks or bags.  Its
+queues come in both blocking and nonblocking forms, while its sets and maps
+are nonblocking only.  Although the queues provide atomic (thread-safe)
+dequeuing and sets provide atomic insertion, it is not possible to combine
+these into a single atomic operation.
+
+In short, libraries for fine-grained concurrency are indispensable, but hard
+to write, hard to extend by composition, and hard to tailor to the exact needs
+of a particular user.
 
@@
 
+We have developed a core library of \emph{composable} primitives for
+fine-grained concurrency.
+
+by no means a silver bullet.
+
+why is this different from STM
+
 \section{Reagents}
+\label{sec:reagents}
 
 \section{Semantics}
+\label{sec:semantics}
 
-\section{Implemention}
+\section{Implementation}
+\label{sec:implementation}
 
 \section{Performance}
+\label{sec:performance}
 
 \section{Related work}
+\label{sec:related}
 

### 2011-10-13 paper/paper.tex
added some macros for semantics
progress on paper

@@
 
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{mathpartir}
+\newcommand{\elide}[1]{}
 
-\newcommand{\ra}{\rightarrow}
-\newcommand{\gives}{\vdash}
-\newcommand{\proves}{\vdash}
-\newcommand{\GA}{\ |\ }
-\newcommand{\kw}[1]{\textsf{#1}}
-\newcommand{\ty}[1]{\textrm{#1}\ }
-\newcommand{\bind}{\mbox{\ \scriptsize$>\!\!>\!=$\ }}
-\newcommand{\letrec}[2]{\kw{letrec}\ #1\ \kw{in}\ #2}
-\newcommand{\LET}[4]{\kw{let}_{#1}\ #2 = #3\ \kw{in}\ #4}
+\input{packages}
 
@@
 % yikes!
-\title{Reagents: expressing and composing patterns of fine-grained concurrency}
+%\title{Reagents: expressing and composing patterns of fine-grained concurrency}
+\title{Reagents: expressing and composing fine-grained concurrency}
 \maketitle
@@
 \begin{abstract}
+blah blah\\
+blah blah\\
+blah blah\\
+blah blah\\
+blah blah\\
 blah blah
@@
 dequeuing and sets provide atomic insertion, it is not possible to combine
-these into a single atomic operation.
+these into a single atomic operation that moves an element from a queue into a
+set.
 
 In short, libraries for fine-grained concurrency are indispensable, but hard
-to write, hard to extend by composition, and hard to tailor to the exact needs
-of a particular user.
+to write, hard to extend by composition, and hard to tailor to the needs of
+particular users.
 
@@
 
-We have developed a core library of \emph{composable} primitives for
-fine-grained concurrency.
+% first class synchronous operations
+
+%% We have developed a small set of primitives for fine-grained concurrency that
+%% are \emph{expressive} and \emph{composable}:
+
+%% We have developed \emph{reagents}, a new abstraction for fine-grained
+%% concurrency that is \emph{expressive} and \emph{composable}:
+
+%% We have developed \emph{reagents}, which are first-class representations of
+%% fine-grained concurrency primitives.  Reagents are \emph{expressive} and
+%% \emph{composable}:
+
+We have developed \emph{reagents}, which abstractly represent fine-grained
+concurrent operations.  Reagents are \emph{expressive} and \emph{composable},
+but retain the scalability and performance of existing algorithms:
+
+%We have developed \emph{reagents},
+
+%% compositional primitives for fine-grained concurrency
+
+%% a small set of primitives for fine-grained concurrency that
+%% are \emph{expressive} and \emph{composable}:
+
+\paragraph{Expressive}  .
+
+\paragraph{Composable} .
+
+We see this work as serving the needs of two distinct groups: concurrency
+experts and concurrency users.  Using reagents, experts can write algorithms
+more easily, because common patterns are expressible as abstractions and many
+are built-in.  Users can compose, extend, and tailor the library without
+detailed knowledge of the algorithms involved.
+
+blend shared-state and message-passing concurrency
+
+\elide{
+# Contributions
+
+  - make it easier to express fine-grained concurrent algorithms,
+  including blocking operations etc.  make it possible to build them
+  compositionally, which already happens in the literature but is
+  largely unremarked upon.
+
+  - we capture exponential backoff, elimination backoff, and other
+  such patterns once and for all
+
+  - can use flat combining as illustration of capturing very general
+  concurrency abstraction
+
+  - smooth blend of message-passing and shared-state concurrency
+  isolation for shared state, coherent communication through messages,
+  uniform blocking/synchronization
+
+  - operational semantics
+
+  - yields completely lock-free implementation of core CML
+
+  - generalizes join calculus implementation, allowing dissolution
+  even for existing channels
+
+  - generalizes transactional events, but with more specialized
+  implementation (doesn't need an underlying STM)
 
@@
 why is this different from STM
+}
+
+\section{Background}
+
+Reagents are directly motivated by
+
+Before introducing reagents, it will be helpful to
+
+readers familiar with these algorithms can safely jump to sec:reagents
+
+also background on Scala
+
+traditional code for Treiber stack (with exp. backoff),
+
+\begin{lstlisting}
+class TreiberStack[A] {
+  private val head = new AtomicReference[List[A]](Nil)
+  def push(a: A) {
+    val backoff = new Backoff
+    while (true) {
+      val cur = head.get()
+      if (head.cas(cur, a :: cur)) return
+      backoff.once()
+    }
+  }
+  def tryPop(): Option[A] = {
+    val backoff = new Backoff
+    while (true) {
+      val cur = head.get() 
+      cur match {
+        case Nil     => return None
+        case a::tail => if (head.cas(cur, tail)) return a
+      }
+      backoff.once()
+    }
+  }
+}
+\end{lstlisting}
+
+need to explain pattern matching and PartialFunction
+
+talk about elimination backoff and blocking versions but don't show code.
+
+discuss the insight behind MCS locks
 
@@
 
+First we give an overview of the library, and then we illustrate its
+expressiveness and composability by showing how to implement several
+
+chemistry metaphor
+
+lack of isolation needed to express many lock-free algorithms.
+message-passing turns out to be a good way to understand what's going on.
+
+\begin{lstlisting}
+class TreiberStack[A] {
+  private val head = new Ref[List[A]](Nil)
+  val push: Reagent[A, Unit] = head.upd {
+    (xs, x) => (x:xs, ())
+  }
+  val tryPop: Reagent[Unit, Option[A]] = head.upd {
+    case (x::xs, ()) => (xs, Some(x))
+    case (Nil,   ()) => (Nil, None)
+  }
+}
+\end{lstlisting}
+
+Examples:
+
+\begin{itemize}
+\item TreiberStack
+\item with blocking
+\item with elimination backoff
+\item MSQueue
+\item with blocking
+\item with buckets?
+\item Counter -> semaphore
+\item RWLocks
+\item Tree barriers
+\item Dining philosophers(ish)
+\item Flat combining
+\end{itemize}
+
 \section{Semantics}
@@
 
+\newcommand{\ra}{\rightarrow}
+\newcommand{\gives}{\vdash}
+\newcommand{\proves}{\vdash}
+\newcommand{\kw}[1]{\textsf{#1}}
+\newcommand{\GA}{\ |\ }
+\newcommand{\step}{\longrightarrow}
+\newcommand{\cstep}[1]{\stackrel{#1}{\Longrightarrow}}
+
+\newcommand{\evalCtx}{\mathbb{E}}
+\newcommand{\procCtx}{\mathbb{P}}
+
+\newcommand{\ret}[1]{\kw{ret}(#1)}
+\newcommand{\never}{\kw{never}}
+\newcommand{\postCommit}[1]{\kw{postCommit}(#1)}
+\newcommand{\computed}[1]{\kw{computed}(#1)}
+\newcommand{\compose}{{\small\texttt{>=>}}}
+\newcommand{\lift}[1]{\kw{lift}(#1)}
+\newcommand{\send}[1]{#1}
+\newcommand{\cas}[3]{\kw{cas}(#1,#2,#3)}
+\newcommand{\rd}[1]{\kw{read}(#1)}
+\newcommand{\newChan}{\kw{new Chan}}
+\newcommand{\newRef}[1]{\kw{new Ref(#1)}}
+
+\elide{
+\[
+\infer
+  {P_i, \sigma_i \cstep{t_i} P'_i, \sigma'_i}
+  {P_1 | \cdots | P_n,\ \sigma_1 \uplus \cdots \uplus \sigma_n\ \step\
+   P'_1 | \cdots | P'_n,\ \sigma'_1 \uplus \cdots \uplus \sigma'_n
+  }
+\]
+
+\[
+\infer
+  {r_i!v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
+  {\procCtx[\prod \evalCtx_i[r_i ! v_i]],\ \biguplus \sigma_i \step
+   \procCtx[\prod \evalCtx_i[\ret{v'_i}]],\ \biguplus \sigma'_i}
+\]
+}
+
+\[
+\infer
+  {r_i!v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
+  {\prod \evalCtx_i[r_i ! v_i],\ \biguplus \sigma_i \step
+   \prod \evalCtx_i[\ret{v'_i}],\ \biguplus \sigma'_i}
+\]
+
 \section{Implementation}
 \label{sec:implementation}
+
+\begin{lstlisting}
+def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
+  r1.tryReact(a, rx, offer) match {
+    case Retry => 
+      r2.tryReact(a, rx, offer) match {
+        case Retry => Retry
+        case Block => Retry // must retry r1
+        case ans   => ans
+      }
+    case Block => r2.tryReact(a, rx, offer)
+    case ans => ans
+  }
+\end{lstlisting}
 
@@
 \bibliographystyle{abbrvnat}
+\bibliography{reagents}
 
@@
 
-# Contributions
-
-  - make it easier to express fine-grained concurrent algorithms,
-  including blocking operations etc.  make it possible to build them
-  compositionally, which already happens in the literature but is
-  largely unremarked upon.
-
-  - we capture exponential backoff, elimination backoff, and other
-  such patterns once and for all
-
-  - can use flat combining as illustration of capturing very general
-  concurrency abstraction
-
-  - smooth blend of message-passing and shared-state concurrency
-  isolation for shared state, coherent communication through messages,
-  uniform blocking/synchronization
-
-  - operational semantics
-
-  - yields completely lock-free implementation of core CML
-
-  - generalizes join calculus implementation, allowing dissolution
-  even for existing channels
-
-  - generalizes transactions events, but with much better
-  implementation
-
 # To do
@@
 
-2.5 Title, abstract, intro, library overview 
+1.2 Title, abstract, intro, 
+1.0 Background
+2.5 Library overview & examples
 1.5 Semantics 
-2.5 Implementation/algorithms 
-1.0 Performance results
-1.5 Related work
-1.0 Conclusion and bib
+1.5 Implementation/algorithms 
+0.5 Performance results
+1.0 Related work
+0.8 Bib
 

### 2011-10-14 paper/paper.tex
slight intro tweak
work on background section

@@
 concurrent operations.  Reagents are \emph{expressive} and \emph{composable},
-but retain the scalability and performance of existing algorithms:
+but when invoked retain the scalability and performance of existing
+algorithms:
 
@@
 
-Reagents are directly motivated by
+Broadly, we are interested in data structures and algorithms for
+communication, synchronization, or both.  This section gives a brief
+survey of the most important technqiues for communication and
+synchronization in a fine-grained setting---exactly the techniques
+that reagents abstract and generalize.  Readers already familiar with
+Treiber stacks~\cite{?}, elimination-backoff stacks~\cite{?}, dual
+stacks~\cite{?}, and MCS locks~\cite{?} can safely skip to
+\secref{reagents}.
 
-Before introducing reagents, it will be helpful to
+Given our target of cache-coherent, shared-memory architectures, the
+most direct way of communicating between threads is modifying shared
+memory.  The challange is to provide both \emph{atomicity} and
+\emph{scalability}: communications must happen concurrently both
+without corruption and without clogging the limited memory bandwidth,
+even when many cores are communicating.  A simple way to provide
+atomicity is to associate a lock with each shared data structure,
+acquiring the lock before performing any operation.  The prevailing
+wisdom\footnote{With some notable exceptions, like the recent work on
+  \emph{flat combining}~\cite{?}, which we explore in \secref{?}} is
+that such \emph{coarse-grained} locking is inherently unscalable:
+\begin{itemize}
+\item It forces operations on the data structure to be serialized,
+  even when they could be performed in parallel.
+\item It adds extra cache-coherence traffic, since each core must
+  acquire the same lock's cache line in exclusive mode before
+  operating.  For fine-grained communication, that means at least one
+  cache miss per operation.
+\item It is susceptible to ... % blocking 
+\end{itemize}
 
-readers familiar with these algorithms can safely jump to sec:reagents
-
-also background on Scala
+%% For example, containers like
+%% stacks and queues provide communication; contructs like locks and
+%% barriers provide synchronization; and (synchronous) message channels
+%% provide both.
 
@@
 
+\begin{figure}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{Treiber's stack}
+\end{figure}
 

### 2011-10-14 paper/paper.tex
expanding background section

@@
 
+\newcommand{\secref}[1]{\S\ref{sec:#1}}
+\newcommand{\figref}[1]{Fig.~\ref{fig:#1}}
 \newcommand{\elide}[1]{}
@@
 communication, synchronization, or both.  This section gives a brief
-survey of the most important technqiues for communication and
+survey of the most important techniques for communication and
 synchronization in a fine-grained setting---exactly the techniques
@@
 most direct way of communicating between threads is modifying shared
-memory.  The challange is to provide both \emph{atomicity} and
+memory.  The challenge is to provide both \emph{atomicity} and
 \emph{scalability}: communications must happen concurrently both
@@
 wisdom\footnote{With some notable exceptions, like the recent work on
-  \emph{flat combining}~\cite{?}, which we explore in \secref{?}} is
+  \emph{flat combining}~\cite{?}, which we explore in \secref{?}.} is
 that such \emph{coarse-grained} locking is inherently unscalable:
@@
   cache miss per operation.
-\item It is susceptible to ... % blocking 
+\item It is susceptible to ... % blocking %TODO
 \end{itemize}
 
-%% For example, containers like
-%% stacks and queues provide communication; contructs like locks and
-%% barriers provide synchronization; and (synchronous) message channels
-%% provide both.
+%% Say something about CONTENTION.
 
-traditional code for Treiber stack (with exp. backoff),
+To combat scalability problems, data structures employ finer-grained locking,
+or eschew locking altogether.  Fine-grained locking associates locks with
+small, independent parts of a data structure, allowing those parts to be
+manipulated in parallel.  Lockless (or \emph{nonblocking}) data structures
+instead perform updates directly, using hardware-level operations (like
+compare-and-set) to ensure atomicity.  Doing the updates directly means that
+there is no extra communication or contention for locks, but it also generally
+means that the entire update must consist of changing a single word of
+memory---a constraint which is often quite challenging to meet.
 
@@
         case Nil     => return None
-        case a::tail => if (head.cas(cur, tail)) return a
+        case a::tail => 
+          if (head.cas(cur, tail)) return Some(a)
       }
@@
 \end{lstlisting}
-\caption{Treiber's stack}
+\nocaptionrule
+\caption{Treiber's stack (in Scala)}
+\label{fig:classic-treiber}
 \end{figure}
 
-need to explain pattern matching and PartialFunction
+To make this discussion concrete, we give a classic example of a fine-grained
+concurrent data structure: Treiber's lock-free stack~\cite{?}, shown in
+\figref{classic-treiber}.
+%notice: changed to less desirable representation to make CAS work
 
-talk about elimination backoff and blocking versions but don't show code.
+%Most of the Scala code in this paper should be 
+%need to explain pattern matching and PartialFunction
+
+%% For example, containers like
+%% stacks and queues provide communication; contructs like locks and
+%% barriers provide synchronization; and (synchronous) message channels
+%% provide both.
+
+Treiber's stack scales better than a lock-based stack mainly because it
+decreases the amount of shared data that must be updated per operation:
+instead of acquiring a shared lock, altering a shared stack pointer, and
+releasing the shared lock, Treiber's stack does a single CAS to update the
+shared pointer directly.  Scalability can be further improved by
+
+%talk about elimination backoff and blocking versions but don't show code.
 
@@
 \item with buckets?
-\item Counter -> semaphore
+\item Counter to semaphore
 \item RWLocks

### 2011-10-17 paper/paper.tex
progress on background and reagents sections

@@
 
-\newcommand{\secref}[1]{\S\ref{sec:#1}}
-\newcommand{\figref}[1]{Fig.~\ref{fig:#1}}
-\newcommand{\elide}[1]{}
-
 \input{packages}
+\input{macros}
 
@@
 
-%% Say something about CONTENTION.
-
 To combat scalability problems, data structures employ finer-grained locking,
@@
 
-To make this discussion concrete, we give a classic example of a fine-grained
-concurrent data structure: Treiber's lock-free stack~\cite{?}, shown in
-\figref{classic-treiber}.
+\figref{classic-treiber} gives a classic example of a fine-grained
+concurrent data structure: Treiber's lock-free stack~\cite{?}.  The
+stack is represented as an \emph{immutable} linked list.  Mutation
+occurs solely through the \lstinline{head} pointer, represented here
+as an \lstinline{AtomicReference}.  The \lstinline{head} is mutated
+using \lstinline{compareAndSet} (here abbreviated as \lstinline{cas}),
+which takes an expected value and a new value, and atomically updates
+the reference
+
 %notice: changed to less desirable representation to make CAS work
@@
 
-%% For example, containers like
-%% stacks and queues provide communication; contructs like locks and
-%% barriers provide synchronization; and (synchronous) message channels
-%% provide both.
+%% Say something about CONTENTION.
 
-Treiber's stack scales better than a lock-based stack mainly because it
-decreases the amount of shared data that must be updated per operation:
-instead of acquiring a shared lock, altering a shared stack pointer, and
-releasing the shared lock, Treiber's stack does a single CAS to update the
-shared pointer directly.  Scalability can be further improved by
+Treiber's stack scales better than a lock-based stack mainly because
+it decreases the amount of shared data that must be updated per
+operation: instead of acquiring a shared lock, altering a shared stack
+pointer, and releasing the shared lock, Treiber's stack does a single
+CAS to update the shared pointer directly.  But the head pointer of
+the stack is still a centralized source of contention.  Spreading out
+the contention requires parallelizing stack operations---but how,
+given that they all involve modifying the top of the stack?
+
+Parallelization requires a change of perspective.  Normally, we view
+concurrent operations on the same part of a data structure as
+competing to atomically update that data structure; we want the
+operations to be isolated.  But if
 
@@
 
-First we give an overview of the library, and then we illustrate its
-expressiveness and composability by showing how to implement several
+In the last section, we saw a range of strategies for communication
+and synchronizion at a fine grain:
+\begin{itemize}
+\item using CAS to make direct, atomic updates;
+\item using exponential backoff to ease contention; 
+\item using side-channel communication to increase parallelism; 
+\item using disjoint spinwaiting to eliminate useless cache traffic; 
+\item using dual data structures to incorporate blocking.
+\end{itemize}
+Reagents unify these disparate techniques into a single framework for
+writing concurrent algorithms.  Some of the techniques are built into
+the framework, while others are expressible as reusable abstractions.
+% extensibility and composability
+
+\subsection{The main idea}
+
+%% The aim of reagents is to make it easy to express these and other
+%% patterns, while providing compositionality blah blah
+
+%% First we give an overview of the library, and then we illustrate its
+%% expressiveness and composability by showing how to implement several
 
@@
 message-passing turns out to be a good way to understand what's going on.
+
+\subsection{Expressiveness by example}
 
@@
 \item Dining philosophers(ish)
-\item Flat combining
+\item Flat combining - doesn't work if commit is two-phase process.
+  probably too much to bring into scope for this paper, anyway
 \end{itemize}
+
+\begin{lstlisting}
+class MSQueue[A] {
+  private final case class Node(
+    data: A, next: Ref[Node] = Ref(null)
+  )
+  private val head = Ref(Node(null))
+  private val tail = Ref(head.read!())
+
+  val enq: Reagent[A, Unit] = computed { (x:A) =>
+    @tailrec def search: Reagent[Unit,Unit] = 
+      tail.read ! () match {
+        case    Node(_, r@Ref(null)) => 
+          r.cas(null, Node(x))
+        case ov@Node(_, Ref(nv)) => 
+          tail.cas(ov,nv) !? (); search
+      }
+    search
+  }
+  val tryDeq: Reagent[Unit, Option[A]] = head.upd {
+    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
+    case emp => (emp, None)
+  }
+}
+\end{lstlisting}
 
@@
 
-\newcommand{\ra}{\rightarrow}
-\newcommand{\gives}{\vdash}
-\newcommand{\proves}{\vdash}
-\newcommand{\kw}[1]{\textsf{#1}}
-\newcommand{\GA}{\ |\ }
-\newcommand{\step}{\longrightarrow}
-\newcommand{\cstep}[1]{\stackrel{#1}{\Longrightarrow}}
+Try to keep this minimal for the sake of space
 
-\newcommand{\evalCtx}{\mathbb{E}}
-\newcommand{\procCtx}{\mathbb{P}}
+Purpose of the semantics is mainly to clarify the relationship between
+the shared-state and message-passing aspects of reagents.
 
-\newcommand{\ret}[1]{\kw{ret}(#1)}
-\newcommand{\never}{\kw{never}}
-\newcommand{\postCommit}[1]{\kw{postCommit}(#1)}
-\newcommand{\computed}[1]{\kw{computed}(#1)}
-\newcommand{\compose}{{\small\texttt{>=>}}}
-\newcommand{\lift}[1]{\kw{lift}(#1)}
-\newcommand{\send}[1]{#1}
-\newcommand{\cas}[3]{\kw{cas}(#1,#2,#3)}
-\newcommand{\rd}[1]{\kw{read}(#1)}
-\newcommand{\newChan}{\kw{new Chan}}
-\newcommand{\newRef}[1]{\kw{new Ref(#1)}}
+\begin{itemize}
+  \item read, cas
+  \item send
+  \item choice
+  \item react
+  \item bind and ret
+  \item computed?
+\end{itemize}
 
@@
 
+  - somewhere say that: we examine fairly simple examples in the
+  paper, due to size constraints.  but we can handle more complex
+  cases like sets/maps represented via skip lists, basket queues, 
+  and are exploring flat combining.
+
 # Outline

### 2011-10-17 paper/paper.tex
paper work

@@
 lost parallelism \emph{and} an extra cache miss.  The extra misses can easily
-yield parallel \emph{slowdown}, more than negating the benefits of the
-remaining parallelism.
+accumulate to yield parallel \emph{slowdown}, more than negating the benefits
+of the remaining parallelism.
 
@@
 blend shared-state and message-passing concurrency
+
+implemented a Scala library
 
@@
 \end{itemize}
-Reagents unify these disparate techniques into a single framework for
-writing concurrent algorithms.  Some of the techniques are built into
-the framework, while others are expressible as reusable abstractions.
+Reagents incorporate these disparate techniques into a uniform framework for
+writing concurrent algorithms.  Some of the techniques are built into the
+framework, while others are expressible as reusable abstractions.
+
+more easily and concisely
 % extensibility and composability
@@
 \subsection{The main idea}
+
+Reagents are a new instance of an old idea: computations can be usefully
+represented as data.
+
+% no reason not to present this as an ADT
 
@@
 class MSQueue[A] {
-  private final case class Node(
-    data: A, next: Ref[Node] = Ref(null)
-  )
+  private case class Node(data: A, next: Ref[Node])
   private val head = Ref(Node(null))
@@
 
-  val enq: Reagent[A, Unit] = computed { (x:A) =>
-    @tailrec def search: Reagent[Unit,Unit] = 
+  @tailrec private def search(x:A): Reagent[Unit,Unit] = 
     tail.read ! () match {
       case Node(_, r@Ref(null)) => 
-          r.cas(null, Node(x))
+        r.cas(null, Node(x, new Ref(null)))
       case ov@Node(_, Ref(nv)) => 
-          tail.cas(ov,nv) !? (); search
+        tail.cas(ov,nv) !? (); search(x)
     }
-    search
-  }
+  val enq: Reagent[A, Unit] = computed search
   val tryDeq: Reagent[Unit, Option[A]] = head.upd {
@@
 the shared-state and message-passing aspects of reagents.
+
+(fractional?) permissions model
 
@@
 \label{sec:related}
+
+We believe it is crucial to have XX, YY together
+
+STM, Communicating transactions and company, elastic transactions and company
+
+CML, Transactional events, join calculus
+
+Paper on composing lock-free data structure operations
 

### 2011-10-18 paper/paper.tex
paper
work on Reagents explanation

@@
 \section{Background}
+\label{sec:background}
 
@@
 
+% OPTIMISTIC CONCURRENCY
+
 %notice: changed to less desirable representation to make CAS work
@@
 
-Treiber's stack scales better than a lock-based stack mainly because
-it decreases the amount of shared data that must be updated per
-operation: instead of acquiring a shared lock, altering a shared stack
-pointer, and releasing the shared lock, Treiber's stack does a single
-CAS to update the shared pointer directly.  But the head pointer of
-the stack is still a centralized source of contention.  Spreading out
-the contention requires parallelizing stack operations---but how,
-given that they all involve modifying the top of the stack?
+Treiber's stack scales better than a lock-based stack mainly because it
+decreases the amount of shared data that must be updated per operation:
+instead of acquiring a shared lock, altering a shared stack pointer, and
+releasing the shared lock, Treiber's stack does a single CAS to update the
+shared pointer directly.  But the head pointer of the stack is still a
+centralized source of contention.  Spreading out the contention requires
+parallelizing stack operations---but how, given that they all involve
+modifying the top of the stack?
 
 Parallelization requires a change of perspective.  Normally, we view
-concurrent operations on the same part of a data structure as
-competing to atomically update that data structure; we want the
-operations to be isolated.  But if
+concurrent operations on the same part of a data structure as competing to
+atomically update that data structure; we want the operations to be isolated.
+But if
 
@@
 
-In the last section, we saw a range of strategies for communication
-and synchronizion at a fine grain:
-\begin{itemize}
-\item using CAS to make direct, atomic updates;
-\item using exponential backoff to ease contention; 
-\item using side-channel communication to increase parallelism; 
-\item using disjoint spinwaiting to eliminate useless cache traffic; 
-\item using dual data structures to incorporate blocking.
-\end{itemize}
-Reagents incorporate these disparate techniques into a uniform framework for
-writing concurrent algorithms.  Some of the techniques are built into the
-framework, while others are expressible as reusable abstractions.
+Reagents are a new instance of an old idea: representing computations as data.
+The computations being represented are fine-grained concurrent operations, so
+a value of type \reagent{A}{B} represents a function from \lstinline{A} to
+\lstinline{B} that internally interacts with a concurrent data structure
+through mutation, synchronization, or both.
 
-more easily and concisely
+Unlike $\lambda$-abstractions, which encompass arbitrary code and treat that
+code opaquely, reagents are built up from a limited set of \emph{combinators}
+that are known to the implementation.  Existing reagents---for example, those
+built by a concurrency expert---can be composed or tailored by library users
+by applying additional combinators, without knowledge of the internal
+implementation.  At any point a user can invoke a reagent, either as a
+\emph{reactant} or a \emph{catalyst}---a distinction explained in
+\secref{easy-examples}.
+
+Reagents are patterned after Concurrent ML~\cite{?} and Haskell's STM
+monad~\cite{?}, which represent (respectively) synchronous and atomic
+operations as first-class values; we discuss the relationship in
+\secref{related}.
+%
+Our contribution is giving a set of combinators appropriate for expressing and
+composing fine-grained concurrent algorithms (\secref{easy-examples}) with a
+clear semantics (\secref{semantics}), ... %TODO
+
+%more easily and concisely
 % extensibility and composability
-
-\subsection{The main idea}
-
-Reagents are a new instance of an old idea: computations can be usefully
-represented as data.
-
-% no reason not to present this as an ADT
 
@@
 
-chemistry metaphor
+\subsection{Basic reagents: updates, messages, and choice}
+\label{sec:easy-examples}
 
-lack of isolation needed to express many lock-free algorithms.
-message-passing turns out to be a good way to understand what's going on.
-
-\subsection{Expressiveness by example}
+In \secref{background}, we saw a range of strategies for communication and
+synchronizion at a fine grain:
+\begin{itemize}
+\item using CAS to make direct, optimistic, and atomic updates;
+\item using exponential backoff to ease contention; 
+\item using side-channel communication to increase parallelism; 
+\item using disjoint spinwaiting to eliminate useless cache traffic; 
+\item using dual data structures to incorporate blocking.
+\end{itemize}
+Reagents incorporate these disparate techniques into a uniform set of
+combinators for writing concurrent algorithms.  Some of the techniques are
+built into the combinators themselves, while others are expressible as
+reusable abstractions.
 
@@
   private val head = new Ref[List[A]](Nil)
-  val push: Reagent[A, Unit] = head.upd {
-    (xs, x) => (x:xs, ())
+  val push: Reagent[A, Unit] = upd(head) {
+    case (xs, x) => (x:xs, ())
   }
-  val tryPop: Reagent[Unit, Option[A]] = head.upd {
+  val tryPop: Reagent[Unit, Option[A]] = upd(head) {
     case (x::xs, ()) => (xs, Some(x))
@@
 \end{lstlisting}
+
+The first two techniques are built into the combinators.  We introduce a new
+type \lstinline{Ref[A]} for atomically updatable references holding data of
+type \lstinline{A}.  Unlike Scala's (and Java's) \lstinline{AtomicReference},
+\lstinline{Ref} values can only be accessed through the following reagent
+combinators:
+\begin{lstlisting}[mathescape=true]
+read: Ref[A] => Reagent[Unit, A]
+cas:  Ref[A] * A * A => Reagent[Unit, Unit]
+upd:  Ref[A] => PartialFunction[(A,B), (A,C)] 
+             => Reagent[B, C]
+\end{lstlisting}
+By restricting access to the combinators, ... %TODO
+
+% Freedom to change kcas implementations
+
+The \lstinline{read} combinator has a straightforward signature: if
+\lstinline{r} has type \lstinline{Ref[A]}, then \lstinline{read(r)} is a
+\emph{value} of type \lstinline{Reagent[Unit, A]} which, when invoked, takes a
+unit\footnote{\lstinline{Unit} in Scala is akin to \lstinline{void}: it is a
+  type with a single member, written \lstinline{()}.} argument and yields an
+\lstinline{A} result.
+
+The \lstinline{cas} combinator takes two additional \lstinline{A} arguments,
+giving the expected and updated values, respectively.  Unlike its counterpart
+for \lstinline{AtomicReference}, a \lstinline{cas} reagent does \emph{not}
+yield a boolean result.  Instead, the reagent library builds in handling of
+failed CASes, including an optimistic retry loop and exponential backoff.  The
+failure mechanics depend on which other combinators are involved, so we delay
+a complete discussion until \secref{failure}.
+
+The \lstinline{upd} combinator
+
+%% THIS SEEMS MORE CONFUSING THAN HELPFUL
+%% For now, it suffices to know that
+%% invoking a \lstinline{cas} will complete and return \lstinline{()} only when
+%% the CAS has successfully completed.
+
+%% \begin{lstlisting}[mathescape=true]
+%% def upd[A,B,C](r: Ref[A], f: (A,B) $\rightharpoonup$ (A,C)): Reagent[B,C] 
+%%   = computed { 
+%%       (b: B) =>       
+%% \end{lstlisting}
+
+chemistry metaphor.  note that catalysts increase parallelism and remove need
+for devoted thread
+
+lack of isolation needed to express many lock-free algorithms.
+message-passing turns out to be a good way to understand what's going on.
+
+key point: traditional lock-free algorithms that use a single cas do not pay
+any penalty for composability/kcas potential
+
+show how, for counter, tryDec and dec are inter-expressible
+
+% TODO: fix types for cases below
+\begin{lstlisting}
+class Counter {
+  private val state = new Ref[Int](0)
+  val get = upd(state) { case i => (i, i) }
+  val inc = upd(state) { case i => (i+1, i) }
+  val dec = upd(state) { case i if (i > 0) => (i-1, i) }
+  val tryDec = upd(state) {
+     case i if (i == 0) => (0,   None)
+     case i              => (i-1, Some(i))
+  }
+}
+\end{lstlisting}
+
+% nice catalysts: zip and unzip
+
+\subsection{Failure is an \textsf{Option}}
+\label{sec:failure}
+
+blocking failure versus retryable failure
+
+\begin{lstlisting}
+  val tryDec = 
+\end{lstlisting}
+
+\subsection{Advanced reagents: CAS, computed, and composed}
+\label{sec:hard-examples}
 
@@
   private case class Node(data: A, next: Ref[Node])
-  private val head = Ref(Node(null))
-  private val tail = Ref(head.read!())
+  private val head = new Ref(Node(null))
+  private val tail = new Ref(head.read!())
 
   @tailrec private def search(x:A): Reagent[Unit,Unit] = 
-    tail.read ! () match {
+    read(tail) ! () match {
       case Node(_, r@Ref(null)) => 
-        r.cas(null, Node(x, new Ref(null)))
+        cas(r, null, Node(x, new Ref(null)))
       case ov@Node(_, Ref(nv)) => 
-        tail.cas(ov,nv) !? (); search(x)
+        cas(tail, ov, nv) !? (); 
+        search(x)
     }
   val enq: Reagent[A, Unit] = computed search
-  val tryDeq: Reagent[Unit, Option[A]] = head.upd {
+  val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
     case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
@@
 \end{lstlisting}
+
+post-commit action for catching up tail?
 
@@
 \begin{lstlisting}
+def !(a: A): B = tryReact(a, Reaction.inert, null) match {
+  case (_: BacktrackCommand) => {
+    val backoff = new Backoff
+    val maySync = this.maySync // cache
+    @tailrec def retryLoop(shouldBlock: Boolean): B = {
+      val wait = maySync || shouldBlock
+      val waiter = if (wait) new Waiter[B](shouldBlock) 
+                   else null
+
+      tryReact(a, Reaction.inert, waiter) match {
+        case (bc: BacktrackCommand) if wait => {
+          bc.bottom(waiter, backoff, snoop(a))
+          waiter.tryAbort      // rescind waiter,
+          waiter.poll match { // but check if completed
+            case Some(ans) => ans.asInstanceOf[B] 
+            case None       => retryLoop(bc.isBlock)
+          }
+        }
+        case Retry => backoff.once; retryLoop(false)
+        case Block => retryLoop(true)
+        case ans   => ans.asInstanceOf[B]
+      }
+    }
+    retryLoop(false)
+  }
+  case ans => ans.asInstanceOf[B]
+}
+\end{lstlisting}
+
+\begin{lstlisting}
 def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
@@
 
+\begin{lstlisting}
+  def tryReact(u: Unit, rx: Reaction, offer: Offer[B]): Any = 
+    if (rx.canCASImmediate(k, offer)) {
+      if (casI(expect, update))
+        k.tryReact((), rx, offer)
+      else Retry
+    } else k.tryReact((), rx.withCAS(Ref.this, expect, update), offer)
+\end{lstlisting}
+
 \section{Performance}
@@
 
-CML, Transactional events, join calculus
+CML, Transactional events, async CML, join calculus
 

### 2011-10-30 paper/paper.tex
a bit of outlining
new benchmarks, lots of writing

@@
 implemented a Scala library
+
+CLEAR COST MODEL
 
@@
 
+\elide{
+In \secref{background}, we saw a range of strategies for communication and
+synchronizion at a fine grain:
+\begin{itemize}
+\item using CAS to make direct, optimistic, and atomic updates;
+\item using exponential backoff to ease contention; 
+\item using side-channel communication to increase parallelism; 
+\item using disjoint spinwaiting to eliminate useless cache traffic; 
+\item using dual data structures to incorporate blocking.
+\end{itemize}
+Reagents incorporate these disparate techniques into a uniform set of
+combinators for writing concurrent algorithms.  Some of the techniques are
+built into the combinators themselves, while others are expressible as
+reusable abstractions.
+}
+
+The goal of reagents is to systematize the expression of these
+disparate techniques into a uniform set of combinators
+
 \section{Reagents}
 \label{sec:reagents}
+
+\begin{figure}
+\begin{lstlisting}[mathescape=true,frame=single]
+// combinators for shared state
+upd:  Ref[A] => PartialFunction[(A,B), (A,C)] 
+             => Reagent[B,C]
+read: Ref[A] => Reagent[Unit, A]
+cas:  Ref[A] * A * A => Reagent[Unit, Unit]
+
+// combinators for message passing
+swap: Endpoint[A,B] => Reagent[A,B]
+
+// combinators for composition
++  : Reagent[A,B] * Reagent[A,B] => Reagent[A,B]
+>> : Reagent[A,B] * Reagent[B,C] => Reagent[A,C]
+|||  : Reagent[A,B] * Reagent[A,C] => Reageng[A,(B,C)]
+
+// combinators for computation
+ret: A => Reagent[Unit, A]
+computed: PartialFunction[A, Reagent[Unit, B]]
+          => Reagent[A,B]
+postCommit: (A => Unit) => Reagent[A,A]
+\end{lstlisting}
+\label{fig:combinators}
+\end{figure}
 
@@
 
-\subsection{Basic reagents: updates, messages, and choice}
-\label{sec:easy-examples}
+\elide{
+want to talk about blocking versus retryable failure
+  - probably can wait until CAS introduced, toward end
+  - on the other hand, need to be clear about how this works 
+    for upd.  maybe introducing the distinction early on makes 
+    sense.
 
-In \secref{background}, we saw a range of strategies for communication and
-synchronizion at a fine grain:
-\begin{itemize}
-\item using CAS to make direct, optimistic, and atomic updates;
-\item using exponential backoff to ease contention; 
-\item using side-channel communication to increase parallelism; 
-\item using disjoint spinwaiting to eliminate useless cache traffic; 
-\item using dual data structures to incorporate blocking.
-\end{itemize}
-Reagents incorporate these disparate techniques into a uniform set of
-combinators for writing concurrent algorithms.  Some of the techniques are
-built into the combinators themselves, while others are expressible as
-reusable abstractions.
+where should we introduce react versus catalyze?
+  - want to talk about react early; perhaps introduce them 
+    separately?
 
-\begin{lstlisting}
-class TreiberStack[A] {
+diagram(s) of reaction?
+  - we'll see if there's space...
+
+  upd
+    nonblocking: 
+      Treiber
+    blocking: 
+      Treiber
+      Counter/semaphore
+  composition
+    Treiber & counter?
+    Treiber & lift => blocking
+  messages and choice
+    elimination Treiber (& counter!  implications for semaphores?)
+    blocking
+  catalysts
+    zip and unzip, tree barriers
+  computed, read, and cas - retryable failure, postCommit?
+    msqueue
+  parallel composition?
+  timeouts?  TODO
+}
+
+\subsection{Atomic updates on \lstinline{Ref}s}
+
+Memory is shared between reagents using the type \lstinline{Ref[A]} of
+atomically-updatable references.  These references are only accessible
+through reagent combinators, which allows them to provide
+functionality beyond the built-in \lstinline{AtomicReference[A]} type,
+as we will see.class TreiberStack[A] {
   private val head = new Ref[List[A]](Nil)
@@
 \end{lstlisting}
+In Scala, anonymous partial functions are written as a series of cases
+enclosed in braces.  For \lstinline{push} and \lstinline{tryPop}, the
+case analyses are exhaustive, so the functions are in fact total.
 
-The first two techniques are built into the combinators.  We introduce a new
-type \lstinline{Ref[A]} for atomically updatable references holding data of
-type \lstinline{A}.  Unlike Scala's (and Java's) \lstinline{AtomicReference},
-\lstinline{Ref} values can only be accessed through the following reagent
-combinators:
-\begin{lstlisting}[mathescape=true]
-read: Ref[A] => Reagent[Unit, A]
-cas:  Ref[A] * A * A => Reagent[Unit, Unit]
-upd:  Ref[A] => PartialFunction[(A,B), (A,C)] 
-             => Reagent[B, C]
+Being reagents, \lstinline{push} and \lstinline{tryPop} are inert
+values.  They can be invoked using the \lstinline{!} method provided
+by reagents; for a \lstinline{Reagent[A,B]} the \lstinline{!} method
+takes an \lstinline{A} and returns a \lstinline{B}.  Scala permits
+infix notation for methods, so we can use a \lstinline{TreiberStack}
+as follows:
+\begin{lstlisting}
+  var s = new TreiberStack[Int]; s.push ! 42
 \end{lstlisting}
-By restricting access to the combinators, ... %TODO
+The key point is that when we invoke these reagents, we are executing
+\emph{exactly} the same algorithms written by hand in
+\secref{background}, including the retry loop with exponential
+backoff; reagents systematize and internalize common patterns of
+fine-grained concurrency.  But by exposing \lstinline{push} and
+\lstinline{tryPop} as reagents, rather than methods, we leave the door
+open for further composition and tailoring by a user of the data
+structure, as we shall see below.
 
-% Freedom to change kcas implementations
+While \lstinline{tryPop} handles both empty and nonempty stacks, we
+can write a variant that drops the empty case:
+\begin{lstlisting}
+val pop: Reagent[Unit, A] = upd(head) {
+  case (x::xs, ()) => (xs, x)
+}
+\end{lstlisting}
+Now our update function really is partial.  An invocation
+\lstinline{s.pop ! ()} will block the calling thread unless or until
+the stack is nonempty.  Blocking and signalling are entirely handled
+by the reagent implementation, using standard
+techniques~(\secref{implementation}); there is therefore no risk of a
+lost wakeup.  %TODO: add more here?
+
+\elide{
+% What's the cost/benefit for this example?
+isomorphic to a stack of unit values
+\begin{lstlisting}
+class Counter {
+  private val c = new Ref[Int](0)
+  val inc = upd(c) { case (i, ()) => (i+1, i) }
+  val dec = upd(c) { case (i, ()) if (i > 0) => (i-1, i) }
+  val tryDec = upd(c) {
+     case (i, ()) if (i == 0) => (0,   None)
+     case (i, ())              => (i-1, Some(i))
+  }
+}
+\end{lstlisting}
+
+Roughly equivalent to the unfair JUC semaphore implementation.
+}
+
+%% THIS SEEMS MORE CONFUSING THAN HELPFUL
+%% For now, it suffices to know that
+%% invoking a \lstinline{cas} will complete and return \lstinline{()} only when
+%% the CAS has successfully completed.
+
+%% \begin{lstlisting}[mathescape=true]
+%% def upd[A,B,C](r: Ref[A], f: (A,B) $\rightharpoonup$ (A,C)): Reagent[B,C] 
+%%   = computed { 
+%%       (b: B) =>       
+%% \end{lstlisting}
+
+\subsection{Synchronization: how reagents react}
+
+%% References allow threads to communicate through atomic changes to
+%% shared state.  The atomic updates appear to happen in a sequence
+%% without overlaps: concurrent threads cannot interfere with each other
+%% or otherwise detect that they are running concurrently.
+
+Another staple of concurrency libraries is synchronization, whereby
+threads are made mutually aware of reaching a particular state.
+Synchronization is completementary to shared state: the former
+provides concurrent interaction, while the latter (through atomic
+update) provides sequential isolation.  % TODO: make this less cryptic
+
+Reagents interact through \emph{synchronous swap channels}, which
+consist of two complementary \emph{endpoints}:
+\begin{lstlisting}
+  mkChan[A,B]: (Endpoint[A,B], Endpoint[B,A])
+\end{lstlisting}
+The combinator for communication is \lstinline{swap} (see
+\figref{combinators}), which lifts an \lstinline{Endpoint[A,B]} to a
+\lstinline{Reagent[A,B]}.  When two reagents communicate on opposite
+endpoints, they provide messages of complementary type (\lstinline{A}
+and \lstinline{B}, for example) and receive each other's value.  We
+call a successful communication a \emph{reaction} between reagents.
+On the other hand, if no complementary message is available,
+\lstinline{swap} will block until a reaction can take place.  Once
+again, the waiting and signalling protocol is managed internally by
+the reagent implementation, so lost wakeups are impossible.
+
+The \lstinline{swap} combinator alone is not powerful enough to
+express many synchronization primitives.  It becomes powerful when
+used together with other combinators, particularly \emph{choice} and
+\emph{sequence}.
+
+%% lack of isolation needed to express many lock-free algorithms.
+%% message-passing turns out to be a good way to understand what's going on.
+
+%% key point: traditional lock-free algorithms that use a single cas do not pay
+%% any penalty for composability/kcas potential
+
+\subsection{Composing reagents: choice}
+
+If \lstinline{r} and \lstinline{s} are two reagents of the same type,
+their \emph{choice} \lstinline{r + s} will behave like one of them,
+nondeterministically, when invoked.  The most straightforward use of
+choice is in the spirit of Unix's \lstinline{select}~\cite{?}: waiting
+on several signals simultaneously, while consuming only one of them.
+For example, if \lstinline{c} and \lstinline{d} are endpoints of the
+same type, \lstinline{swap(c) + swap(d)} is a reagent that will accept
+exactly one message, either from \lstinline{c} or from \lstinline{d}.
+If neither endpoint has a message available, the reagent will block
+until one of them does.  As usual, the blocking protocol is handled by
+the reagent library.
+
+A more novel use of choice is adding backoff strategies (of the kind
+described in \secref{?}).  For example, we can build an elimination
+backoff stack as follows:
+\begin{lstlisting}
+class EliminationStack[A] {
+  private val s = new TreiberStack[A]
+  private val (elimPop, elimPush) = mkChan[Unit,A]
+  val push: Reagent[A,Unit] = s.push + swap(elimPush)
+  val pop:  Reagent[Unit,A] = s.pop  + swap(elimPop)
+}
+\end{lstlisting}
+Choice is left-biased,\footnote{The left bias makes choice more
+  similar to \lstinline{orElse} in Haskell's STM than to CML's choice,
+  but there are differences from both.  See \secref{related}.  } so
+when \lstinline{push} is invoked, it will first attempt to push onto
+the underlying Treiber stack.  If the underlying push fails (due to a
+lost CAS race), \lstinline{push} will attempt to send a message along
+\lstinline{elimPush}, \emph{i.e.}, to synchronize with a concurrent
+popper.  If it succeeds, the \lstinline{push} reagent completes
+without ever having modified its underlying stack.
+
+We noted that \lstinline{s.push} can only fail due to a failed CAS.
+Such a failure is \emph{transient}: it can only be caused by active
+interference from another thread.  A reagent that has failed
+transiently should retry, rather than block, following the concurrency
+patterns laid out in \secref{background}.  On the other hand,
+\lstinline{s.pop} can fail in a \emph{permanent} way: the underlying
+stack \lstinline{s} may be empty.  This failure is permanent in the
+sense that only activity by another thread can enable the reagent to
+proceed.  When faced with a permanent failure, a reagent should block
+until signalled that the underlying state has changed.
+
+For choice, failure depends on the underlying reagents.  A choice
+reagent fails permanently only when \emph{both} of its underlying
+reagents have failed permanently.  On the other hand, if either fails
+transiently, the choice reagent has failed transiently and should
+therefore retry.  Reasoning along these lines, we deduce that
+\lstinline{push} never blocks, since the underlying \lstinline{s.push}
+can only fail transiently.  On the other hand, \lstinline{pop} can
+block because \lstinline{s.pop} can fail permanently on an empty stack
+and \lstinline{swap(elimPop)} can fail permanently if there are no
+offers from pushers.
+
+When \lstinline{push} or \lstinline{pop} retry, they follow an
+exponential backoff pattern with \emph{spinwaiting} for elimination
+offers, exactly as described in \secref{background}.  The backoff
+logic follows directly from the implementation of reagents; see
+\secref{implementation} for details.
+
+\elide{
+On each retry, \lstinline{push} will \emph{spinwait} for
+another thread to accept its message along \lstinline{elimPush}; the
+length of the wait grows exponentially, as part of the exponential
+backoff logic.  Once the waiting time is up, the communication attempt
+is cancelled, and the whole reagent is retried.  This protocol is
+elaborated in \secref{implementation}.
+}
+
+\elide{
+This example demonstrates
+
+although we have specified \lstinline{TreiberStack} here
+
+choice on messages, choice with timeout, choice with mix of reagents
+
+emphasize: combine without knowledge of implementation
+
+isolation and interaction
+}
+
+\subsection{Composing reagents: sequencing and pairing}
+
+Choice offers a kind of disjunction on reagents.  There are also two
+ways of \emph{conjoining} two reagents, so that the composed reagent
+has the effect of both underlying reagents:
+\begin{itemize} %TODO: graphical depiction?
+\item End-to-end composition, via \emph{sequencing}: if 
+  \lstinline{r: Reagent[A,B]} and \lstinline{s: Reagent[B,C]} then 
+  \lstinline{r >> s: Reagent[A,C]}.
+\item Side-by-side composition, via \emph{pairing}: if 
+  \lstinline{r: Reagent[A,B]} and \lstinline{s: Reagent[B,C]} then 
+  \lstinline{r ||| s: Reagent[A,(B,C)]}.  
+\end{itemize}
+These combinators differ only in information flow.  Each guarantees
+that the atomic actions of both underlying reagents become a single
+atomic action for the composition.  For example, if \lstinline{s1} and
+\lstinline{s2} are both stacks, then \lstinline{s1.pop >> s2.push} is
+a reagent that will atomically transfer an element from the top of one
+to the top of the other.  The reagent will block if \lstinline{s1} is
+empty.  Similarly, \lstinline{s1.pop ||| s2.pop} will pop, in one
+atomic action, the top elements of both stacks, or block if either is
+empty.
+
+Here we begin to see the benefits of the reagent abstraction.  Both of
+the example combinations work regardless of how the underlying stacks
+are implemented.  If both stacks use elimination backoff, the
+conjoined operations will potentially use elimination on both
+simultaneously.  This behavior is entirely emergent; it does not
+require any code on the part of the stack author, and it does not
+require the stack user to know anything about the implementation.
+Reagents can be composed in unanticpated ways.
+
+The double-pop reagent implicitly gives a solution to the Dining
+Philosophers problem~\cite{?}: two resources are atomically acquired.
+Internally, the reagent library ensures dead- and live-lock freedom
+for conjoined reagents.
+
+The implementation details for conjunctions are discussed
+later~(\secref{implementation}), but a key point is that the
+performance cost is \emph{pay as you go}.  Single atomic reagents like
+\lstinline{push} and \lstinline{pop} execute a single CAS---just like
+the standard nonblocking algorithms they are meant to implement---even
+though these operations can be combined into larger atomic operations.
+The cost of conjunction is only incurred when a conjoined reagent is
+actually used.  This is a crucial difference from STM, which must
+generally incur overheads regardless of the size of the atomic blocks;
+see~\secref{related} for more discussion.
+
+\elide{
+Unlike function composition, however, the
+atomicity and blocking behavior of the reagents is combined.
+
+fills a key gap in current libraries for fine-grained concurrency.
+
+KEY: pay as you go!
+
+dining philosophers
+
+pop from stack, send as message
+
+exponential search for message passing
+
+show how, for counter, tryDec and dec are inter-expressible
+}
+
+\subsection{Catalysts: persistent reagents}
+\label{sec:catalysts}
+
+In chemistry, a reagent is a participant in a reaction, and reagents
+are further subdivided into \emph{reactants}, which are consumed
+during reaction, and \emph{catalysts}, which enable reactions but are
+not consumed by them.  Invoking a reagent using the \lstinline{!}
+operator
+
+chemistry metaphor.  note that catalysts increase parallelism and remove need
+for devoted thread
+
+nice catalysts: zip and unzip, tree barrier
+
+\elide{
+class TreeBarrier {
+  public readonly Synchronous.Channel[] Arrive;
+  private readonly Join j; // create j, init chans ...
+  public TreeBarrier(int n) {Wire(0, n-1, () => {});}
+  private void Wire(int low, int high, Action Done) {
+    if (low == high) j.When(Arrive[low]).Do(Done);
+    else if (low + 1 == high)
+      j.When(Arrive[low]).And(Arrive[high]).Do(Done);
+    else { // low + 1 < high
+      Synchronous.Channel Left, Right; // init chans       
+      j.When(Left).And(Right).Do(Done);
+      int mid = (low + high) / 2;
+      Wire(low, mid, () => Left());
+      Wire(mid + 1, high, () => Right()); 
+    }
+  }
+}
+}
+
+\begin{lstlisting}
+def zip(in1: Endpoint[Unit,  A], 
+        in2: Endpoint[Unit,  B], 
+        out: Endpoint[(A,B), Unit]) = 
+  dissolve((swap(in1) ||| swap(in2)) >> out)
+\end{lstlisting}
+
+can be made more abstract: use reagent arguments instead!
+
+\subsection{Dynamic reagents and \lstinline{cas}}
+
+The combinators introduced so far are powerful, but they impose a
+strict phase separation: reagents are constructed prior to, and
+independently from, the data that flows through them.  Phase
+separation is useful, because it allows reagent execution to be
+optimized based on complete knowledge of the computation to be
+performed (see \secref{implementation}).
 
@@
 
-The \lstinline{upd} combinator
-
-%% THIS SEEMS MORE CONFUSING THAN HELPFUL
-%% For now, it suffices to know that
-%% invoking a \lstinline{cas} will complete and return \lstinline{()} only when
-%% the CAS has successfully completed.
-
-%% \begin{lstlisting}[mathescape=true]
-%% def upd[A,B,C](r: Ref[A], f: (A,B) $\rightharpoonup$ (A,C)): Reagent[B,C] 
-%%   = computed { 
-%%       (b: B) =>       
-%% \end{lstlisting}
-
-chemistry metaphor.  note that catalysts increase parallelism and remove need
-for devoted thread
-
-lack of isolation needed to express many lock-free algorithms.
-message-passing turns out to be a good way to understand what's going on.
-
-key point: traditional lock-free algorithms that use a single cas do not pay
-any penalty for composability/kcas potential
-
-show how, for counter, tryDec and dec are inter-expressible
-
-% TODO: fix types for cases below
-\begin{lstlisting}
-class Counter {
-  private val state = new Ref[Int](0)
-  val get = upd(state) { case i => (i, i) }
-  val inc = upd(state) { case i => (i+1, i) }
-  val dec = upd(state) { case i if (i > 0) => (i-1, i) }
-  val tryDec = upd(state) {
-     case i if (i == 0) => (0,   None)
-     case i              => (i-1, Some(i))
-  }
-}
-\end{lstlisting}
-
-% nice catalysts: zip and unzip
-
-\subsection{Failure is an \textsf{Option}}
-\label{sec:failure}
-
-blocking failure versus retryable failure
-
-\begin{lstlisting}
-  val tryDec = 
-\end{lstlisting}
-
-\subsection{Advanced reagents: CAS, computed, and composed}
-\label{sec:hard-examples}
-
-Examples:
-
-\begin{itemize}
-\item TreiberStack
-\item with blocking
-\item with elimination backoff
-\item MSQueue
-\item with blocking
-\item with buckets?
-\item Counter to semaphore
-\item RWLocks
-\item Tree barriers
-\item Dining philosophers(ish)
-\item Flat combining - doesn't work if commit is two-phase process.
-  probably too much to bring into scope for this paper, anyway
-\end{itemize}
-
+\begin{figure}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\end{figure}
+
+This is where the differences from STM really start to show.
 
 post-commit action for catching up tail?
+
+\subsection{A caveat for composition}
 
@@
 Purpose of the semantics is mainly to clarify the relationship between
-the shared-state and message-passing aspects of reagents.
+the shared-state and message-passing aspects of reagents.  We have
+ruthlessly stripped out ...
 
 (fractional?) permissions model
+
+Left bias in choice operator
 
@@
 \end{itemize}
+
+Computed and lift cause problems: can potentially invoke a reagent
+inside another invocation.  Just drop these.  And probably ret too.
+Hmm.  This doesn't even allow us to express upd.  But upd itself is
+problematic for the same reasons.
+
+Could easily split out ``pure'' expressions.  Unsatisfying...  but
+this is essentially what Haskell's STM is doing.
+
+% NOTE: this leaves a nice opportunity later to give a full semantic
+% story
 
@@
 \infer
-  {r_i!v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
+  {r_i\ \lstinline{!}\ v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
   {\prod \evalCtx_i[r_i ! v_i],\ \biguplus \sigma_i \step
@@
 \label{sec:implementation}
+
+bags allow greatest degree of freedom and parallelism
 
@@
 
+The design of reagents builds on a foundation laid by Concurrent ML
+and extended in Haskell's STM
+
+\subsection{Concurrent ML}
+
+parallel CML's implementation of choice
+
+\subsection{Software transactional memory}
+
+STM in general
+
+Haskell's STM
+
+left-biased choice, but differently from \lstinline{orElse} in
+Haskell's STM.
+
+\subsection{Transactions that communicate}
+
+Communicating transactions and company, transactional events
+
+\subsection{Composing lock-free data structures}
+
+elastic transactions and company, Paper on composing lock-free data
+structure operations
+
+\section{Conclusion}
+
 We believe it is crucial to have XX, YY together
-
-STM, Communicating transactions and company, elastic transactions and company
-
-CML, Transactional events, async CML, join calculus
-
-Paper on composing lock-free data structure operations
 
@@
 unsplittability is rather more complicated than splittability.
+
+Examples:
+
+\begin{itemize}
+\item TreiberStack
+\item with blocking
+\item with elimination backoff
+\item MSQueue
+\item with blocking
+\item with buckets?
+\item Counter to semaphore
+\item RWLocks
+\item Tree barriers
+\item Dining philosophers(ish)
+\item Flat combining - doesn't work if commit is two-phase process.
+  probably too much to bring into scope for this paper, anyway
+\end{itemize}

### 2011-10-30 src/main/scala/KCASRef.scala

@@ -5,30 +5,33 @@ package chemistry
 import scala.annotation.tailrec
 import java.util.concurrent.atomic._
 
+/*
 // a k-cas-able reference cell
 class KCASRef[A <: AnyRef](init: A) {
   private[chemistry] val data = new AtomicReference[AnyRef](init)
 
   private[chemistry] def afterCAS {}
 
-  private[chemistry] final def getI = (data.get match {
+  @inline private[chemistry] final def getI = (data.get match {
     case (owner: KCAS) => owner.curVal(this)
     case v => v
   }).asInstanceOf[A]
 
-  private [chemistry] final def casI(ov: A, nv: A): Boolean = {
+  @inline private [chemistry] final def casI(ov: A, nv: A): Boolean = {
     val success = data.compareAndSet(ov, nv)
     if (success) afterCAS
     success
   }
 }
+*/
 
 private final case class CAS[A <: AnyRef](
-  ref: KCASRef[A], ov: A, nv: A
+  ref: Ref[A], ov: A, nv: A
 ) {
-  def execAsSingle: Boolean = ref.casI(ov, nv)
+  def execAsSingle: Boolean = ref.data.compareAndSet(ov, nv)
 }
 
+/*
 private object KCAS {
   private sealed abstract class Status
   private final case object Pending extends Status
@@ -100,6 +103,54 @@ private final class KCAS(val casList: List[CAS[_]]) {
 	if (complete) { rollForward(casList); true }
 	else	      { rollBackBetween(casList, Nil); false }
       case end =>     { rollBackBetween(casList, end); false }
+    }
+  }
+}
+*/
+
+private object KCAS {
+  def tryCommit(casList: List[CAS[_]]): Boolean = {
+    // attempt to place KCAS record in each CASed reference.  returns null
+    // if successful, and otherwise the point in the CAS list to rollback
+    // from.
+    @tailrec def acquire(casList: List[CAS[_]]): List[CAS[_]] = 
+      casList match {
+	case Nil => null
+	case CAS(ref, ov, nv) :: rest => ref.data.get match {
+	  case null => casList // blocking version
+	  case curVal if (curVal == ov) => 
+	    if (ref.data.compareAndSet(ov, null)) acquire(rest)
+	    else casList
+	  case _ => casList
+	}
+      }
+
+    @tailrec def rollBackBetween(start: List[CAS[_]], end: List[CAS[_]]) { 
+      if (start != end) start match {
[…]

### 2011-10-30 src/main/scala/MSQueue.scala

@@ -4,22 +4,84 @@ package chemistry
 
 import scala.annotation.tailrec
 
+/*
 final class MSQueue[A >: Null] {
   private final case class Node(data: A, next: Ref[Node] = Ref(null))
   private val head = Ref(Node(null))
-  private val tail = Ref(head.read!())
+//  private val tail = Ref(head.read!())
+  private var tail = head.read!()
 
   val enq: Reagent[A, Unit] = computed { (x:A) =>
-    @tailrec def search: Reagent[Unit,Unit] = tail.read ! () match {
-      case    Node(_, r@Ref(null)) => r.cas(null, Node(x))
-      case ov@Node(_, Ref(nv))     => tail.cas(ov,nv) !? (); search
+    val newNode = Node(x)
+    @tailrec def search: Reagent[Unit,Unit] = tail match {
+      case    Node(_, r@Ref(null)) => r.cas(null, newNode)
+      case ov@Node(_, Ref(nv))     => tail = nv; search //tail.casI(ov,nv); search
     }
     search
   }
-  val tryDeq: Reagent[Unit, Option[A]] = head.upd[Option[A]] {
-    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
-    case emp => (emp, None)
+  // val tryDeq: Reagent[Unit, Option[A]] = head.upd[Option[A]] {
+  //   case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
+  //   case emp => (emp, None)
+  // }
+  object tryDeq extends Reagent[Unit,Option[A]] {
+    @inline def tryReact(u:Unit, rx: Reaction, offer: Offer[Option[A]]): Any = 
+      head.getI match {
+	case old@Node(_, Ref(n@Node(x, _))) =>
+	  if (head.casI(old,n)) Some(x) else Retry
+	case _ => None
+      }
+    def composeI[B](next: Reagent[Option[A],B]) = throw Util.Impossible
+    def maySync = false
+    def alwaysCommits = false
+    def snoop(a: Unit) = false
   }
+
+  val deq: Reagent[Unit, A] = head.upd[A] {
+    case Node(_, Ref(n@Node(x, _))) => (n, x)
+  }
+}
+*/
+
+final class MSQueue[A >: Null] {
+  private abstract class Q
+  private final case class Node(data: A, next: Ref[Q] = Ref(Emp)) extends Q
+  private final case object Emp extends Q
[…]

### 2011-10-30 src/main/scala/Offer.scala

@@ -56,7 +56,7 @@ private final class Waiter[-A](val blocking: Boolean)
 	      extends Offer[A] with DeletionFlag {
   import Waiter._
 
-  private[chemistry] val status = new KCASRef[AnyRef](Waiting)
+  private[chemistry] val status = new Ref[AnyRef](Waiting)
 
   // the thread that *created* the Waiter
   private val waiterThread = Thread.currentThread() 
@@ -64,26 +64,27 @@ private final class Waiter[-A](val blocking: Boolean)
     if (blocking) LockSupport.unpark(waiterThread)
   }
   
-  @inline def isActive: Boolean = status.getI == Waiting
+  @inline def isActive: Boolean = status.data.get == Waiting
 
   // Poll current waiter value:
   //   - None if Waiting or Aborted
   //   - Some(ans) if completed with ans
   // sadly, have to use `Any` to work around variance problems
-  @inline def poll: Option[Any] = status.getI match {
+  @inline def poll: Option[Any] = status.data.get match {
+    case null => None
     case (_:WaiterStatus) => None
     case ans => Some(ans)
   }
   
   // Attempt to abort, returning true iff successful
-  @inline def tryAbort = status.casI(Waiting, Aborted)
+  @inline def tryAbort = status.data.compareAndSet(Waiting, Aborted)
   @inline def rxWithAbort(rx: Reaction): Reaction =
     rx.withCAS(status, Waiting, Aborted)
 
   def abortAndWake = if (tryAbort) wake()
 
   @inline def tryComplete(a: A) = 
-    status.casI(Waiting, a.asInstanceOf[AnyRef])
+    status.data.compareAndSet(Waiting, a.asInstanceOf[AnyRef])
   @inline def rxWithCompletion(rx: Reaction, a: A): Reaction = 
     rx.withCAS(status, Waiting, a.asInstanceOf[AnyRef])
 

### 2011-10-30 src/main/scala/Reaction.scala

@@ -5,8 +5,8 @@ package chemistry
 
 private sealed class Reaction private (
   val casList: List[CAS[_]],		// k-cas built up so far
-  val pcList: List[Unit => Unit],	// post-commit actions
-  val msgSet: HashSet[Message[_,_,_]]   // messages intended for consumption
+  val pcList: List[Unit => Unit]	// post-commit actions
+//  val msgSet: HashSet[Message[_,_,_]]   // messages intended for consumption
 ) {
   import Reaction._
   
@@ -23,7 +23,7 @@ private sealed class Reaction private (
 
   def withPostCommit(postCommit: Unit => Unit): Reaction =
     new Reaction(casList, postCommit +: pcList)
-  def withCAS[A <: AnyRef](ref: KCASRef[A], ov: A, nv: A): Reaction =
+  def withCAS[A <: AnyRef](ref: Ref[A], ov: A, nv: A): Reaction =
     new Reaction(CAS(ref, ov, nv) +: casList, pcList)
 
   def ++(rx: Reaction): Reaction = 
@@ -33,7 +33,7 @@ private sealed class Reaction private (
     val success: Boolean = casCount match {
       case 0 => true
       case 1 => casList.head.execAsSingle
-      case _ => new KCAS(casList).tryCommit
+      case _ => KCAS.tryCommit(casList)
     }
     if (success)
       pcList.foreach(_.apply())  // perform the post-commit actions

### 2011-10-30 src/main/scala/Reagent.scala

@@ -166,7 +166,7 @@ object never extends Reagent[Any, Nothing] {
   def alwaysCommits = false
   def maySync = false
 }
-
+/*
 object computed {
   private final case class Computed[A,B,C](c: A => Reagent[Unit,B], 
 					   k: Reagent[B,C]) 
@@ -180,6 +180,20 @@ object computed {
   }
   @inline def apply[A,B](c: A => Reagent[Unit,B]): Reagent[A,B] = 
     Computed(c, Commit[B]())
+}
+*/
+object computed {
+  private final case class Computed[A,B](c: A => Reagent[Unit,B]) 
+		     extends Reagent[A,B] {
+    def snoop(a: A) = false
+    def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
+      c(a).tryReact((), rx, offer)
+    def composeI[C](next: Reagent[B,C]) = throw Util.Impossible
+    def alwaysCommits = false
+    def maySync = true
+  }
+  @inline def apply[A,B](c: A => Reagent[Unit,B]): Reagent[A,B] = 
+    Computed(c)
 }
 
 object lift {

### 2011-10-30 src/main/scala/Ref.scala

@@ -5,10 +5,11 @@ package chemistry
 import scala.annotation.tailrec
 import java.util.concurrent.atomic._
 
-final class Ref[A <: AnyRef](init: A) extends KCASRef[A](init) {
-  private val offers = new CircularPool[Offer[_]]
+final class Ref[A <: AnyRef](init: A) {
+  private[chemistry] val data = new AtomicReference(init)
+  private[chemistry] val offers = new CircularPool[Offer[_]]
 
-  override def afterCAS {
+  private[chemistry] def afterCAS {
     @tailrec def wakeFrom(n: offers.Node): Unit = if (n != null) {
       n.data.abortAndWake
       wakeFrom(n.next)
@@ -19,12 +20,18 @@ final class Ref[A <: AnyRef](init: A) extends KCASRef[A](init) {
   private final case class Read[B](k: Reagent[A,B]) extends Reagent[Unit, B] {
     def tryReact(u: Unit, rx: Reaction, offer: Offer[B]): Any = {
       if (offer != null) offers.put(offer)
-      getI
+      data.get match {
+	case null => Retry
+	case ans => ans
+      }
     }
     def composeI[C](next: Reagent[B,C]) = Read(k >=> next)
     def maySync = k.maySync
     def alwaysCommits = k.alwaysCommits
-    def snoop(u: Unit) = k.snoop(getI)
+    def snoop(u: Unit) = data.get match {
+      case null => false
+      case ans => k.snoop(ans)
+    }
   }
   @inline def read: Reagent[Unit,A] = Read(Commit[A]())
 
@@ -34,7 +41,7 @@ final class Ref[A <: AnyRef](init: A) extends KCASRef[A](init) {
     // ref cell to the continuation k.
     def tryReact(u: Unit, rx: Reaction, offer: Offer[B]): Any = 
       if (rx.canCASImmediate(k, offer)) {
-	if (casI(expect, update))
+	if (data.compareAndSet(expect, update))
 	  k.tryReact((), rx, offer)
 	else Retry
       } else k.tryReact((), rx.withCAS(Ref.this, expect, update), offer)
@@ -53,21 +60,17 @@ final class Ref[A <: AnyRef](init: A) extends KCASRef[A](init) {
[…]

### 2011-10-30 src/main/scala/SimpleQueue.scala

@@ -1,0 +1,41 @@
+package chemistry
+
+import scala.collection.immutable._
+
+final class SimpleQueue[A] {
+  private val head = Ref(Queue.empty[A])
+  object enq extends Reagent[A,Unit] {
+    def tryReact(x:A, rx: Reaction, offer: Offer[Unit]): Any = {
+      val cur = head.data.get
+      if (cur eq null) 
+	Retry
+      else {
+	val upd = cur.enqueue(x)
+	if (head.data.compareAndSet(cur,upd)) () 
+	else Retry
+      }
+    }
+    def composeI[B](next: Reagent[Unit,B]) = throw Util.Impossible
+    def maySync = false
+    def alwaysCommits = false
+    def snoop(a: A) = false
+  }
+
+  object tryDeq extends Reagent[Unit,Option[A]] {
+    @inline def tryReact(u:Unit, rx: Reaction, offer: Offer[Option[A]]): Any = {
+      val cur = head.data.get
+      if (cur eq null)
+	Retry
+      else if (cur.isEmpty)
+	None
+      else {
+	val (x, rest) = cur.dequeue
+	if (head.data.compareAndSet(cur,rest)) Some(x) else Retry
+      }
+    }
+    def composeI[B](next: Reagent[Option[A],B]) = throw Util.Impossible
+    def maySync = false
+    def alwaysCommits = false
+    def snoop(a: Unit) = false
+  }
+}

### 2011-10-30 src/main/scala/TreiberStack.scala

@@ -4,8 +4,39 @@ package chemistry
 
 import scala.annotation.tailrec
 import java.util.concurrent.atomic._
+import scala.collection.immutable._
 
 final class TreiberStack[A >: Null] {
+/*
+    new Reagent[B,C] {
+      private val k = Commit[C]()
+      def tryReact(b: B, rx: Reaction, offer: Offer[C]): Any = {
+	if (rx.canCASImmediate(k, offer)) {
+	  // no need to store offer here, as we will either succeed or retry
+	  // (never block)
+
+	  val ov = r.getI
+	  val (nv, retVal) = f(ov, b)
+	  if (r.casI(ov, nv)) retVal
+	  else Retry
+	} else {
+//	  if (offer != null) offers.put(offer) 
+
+	  val ov = r.getI
+	  val (nv, retVal) = f(ov, b)
+	  k.tryReact(retVal, rx.withCAS(r, ov, nv), offer)
+	}
+      }
+      def composeI[D](next: Reagent[C,D]) = throw Util.Impossible/*
+	new r.InnerUpd[B,C,D](next) {
+	  @inline def newValue(a: A, b: B): A = f(a,b)._1
+	  @inline def retValue(a: A, b: B): C = f(a,b)._2
+	}*/
+      def maySync = false
+      def alwaysCommits = false
+      def snoop(b: B) = false
+    }
+    */
 
 /*
   val push: Reagent[A,Unit] = head.fastUpd[A,Unit](
@@ -20,30 +51,31 @@ final class TreiberStack[A >: Null] {
       case (emp,   _) => None }
   )
 */
+
 /*
   private val head = Ref[List[A]](List())
 
-  val push: Reagent[A,Unit] = head.updIn { 
-    (xs,x) => x::xs
-  }
+  val push: Reagent[A,Unit] = upd.fast[List[A],A,Unit](head, { 
+    (xs,x) => (x::xs, ())
+  })
 
-  val tryPop: Reagent[Unit,Option[A]] = head.upd[Option[A]] {
-    case (x::xs) => (xs,  Some(x))
-    case emp     => (emp, None)
-  }
+  val tryPop: Reagent[Unit,Option[A]] = upd.fast[List[A],Unit,Option[A]](head, {
+    case ((x::xs),()) => (xs,  Some(x))
+    case (Nil, ())    => (Nil, None)
+  })
 
   val pop: Reagent[Unit,A] = head.upd[A] {
[…]

### 2011-10-30 src/main/scala/bench/Bench.scala

@@ -251,7 +251,8 @@ object Bench extends App {
 //    List(100, 250, 500)
 //    List(0) ++ (for (i <- 0 to 15) yield pow(10, 1+i.toDouble * 0.25).toInt)
 //    (0 to 600 by 50)
-    (50 to 1000 by 50).toList
+    List(100, 1000)
+//    (50 to 1000 by 50).toList
   else config.workList = config.workList.reverse
 
   log("Beginning benchmark")
@@ -263,7 +264,7 @@ object Bench extends App {
   } yield (b, 0, 1, 1)
   private val concBenches = for {
 //    b <- List(PushPop, EnqDeq, IncDec)
-    b <- List(PushPop)
+    b <- List(PushPop, EnqDeq)
     w <- config.workList
   } yield (b, w, config.minCores, config.maxCores)
 

### 2011-10-30 src/main/scala/bench/benchmarks/EnqDeq.scala

@@ -31,10 +31,38 @@ object EnqDeq extends Benchmark {
       }
     }
   } 
+  private object stm extends Entry {
+    def name = "stm"
+    type S = STMQueue[AnyRef]
+    def setup = new STMQueue()
+    def run(q: S, work: Int, iters: Int) {
+      val r = new Random
+      for (_ <- 1 to iters) {
+	q.enq(SomeData)
+	Util.noop(r.fuzz(work))
+	untilSome {q.tryDeq}
+	Util.noop(r.fuzz(work))
+      }
+    }
+  } 
   private object reagent extends Entry {
     def name = "reagent"
     type S = MSQueue[AnyRef]
     def setup = new MSQueue()
+    def run(q: S, work: Int, iters: Int) {
+      val r = new Random
+      for (_ <- 1 to iters) {
+	q.enq ! SomeData;
+	Util.noop(r.fuzz(work))
+	untilSome {q.tryDeq ! ()}
+	Util.noop(r.fuzz(work))
+      }
+    }
+  }
+  private object simple extends Entry {
+    def name = "simpleR"
+    type S = SimpleQueue[AnyRef]
+    def setup = new SimpleQueue()
     def run(q: S, work: Int, iters: Int) {
       val r = new Random
       for (_ <- 1 to iters) {
@@ -145,7 +173,7 @@ object EnqDeq extends Benchmark {
   }
 */
 
-  def entries = List(reagent, lagging, optimistic, hand, juc, lock)
+  def entries: List[Entry] = List(stm) //, simple, reagent, hand, juc)
   // fc livelocks
   // basket is not licensed for distribution
 }

### 2011-10-30 src/main/scala/bench/benchmarks/PushPop.scala

@@ -24,15 +24,19 @@ object PushPop extends Benchmark {
 	while (n < iters) {
 	  n += 1
 	  push(s, SomeData)
+//	  push(s, SomeData)
 	  Util.noop(r.fuzz(work))
 	  untilSome { tryPop(s) }
+//	  untilSome { tryPop(s) }
 	  Util.noop(r.fuzz(work))
 	}
       } else {
 	while (n < iters) {
 	  n += 1
 	  push(s,SomeData)
+//	  push(s,SomeData)
 	  untilSome { tryPop(s) }
+//	  untilSome { tryPop(s) }
 	}
       }
 
@@ -75,7 +79,7 @@ object PushPop extends Benchmark {
     while (n < iters) {
       n += 1
       Util.noop(r.fuzz(work))
-      Util.noop(r.fuzz(work))
+//      Util.noop(r.fuzz(work))
     }
   }
   private object handElim extends Entry with Generic {
@@ -112,6 +116,13 @@ object PushPop extends Benchmark {
     def name = "hand"
     type S = HandStack[AnyRef]
     def setup = new HandStack()
+    def push(s: S, x: AnyRef) = s.push(x)
+    def tryPop(s: S): Option[AnyRef]  = s.tryPop
+  } 
+  private object stm extends Entry with Generic{
+    def name = "stm"
+    type S = STMStack[AnyRef]
+    def setup = new STMStack()
     def push(s: S, x: AnyRef) = s.push(x)
     def tryPop(s: S): Option[AnyRef]  = s.tryPop
   } 
@@ -180,5 +191,5 @@ private object rTreiberFast extends Entry with Generic {
 //  def entries = List(rTreiber, rElim, hand)
 //  def entries = List(rElim, rTreiber, handElim, hand, handPool)
 //  def entries: List[Entry] = List(rElim, handPool, handElim)
-  def entries: List[Entry] = List(handAloc, hand)
+  def entries: List[Entry] = List(stm) //List(rTreiber, stm, hand)
 }

### 2011-10-30 src/main/scala/bench/benchmarks/Transfer.scala

@@ -1,0 +1,22 @@
+package chemistry.bench.benchmarks
+
+import chemistry._
+import chemistry.bench._
+import chemistry.bench.competition._
+import chemistry.Util._
+/*
+object Transfer extends Benchmark {
+  def pureWork(work: Int, iters: Int) = {
+    val r = new Random
+    for (_ <-1 to iters) {
+      Util.noop(r.fuzz(work * 20))
+    }
+  }
+
+  private object reagent extends Entry {
+    def name = "reagent"
+    type S = 
+  }
+}
+
+*/

### 2011-10-30 src/main/scala/bench/competition/HandStack.scala

@@ -2,31 +2,34 @@ package chemistry.bench.competition
 
 import java.util.concurrent.atomic._
 import scala.annotation.tailrec
+import scala.collection.immutable._
 
 // Standard lock-free stack, due to Treiber.  Doesn't yet perform
 // exponential backoff.
 class HandStack[A >: Null] {
-  class Node(val data: A, var next: Node) 
-
   // head always points to top of stack,
   //   from which the rest of the stack is reachable
-  private val head = new AtomicReference[Node](null)
+  private val head = new AtomicReference[List[A]](Nil)
 
   def push(x: A) {
-    val n = new Node(x, null)
+    val backoff = new chemistry.Backoff
     while (true) {
-      n.next = head.get
-      if (head compareAndSet (n.next, n)) return
+      val cur = head.get
+      val upd = x +: cur
+      if (head compareAndSet (cur, upd)) return
+      backoff.once
     } 
   }
 
   def tryPop(): Option[A] = {
+    val backoff = new chemistry.Backoff
     while (true) {
-      val h = head.get
-      if (h eq null) 
+      val cur = head.get
+      if (cur.isEmpty) 
 	return None
-      else if (head compareAndSet (h, h.next)) 
-	return Some(h.data) 
+      else if (head.compareAndSet(cur,cur.tail)) 
+	return Some(cur.head)      
+      backoff.once
     }
     throw new Exception("Impossible")
   } 

### 2011-10-30 src/main/scala/bench/competition/STMQueue.scala

@@ -1,0 +1,21 @@
+package chemistry.bench.competition
+
+import akka.stm._
+import scala.collection.immutable._
+
+class STMQueue[A] {
+  private val state = Ref(Queue.empty[A])
+  def enq(x: A) = atomic { 
+    state alter (_.enqueue(x))
+  }
+  def tryDeq() = atomic {
+    val cur = state.get
+    if (cur.isEmpty)
+      None
+    else {
+      val (x, xs) = cur.dequeue
+      state.set(xs)
+      Some(x)
+    }
+  }
+}

### 2011-10-30 src/main/scala/bench/competition/STMStack.scala

@@ -1,0 +1,20 @@
+package chemistry.bench.competition
+
+import akka.stm._
+import scala.collection.immutable._
+
+class STMStack[A] {
+  private val state = Ref[List[A]](Nil)
+  def push(x: A) = atomic {
+    state alter (x +: _)
+  }
+  def tryPop(): Option[A] = atomic {
+    val cur = state.get
+    if (cur.isEmpty)
+      None
+    else {
+      state.set(cur.tail)
+      Some(cur.head)
+    }
+  }
+}

### 2011-10-30 paper/paper.tex
more work on reagents section

@@
   cache miss per operation.
-\item It is susceptible to ... % blocking %TODO
+\item It is susceptible to unfortunate preemptions or stalls on the part of a
+  thread holding a lock, which prevents other threads from making progress.
 \end{itemize}
@@
 Memory is shared between reagents using the type \lstinline{Ref[A]} of
-atomically-updatable references.  These references are only accessible
-through reagent combinators, which allows them to provide
-functionality beyond the built-in \lstinline{AtomicReference[A]} type,
-as we will see.class TreiberStack[A] {
+atomically-updatable references.\footnote{ These references are only
+  accessible through reagent combinators, which allows them to provide
+  functionality beyond the built-in \lstinline{AtomicReference[A]} type, such
+  as the blocking logic we will see shortly.  }
+The \lstinline{upd} combinator~(\figref{combinators}) represents simple atomic
+updates to references.  It takes an \emph{update function}, which tells how to
+transform a snapshot of the reference cell and some input into an updated
+value for the cell and some output.  Using \lstinline{upd}, we can rewrite
+\lstinline{TreiberStack} in a more readable and concise way:
+\begin{lstlisting}
+class TreiberStack[A] {
   private val head = new Ref[List[A]](Nil)
@@
 In Scala, anonymous partial functions are written as a series of cases
-enclosed in braces.  For \lstinline{push} and \lstinline{tryPop}, the
-case analyses are exhaustive, so the functions are in fact total.
+enclosed in braces.  For \lstinline{push} and \lstinline{tryPop}, the case
+analyses are exhaustive, so the update functions are in fact total.
 
-Being reagents, \lstinline{push} and \lstinline{tryPop} are inert
-values.  They can be invoked using the \lstinline{!} method provided
-by reagents; for a \lstinline{Reagent[A,B]} the \lstinline{!} method
-takes an \lstinline{A} and returns a \lstinline{B}.  Scala permits
-infix notation for methods, so we can use a \lstinline{TreiberStack}
-as follows:
+Being reagents, \lstinline{push} and \lstinline{tryPop} are inert values.
+They can be invoked using the \lstinline{!} method provided by reagents; for a
+\lstinline{Reagent[A,B]} the \lstinline{!} method takes an \lstinline{A} and
+returns a \lstinline{B}.  Scala permits infix notation for methods, so we can
+use a \lstinline{TreiberStack} as follows:
 \begin{lstlisting}
@@
 The key point is that when we invoke these reagents, we are executing
-\emph{exactly} the same algorithms written by hand in
-\secref{background}, including the retry loop with exponential
-backoff; reagents systematize and internalize common patterns of
-fine-grained concurrency.  But by exposing \lstinline{push} and
-\lstinline{tryPop} as reagents, rather than methods, we leave the door
-open for further composition and tailoring by a user of the data
-structure, as we shall see below.
+\emph{exactly} the same algorithms written by hand in \secref{background},
+including the retry loop with exponential backoff; reagents systematize and
+internalize common patterns of fine-grained concurrency.  But by exposing
+\lstinline{push} and \lstinline{tryPop} as reagents, rather than methods, we
+leave the door open for further composition and tailoring by a user of the
+data structure (\secref{choice}, \secref{conjunction}).
 
@@
 \end{lstlisting}
-Now our update function really is partial.  An invocation
-\lstinline{s.pop ! ()} will block the calling thread unless or until
-the stack is nonempty.  Blocking and signalling are entirely handled
-by the reagent implementation, using standard
-techniques~(\secref{implementation}); there is therefore no risk of a
-lost wakeup.  %TODO: add more here?
+Now our update function is partial.  An invocation \lstinline{s.pop ! ()} will
+block the calling thread unless or until the stack is nonempty.  Blocking and
+signalling are entirely handled by the reagent implementation, using standard
+techniques~(\secref{implementation}); there is therefore no risk of a lost
+wakeup.  %TODO: add more here?
 
@@
 threads are made mutually aware of reaching a particular state.
-Synchronization is completementary to shared state: the former
+Synchronization is complementary to shared state: the former
 provides concurrent interaction, while the latter (through atomic
@@
 \subsection{Composing reagents: choice}
+\label{sec:choice}
 
@@
 \subsection{Composing reagents: sequencing and pairing}
+\label{sec:conjunction}
 
@@
 The implementation details for conjunctions are discussed
-later~(\secref{implementation}), but a key point is that the
-performance cost is \emph{pay as you go}.  Single atomic reagents like
-\lstinline{push} and \lstinline{pop} execute a single CAS---just like
-the standard nonblocking algorithms they are meant to implement---even
-though these operations can be combined into larger atomic operations.
-The cost of conjunction is only incurred when a conjoined reagent is
-actually used.  This is a crucial difference from STM, which must
-generally incur overheads regardless of the size of the atomic blocks;
-see~\secref{related} for more discussion.
+later~(\secref{implementation}), but a key point is that the performance cost
+is \emph{pay as you go}.  Single atomic reagents like \lstinline{push} and
+\lstinline{pop} execute a single CAS---just like the standard nonblocking
+algorithms they are meant to implement---even though these operations can be
+combined into larger atomic operations.  The cost of conjunction is only
+incurred when a conjoined reagent is actually used.  This is a crucial
+difference from STM, which generally incurs overheads regardless of the size
+of the atomic blocks; see~\secref{related} for more discussion.
 
@@
 
-In chemistry, a reagent is a participant in a reaction, and reagents
-are further subdivided into \emph{reactants}, which are consumed
-during reaction, and \emph{catalysts}, which enable reactions but are
-not consumed by them.  Invoking a reagent using the \lstinline{!}
-operator
+In chemistry, a reagent is a participant in a reaction, and reagents are
+further subdivided into \emph{reactants}, which are consumed during reaction,
+and \emph{catalysts}, which enable reactions but are not consumed by them.
 
-chemistry metaphor.  note that catalysts increase parallelism and remove need
-for devoted thread
+The \lstinline{!} operator invokes a reagent as a reactant: the invocation
+lasts for a single reaction, and any messages sent by the reagent are
+consumed.  But sometimes it is useful for reagents to persist beyond a single
+reaction, \emph{i.e.}, to act as catalysts.  For example, the following
+function creates a catalyst that merges input from two endpoints and sends the
+resulting pairs to another endpoint:
+\begin{lstlisting}
+def zip(in1: Endpoint[Unit,  A], 
+        in2: Endpoint[Unit,  B], 
+        out: Endpoint[(A,B), Unit]) = 
+  dissolve((swap(in1) ||| swap(in2)) >> swap(out))
+\end{lstlisting}
+The \lstinline{dissolve} function takes a \lstinline{Reagent[Unit, Unit]} and
+introduces it as a catalyst.\footnote{For simplicity, we have not given a way
+  to \emph{cancel} catalysts after they have been introduced, but cancellation
+  is easy to add.}  Operationally, in this example, that just means sending
+messages along \lstinline{in1} and \listinline{in2} that are marked as
+catalyzing messages, and hence are not consumed during reaction.  The upshot
+is that senders along \lstinline{in1} will see the catalyzing messages, look
+for messages along \lstinline{in2} to pair with, and ultimately send messages
+along \lstinline{out} (and similarly in the other order).
 
-nice catalysts: zip and unzip, tree barrier
+Catalysts could instead be expressed using a thread that repeatedly invokes a
+reagent as a reactant.  Allowing direct expression through
+\lstinline{dissolve} is more efficient (since it does not tie up a thread) and
+allows greater parallelism (since, as with the \lstinline{zip} example above,
+multiple reagents can react with it in parallel).
+
+Catalysts are not limited to message passing.  The \lstinline{zip} example
+above could be rephrased to take reagents, rather than endpoints, as
+arguments.
+
+%TODO: barrier?
+%TODO: explain Unit -> Unit?
+
+% nice catalysts: zip and unzip, tree barrier
 
@@
 }
-
-\begin{lstlisting}
-def zip(in1: Endpoint[Unit,  A], 
-        in2: Endpoint[Unit,  B], 
-        out: Endpoint[(A,B), Unit]) = 
-  dissolve((swap(in1) ||| swap(in2)) >> out)
-\end{lstlisting}
-
-can be made more abstract: use reagent arguments instead!
-
 \subsection{Dynamic reagents and \lstinline{cas}}
 
-The combinators introduced so far are powerful, but they impose a
-strict phase separation: reagents are constructed prior to, and
-independently from, the data that flows through them.  Phase
-separation is useful, because it allows reagent execution to be
-optimized based on complete knowledge of the computation to be
-performed (see \secref{implementation}).
+The combinators introduced so far are powerful, but they impose a strict phase
+separation: reagents are constructed prior to, and independently from, the
+data that flows through them.  Phase separation is useful, because it allows
+reagent execution to be optimized based on complete knowledge of the
+computation to be performed (see \secref{implementation}).  But in many cases
+the structure of the computation to be performed depends on the input or other
+dynamic data.
 
+\begin{figure}
+\begin{lstlisting}
+class MSQueue[A] {
+  private case class Node(data: A, next: Ref[Node])
+  private val head = new Ref(Node(null))
+  private val tail = new Ref(head.read ! ())
+  val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
+    case (Node(_, Ref(n@Node(x, _))), ()) => (n, Some(x))
+    case (emp,                        ()) => (emp, None)
+  }
+  private def findAndEnq(n: Node): Reagent[Unit,Unit] = 
+    read(tail) ! () match {
+      case ov@Node(_, r@Ref(null)) => 
+        cas(r, null, n) >>
+        postCommit { _ => cas(tail, ov, n) !? () }
+      case ov@Node(_, Ref(nv)) => 
+        cas(tail, ov, nv) !? (); 
+        findAndEnq(n)
+    }
+  val enq: Reagent[A, Unit] = computed { 
+    (x: A) => findAndEnq(Node(x, new Ref(null)))
+  }
+}
+\end{lstlisting}
+\nocaptionrule
+\caption{The Michael-Scott queue, with reagents}
+\label{fig:msqueue}
+\end{figure}
+
+An example of this phenomenon is the classic Michael-Scott lock-free
+queue~\cite{?}.
+
+\elide{
 The \lstinline{read} combinator has a straightforward signature: if
@@
 a complete discussion until \secref{failure}.
+}
 
-\begin{figure}
-\begin{lstlisting}
-class MSQueue[A] {
-  private case class Node(data: A, next: Ref[Node])
-  private val head = new Ref(Node(null))
-  private val tail = new Ref(head.read!())
-
-  @tailrec private def search(x:A): Reagent[Unit,Unit] = 
-    read(tail) ! () match {
-      case Node(_, r@Ref(null)) => 
-        cas(r, null, Node(x, new Ref(null)))
-      case ov@Node(_, Ref(nv)) => 
-        cas(tail, ov, nv) !? (); 
-        search(x)
-    }
-  val enq: Reagent[A, Unit] = computed search
-  val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
-    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
-    case emp => (emp, None)
-  }
-}
-\end{lstlisting}
-\end{figure}
+% tail recursive
 
@@
 
-post-commit action for catching up tail?
+post-commit action for catching up tail
 
-\subsection{A caveat for composition}
+\paragraph{A caveat for composition}
 
@@
 1.0 Background
-2.5 Library overview & examples
+2.5 Library overview & examples         3.0
 1.5 Semantics 
-1.5 Implementation/algorithms 
-0.5 Performance results
+1.5 Implementation/algorithms           3.0
+0.5 Performance results                 1.0
 1.0 Related work

### 2011-10-31 paper/paper.tex
filled in background section; dropped semantics

@@
   cache miss per operation.
-\item It is susceptible to unfortunate preemptions or stalls on the part of a
+\item It is susceptible to preemptions or stalls of a
   thread holding a lock, which prevents other threads from making progress.
@@
 
-To combat scalability problems, data structures employ finer-grained locking,
-or eschew locking altogether.  Fine-grained locking associates locks with
-small, independent parts of a data structure, allowing those parts to be
-manipulated in parallel.  Lockless (or \emph{nonblocking}) data structures
-instead perform updates directly, using hardware-level operations (like
-compare-and-set) to ensure atomicity.  Doing the updates directly means that
-there is no extra communication or contention for locks, but it also generally
-means that the entire update must consist of changing a single word of
-memory---a constraint which is often quite challenging to meet.
+To achieve scalability, data structures employ finer-grained locking,
+or eschew locking altogether.  Fine-grained locking associates locks
+with small, independent parts of a data structure, allowing those
+parts to be manipulated in parallel.  Lockless (or \emph{nonblocking})
+data structures instead perform updates directly, using hardware-level
+operations (like compare-and-set) to ensure atomicity.  Doing the
+updates directly means that there is no extra communication or
+contention for locks, but it also generally means that the entire
+update must consist of changing a single word of memory---a constraint
+which is often quite challenging to meet.
 
@@
 
-\figref{classic-treiber} gives a classic example of a fine-grained
-concurrent data structure: Treiber's lock-free stack~\cite{?}.  The
-stack is represented as an \emph{immutable} linked list.  Mutation
-occurs solely through the \lstinline{head} pointer, represented here
-as an \lstinline{AtomicReference}.  The \lstinline{head} is mutated
-using \lstinline{compareAndSet} (here abbreviated as \lstinline{cas}),
-which takes an expected value and a new value, and atomically updates
-the reference
-
-% OPTIMISTIC CONCURRENCY
+\figref{classic-treiber} gives a classic example of a fine-grained concurrent
+data structure: Treiber's lock-free stack~\cite{?}.  The stack is represented as
+an \emph{immutable} linked list.  Mutation occurs solely through the
+\lstinline{head} pointer, represented here as an \lstinline{AtomicReference}.
+The \lstinline{head} is mutated using \lstinline{compareAndSet} (here
+abbreviated as \lstinline{cas}), which takes an expected value and a new value,
+and atomically updates the reference if it has the expected value.  It returns
+\lstinline{true} iff the update was successful.
+
+A word about Scala: the syntax 
+\begin{lstlisting}
+  exp match { case pat1 => body1 ... case patN => bodyN }
+\end{lstlisting}
+denotes pattern matching, where the value of \lstinline{exp} is matched in order
+against \lstinline{pat1} up to \lstinline{patN}.  In \figref{classic-treiber},
+we use the two list constructors \lstinline{Nil} and \lstinline{::} (an inline
+cons) in patterns.
+
+The \lstinline{push} and \lstinline{tryPop} operations are implemented in a
+typical \emph{optimistic} style: they take a snapshot of the head, perform some
+local computation with it, and then attempt to update it accordingly.  The
+computation is optimistic because it is performed without holding a lock.  The
+head might concurrently change, invalidating the \lstinline{cas} intended to
+update it.  To account for possible interference, the snapshot-compute-update
+code executes within a retry loop.  While the loop may retry forever, it can
+only do so by repeatedly failing to \lstinline{cas}, which in turn means that an
+unlimited number of other operations on the stack are succeeding.  In other
+words, \lstinline{push} and \lstinline{tryPop} are formally \emph{lock
+  free}~\cite{?}.
+
+To successfully \lstinline{cas}, a core must acquire the relevant cache line in
+\emph{exclusive} mode~\cite{?}, or something conceptually equivalent.  When
+several cores attempt to \lstinline{cas} a common reference, they must
+coordinate at the hardware level, using precious memory bandwidth and wasting
+processor cycles.  A failed \lstinline{cas} is evidence that there is
+\emph{contention} over a common location.  To increase the \lstinline{cas}
+success rate and conserve memory bandwidth, fine-grained concurrent algorithms
+employ a \emph{backoff} scheme, which here we have abstracted into a lightweight
+\lstinline{Backoff} class.  The class encapsulates the simplest scheme:
+busy-wait for a random amount of time that grows exponentially each time
+\lstinline{once} is invoked.
 
@@
 
-%% Say something about CONTENTION.
-
 Treiber's stack scales better than a lock-based stack mainly because it
-decreases the amount of shared data that must be updated per operation:
-instead of acquiring a shared lock, altering a shared stack pointer, and
-releasing the shared lock, Treiber's stack does a single CAS to update the
-shared pointer directly.  But the head pointer of the stack is still a
-centralized source of contention.  Spreading out the contention requires
-parallelizing stack operations---but how, given that they all involve
-modifying the top of the stack?
-
-Parallelization requires a change of perspective.  Normally, we view
-concurrent operations on the same part of a data structure as competing to
-atomically update that data structure; we want the operations to be isolated.
-But if
+decreases the amount of shared data that must be updated per operation: instead
+of acquiring a shared lock, altering a shared stack pointer, and releasing the
+shared lock, Treiber's stack does a single CAS to update the shared pointer
+directly.  However, that pointer is still a centralized source of contention.
+While exponential backoff helps relieve the contention, we can do better by
+parallelizing stack operations---which is counterintuitive, given that they all
+involve modifying the top of the stack.
+
+Parallelization requires a change of perspective.  Normally, we view concurrent
+operations on the same part of a data structure as competing to atomically
+update that data structure; we want the operations to be isolated.  On the other
+hand, sometimes operations can ``help'' each other: a \lstinline{push} and a
+\lstinline{tryPop} effectively cancel each other out.  This insight leads to a
+scheme called \emph{elimination backoff}~\cite{?}.  Operations first try the
+usual \lstinline{cas}-based code.  If the \lstinline{cas} fails, rather than
+busy-waiting, the operations advertise their presence on a side-channel,
+reducing the contention for the head pointer.  If a \lstinline{push} and
+\lstinline{tryPop} detect their mutual presence, the \lstinline{push} can pass
+its argument directly to \lstinline{tryPop} through the side-channel, and no
+change to the head pointer is necessary.  Atomicity is not violated, because had
+the \lstinline{push} and \lstinline{pop} executed in sequence, the head pointer
+would have been returned to its original value anyway.
+
+Both \lstinline{push} and \lstinline{tryPop} are \emph{total} operations: they
+can in principle succeed no patter what state the stack is in, and fail (and
+retry) only due to active interference from concurrent threads.  A true
+\lstinline{pop} operation, on the other hand, is \emph{partial}: it is undefined
+when the stack is empty.  Often this is taken to mean that the operation should
+block until another thread moves the stack into a state on which the operation
+is defined.  Partial operations introduce considerable complexity, because
+\emph{all} of the operations on the data structure must potentially signal
+blocked threads, depending on the changes being performed.  In some cases, it is
+possible to cleverly treat signalling in essentially the same way as atomic
+updates~\cite{?}.
+
+Synchronization and signalling is also subject to cache and memory bandwidth
+concerns.  For example, when a lock is released, how much memory traffic is
+generated?  If the lock is implemented as a single bit which threads attempt to
+\lstinline{cas} in a tight loop, any change to that bit will result in a cascade
+of memory traffic---even though only one waiting thread can actually win and
+acquire the lock.  The most important strategy for cutting down this traffic was
+introduced by Mellor-Crummey and Scott~\cite{?}, who suggested that waiting
+threads should place themselves in a lock-free queue, and the release operation
+should  % TODO: room for this??
 
@@
 
-discuss the insight behind MCS locks
+%discuss the insight behind MCS locks
+
+The goal of reagents is to systematize these techniques and others into a
+uniform set of combinators that make algorithms easier to express, and allows
+algorithms to be composed and tailored by users.  In the next section, we will
+see how to write all of the above examples, concisely and composably.
 
@@
 }
-
-The goal of reagents is to systematize the expression of these
-disparate techniques into a uniform set of combinators
 
@@
 
-\section{Semantics}
-\label{sec:semantics}
-
-Try to keep this minimal for the sake of space
-
-Purpose of the semantics is mainly to clarify the relationship between
-the shared-state and message-passing aspects of reagents.  We have
-ruthlessly stripped out ...
-
-(fractional?) permissions model
-
-Left bias in choice operator
-
-\begin{itemize}
-  \item read, cas
-  \item send
-  \item choice
-  \item react
-  \item bind and ret
-  \item computed?
-\end{itemize}
-
-Computed and lift cause problems: can potentially invoke a reagent
-inside another invocation.  Just drop these.  And probably ret too.
-Hmm.  This doesn't even allow us to express upd.  But upd itself is
-problematic for the same reasons.
-
-Could easily split out ``pure'' expressions.  Unsatisfying...  but
-this is essentially what Haskell's STM is doing.
-
-% NOTE: this leaves a nice opportunity later to give a full semantic
-% story
-
-\elide{
-\[
-\infer
-  {P_i, \sigma_i \cstep{t_i} P'_i, \sigma'_i}
-  {P_1 | \cdots | P_n,\ \sigma_1 \uplus \cdots \uplus \sigma_n\ \step\
-   P'_1 | \cdots | P'_n,\ \sigma'_1 \uplus \cdots \uplus \sigma'_n
-  }
-\]
-
-\[
-\infer
-  {r_i!v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
-  {\procCtx[\prod \evalCtx_i[r_i ! v_i]],\ \biguplus \sigma_i \step
-   \procCtx[\prod \evalCtx_i[\ret{v'_i}]],\ \biguplus \sigma'_i}
-\]
-}
-
-\[
-\infer
-  {r_i\ \lstinline{!}\ v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
-  {\prod \evalCtx_i[r_i ! v_i],\ \biguplus \sigma_i \step
-   \prod \evalCtx_i[\ret{v'_i}],\ \biguplus \sigma'_i}
-\]
-
 \section{Implementation}
@@
 
-1.2 Title, abstract, intro, 
-1.0 Background
+1.2 Title, abstract, intro,             1.0
+1.0 Background                          1.0
 2.5 Library overview & examples         3.0
 1.5 Semantics 
-1.5 Implementation/algorithms           3.0
+1.5 Implementation/algorithms           2.0
 0.5 Performance results                 1.0
-1.0 Related work
-0.8 Bib
+1.0 Related work                        1.0
+0.3 Conclusion                          0.2
+0.8 Bib                                 0.8
 
@@
 \end{itemize}
+
+\elide{
+\section{Semantics}
+\label{sec:semantics}
+
+Try to keep this minimal for the sake of space
+
+Purpose of the semantics is mainly to clarify the relationship between
+the shared-state and message-passing aspects of reagents.  We have
+ruthlessly stripped out ...
+
+(fractional?) permissions model
+
+Left bias in choice operator
+
+\begin{itemize}
+  \item read, cas
+  \item send
+  \item choice
+  \item react
+  \item bind and ret
+  \item computed?
+\end{itemize}
+
+Computed and lift cause problems: can potentially invoke a reagent
+inside another invocation.  Just drop these.  And probably ret too.
+Hmm.  This doesn't even allow us to express upd.  But upd itself is
+problematic for the same reasons.
+
+Could easily split out ``pure'' expressions.  Unsatisfying...  but
+this is essentially what Haskell's STM is doing.
+
+% NOTE: this leaves a nice opportunity later to give a full semantic
+% story
+
+\elide{
+\[
+\infer
+  {P_i, \sigma_i \cstep{t_i} P'_i, \sigma'_i}
+  {P_1 | \cdots | P_n,\ \sigma_1 \uplus \cdots \uplus \sigma_n\ \step\
+   P'_1 | \cdots | P'_n,\ \sigma'_1 \uplus \cdots \uplus \sigma'_n
+  }
+\]
+
+\[
+\infer
+  {r_i!v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
+  {\procCtx[\prod \evalCtx_i[r_i ! v_i]],\ \biguplus \sigma_i \step
+   \procCtx[\prod \evalCtx_i[\ret{v'_i}]],\ \biguplus \sigma'_i}
+\]
+}
+
+\[
+\infer
+  {r_i\ \lstinline{!}\ v_i, \sigma_i \cstep{t_i} \ret{v'_i}, \sigma'_i}
+  {\prod \evalCtx_i[r_i ! v_i],\ \biguplus \sigma_i \step
+   \prod \evalCtx_i[\ret{v'_i}],\ \biguplus \sigma'_i}
+\]
+}

### 2011-10-31 paper/paper.tex
work on bib

@@
 
+kcas implementations: \cite{Attiya2008,Fraser2007,Luchangco2003}
+
 \begin{lstlisting}
@@
 
-parallel CML's implementation of choice
+pure cml~\cite{Reppy1991}
+
+parallel CML's implementation of choice~\cite{Reppy2009}
 
@@
 
-Haskell's STM
+Haskell's STM~\cite{Harris2005a}
 
@@
 
-Communicating transactions and company, transactional events
+Communicating transactions~\cite{Lesani2011}, 
+Transaction synchronizers~\cite{Luchangco2005},
+Transaction communicators~\cite{Luchangco2011},
+transactional events~\cite{Donnelly2006}
 
@@
 
-elastic transactions and company, Paper on composing lock-free data
-structure operations
+elastic transactions~\cite{Felber2009} and company,
+
+Paper on composing lock-free data structure operations~\cite{Cederman2010}
 

### 2011-11-01 paper/paper.tex
started related work section

@@
 
+\newpage
 \section{Related work}
@@
 
-pure cml~\cite{Reppy1991}
+Concurrent ML~\cite{Reppy1991} was designed to resolve an apparent tension
+between abstraction and choice: if protocols are represented abstractly as
+functions, it is impossible to express the choice of two abstract protocols.
+The solution is \emph{higher-order concurrency}, in which synchronous
+message-passing protocols are represented abstractly as \emph{events}.  CML's
+events are built up from combinators, including a choice combinator,
+communication combinators, and combinators for arbitrary computations not
+involving communication.  Events are an abstract data type: externally to the
+CML library, they appear as abstract, composable values.  Internally to the
+library, however, events can be decomposed into their constituent combinators,
+which allows CML to execute them.
+
+Although CML events do not form a monad~\cite{?}, the basic pattern of
+representing computations as abstract, composable values has become extremely
+common in monadic programming
 
@@
 Communicating transactions~\cite{Lesani2011},
+
 Transaction synchronizers~\cite{Luchangco2005},
 Transaction communicators~\cite{Luchangco2011},
+
 transactional events~\cite{Donnelly2006}
 
+join patterns~\cite{?}
+
 \subsection{Composing lock-free data structures}
@@
 
-Paper on composing lock-free data structure operations~\cite{Cederman2010}
+composing lock-free data structure operations~\cite{Cederman2010}
 
@@
 
+\newpage
 \bibliographystyle{abbrvnat}

### 2011-11-01 paper/paper.tex
related work

@@
 acquiring the lock before performing any operation.  The prevailing
-wisdom\footnote{With some notable exceptions, like the recent work on
-  \emph{flat combining}~\cite{?}, which we explore in \secref{?}.} is
+wisdom\footnote{With some dissenting opinions, as in the recent work on
+  \emph{flat combining}~\cite{Hendler2010}.} is
 that such \emph{coarse-grained} locking is inherently unscalable:
@@
 
+\elide{
 The design of reagents builds on a foundation laid by Concurrent ML
 and extended in Haskell's STM 
+}
 
@@
 
+Reagents are clearly influenced by the design of CML's events, and include
+variants of CML's core event combinators (communication and choice).  Where
+CML is aimed squarely at capturing synchronous communication protocols,
+reagents are designed for writing and tailoring fine-grained concurrent data
+structures and synchronization primitives.  This difference in motivation led
+us to include a number of additional combinators, including those dealing
+directly with shared state.
+
+Originally, CML was focused on managing concurrency rather than profiting from
+parallelism, and this focus was reflected in its implementation.  More
+recently, a parallel implementation of CML was proposed~\cite{Reppy2009}.  The
+key challenge is resolving uses of choice both consistently and scalably.  It
+is addressed by \emph{sharing}: an event making a choice is enrolled as
+offering a communication corresponding to each possible choice, and when a
+communication is accepted, the (single, shared) event is atomically marked as
+consumed.  We follow a similar strategy in dealing with message passing, but
+where Parallel CML uses lock-based queues to store messages, we show how to
+use lock-free bags for increased parallelism~(\secref{implementation}).  We
+also show how to incorporate choice resolution with simultaneous shared-state
+updates and our conjunction combinators.
+
+\elide{
 Although CML events do not form a monad~\cite{?}, the basic pattern of
-representing computations as abstract, composable values has become extremely
-common in monadic programming
-
-parallel CML's implementation of choice~\cite{Reppy2009}
+representing effectful computations as abstract, composable values has become
+common in monadic programming~\cite{?}, including Haskell's STM and
+Transactional Events, both discussed below.
+}
 
@@
 
-STM in general
+Software transactional memory was originally intended ``to provide a general
+highly concurrent method for translating sequential object implementations
+into non-blocking ones''~\cite{Shavit1997}.  This ambitious goal has led to a
+remarkable research literature, which has been summarized in textbook
+form~\cite{Larus2006}.  Much of the research is devoted to achieving
+scalability on multiprocessors or multicores, sometimes by relaxing
+consistency guarantees~\cite{?} or only providing obstruction-freedom rather
+than lock-freemdom~\cite{?}.  %Work in the area is
+
+Reagents, on the other hand, are aimed at a less ambitious goal: enabling the
+concise expression, user tailoring, and composition of fine-grained concurrent
+algorithms.  That is, unlike STM, reagents do not attempt to provide a
+\emph{universal} fine-grained concurrent algorithm.  Instead, they assist in
+writing and using specific algorithms.
+
+There is a clear tradeoff.  Using STM, one can implement a concurrent stack or
+queue by simply wrapping a sequential version in an \lstinline{atomic} block,
+which requires no algorithmic insight and is simpler than the stack or queue
+we give in~\secref{reagents}.  But even with a very clever STM, these
+implementations are unlikely to scale as well as our elimination stack or
+Michael-Scott queue; some evidence for that is shown in~\secref{performance}.
+
+Reagents carve out a middle ground between completely hand-written algorithms
+and the completely automatic \lstinline{atomic} blocks of STM.  When used in
+isolation, reagents are \emph{guaranteed} to perform only the CASes that the
+hand-written algorithm would, so they introduce no overhead on shared-memory
+operations\elide{; we know of no STM implementation that can provide similar
+  guarantees or a similar cost model}.  Yet, unlike hand-written algorithms,
+reagents can be tailored and composed using choice, adding blocking behavior,
+or combining sequentially into larger atomic blocks. %TODO
 
@@
 
-Communicating transactions~\cite{Lesani2011},
-
-Transaction synchronizers~\cite{Luchangco2005},
-Transaction communicators~\cite{Luchangco2011},
-
-transactional events~\cite{Donnelly2006}
-
-join patterns~\cite{?}
+A central tenet of transactions is \emph{isolation}: transactions should not
+be aware of concurrent changes to memory from other transactions.  But
+sometimes it is desirable for concurrent transactions to coordinate or
+otherwise communicate while executing.  Two very recent papers have proposed
+mechanisms for incorporating communication with STM, in the form of
+Communicating Transactions~\cite{Lesani2011} and Transaction
+Communicators~\cite{Luchangco2011b} (the latter extends earlier work on
+Transaction Synchronizers~\cite{Luchangco2005}).  In both cases, a key
+question is how the expected isolation of shared memory can safely coexist
+with concurrent communication.  Communicating Transactions use explicit,
+asynchronous message passing to communicate; the mechanism is entirely
+separate from shared memory, which retains isolation.  On the other hand,
+Transaction Communicators allow isolation to be weakened in a controlled way.
+
+Our mixture of message-passing and shared-state combinators most closely
+resembles the work on Communicating Transactions.   %TODO
+
+\elide{
+There has also been work treating pure message-passing in a transactional way,
+both explicitly in Transactional Events~\cite{Donnelly2006} and implicitly in
+the Join Calculus~\cite{?}.  The former combines CML with an atomic sequencing
+operator, while the latter introduces \emph{join patterns} that can receive
+from multiple channels simultaneously and atomically.  Both are expressible
+using reagents, through the combination of \lstinline{swap} and the
+conjunction combinators.  Thus, reagents provide a new implementation for both
+of these systems.  Previously, Transactional Events was implemented on top of
+Haskell's STM, while join patterns have largely been implemented via
+coarse-grained locking~\cite{?}.  
+}
+
+There has also been work treating pure message-passing in a transactional way:
+Transactional Events~\cite{Donnelly2006} combines CML with an atomic
+sequencing operator.  Previously, Transactional Events were implemented on top
+of Haskell's STM, relied on lightweight search threads for matching
+communications, and used an STM-based representation of channels.  However,
+Transactional Events are expressible using reagents, through the combination
+of \lstinline{swap} and the conjunction combinators.  Doing so yields a new
+implementation that does not require search threads, performs parallel
+matching for communication, and represents channels as lock-free bags.  We are
+not in a position to do a head-to-head comparison, but based on the results
+in~\secref{performance}, we expect the reagent-based implementation to scale
+better on fine-grained workloads.
 
@@
 
+Most of the literature on fine-grained concurrent data structures studies only
+``within-object'' atomicity, for example developing algorithms for inserting
+or removing elements into a collection atomically.  However, there has been
+some nice work studying the problem of transferring data atomically between
+such fine-grained data structures~\cite{Cederman2010}.  The basic approach
+relies on a \emph{k}CAS operation in much the same way that reagent sequencing
+does.  The transfer methods must be defined manually, in advance, and with
+access to the internals of the relevant data structures, whereas reagents
+allow arbitrary new compositions, without manual definition, and without
+access to the code or internals of the involved data structures.
+
+Atomic transfers between concurrent data structures
+
 elastic transactions~\cite{Felber2009} and company,
@@
 
-\newpage
 \bibliographystyle{abbrvnat}
@@
 1.0 Related work                        1.0
-0.3 Conclusion                          0.2
-0.8 Bib                                 0.8
+0.3 Conclusion                          0.3
+0.8 Bib                                 0.7
 

### 2011-11-02 paper/paper.tex
related work, and new locking benchmarks

@@
 \authorinfo{}{}{}
-% yikes!
-%\title{Reagents: expressing and composing patterns of fine-grained concurrency}
+%\authorinfo{Aaron Turon}{Northeastern University}{turon@ccs.neu.edu}
+
 \title{Reagents: expressing and composing fine-grained concurrency}
@@
 new platforms.  They tend to be conservative, implementing only those data
-structures and primitives likely to fulfill common needs.  In addition, it is
-generally not possible to safely combine the facilities of the library.  For
-example, JUC provides queues, sets and maps, but not stacks or bags.  Its
-queues come in both blocking and nonblocking forms, while its sets and maps
-are nonblocking only.  Although the queues provide atomic (thread-safe)
-dequeuing and sets provide atomic insertion, it is not possible to combine
-these into a single atomic operation that moves an element from a queue into a
-set.
+structures and primitives likely to fulfill common needs, and it is generally
+not possible to safely combine the facilities of the library.  For example,
+JUC provides queues, sets and maps, but not stacks or bags.  Its queues come
+in both blocking and nonblocking forms, while its sets and maps are
+nonblocking only.  Although the queues provide atomic (thread-safe) dequeuing
+and sets provide atomic insertion, it is not possible to combine these into a
+single atomic operation that moves an element from a queue into a set.
 
@@
 
+\cite{Shavit1997a} %TODO: cite pure elimination
+
 Broadly, we are interested in data structures and algorithms for
-communication, synchronization, or both.  This section gives a brief
-survey of the most important techniques for communication and
-synchronization in a fine-grained setting---exactly the techniques
-that reagents abstract and generalize.  Readers already familiar with
-Treiber stacks~\cite{?}, elimination-backoff stacks~\cite{?}, dual
-stacks~\cite{?}, and MCS locks~\cite{?} can safely skip to
-\secref{reagents}.
+communication, synchronization, or both.  This section gives a brief survey of
+the most important techniques for communication and synchronization in a
+fine-grained setting---exactly the techniques that reagents abstract and
+generalize.  Readers already familiar with Treiber stacks~\cite{Treiber1986},
+elimination-backoff stacks~\cite{Hendler2004}, dual stacks~\cite{Scherer2004},
+and MCS locks~\cite{Mellor-Crummey1991} can safely skip to \secref{reagents}.
 
@@
 \subsection{Dynamic reagents and \lstinline{cas}}
+\label{sec:dynamic}
 
@@
 communication combinators, and combinators for arbitrary computations not
-involving communication.  Events are an abstract data type: externally to the
+involving communication.  \elide{
+Events are an abstract data type: externally to the
 CML library, they appear as abstract, composable values.  Internally to the
@@
 which allows CML to execute them.
-
+}
 Reagents are clearly influenced by the design of CML's events, and include
-variants of CML's core event combinators (communication and choice).  Where
-CML is aimed squarely at capturing synchronous communication protocols,
-reagents are designed for writing and tailoring fine-grained concurrent data
-structures and synchronization primitives.  This difference in motivation led
-us to include a number of additional combinators, including those dealing
-directly with shared state.
+variants of CML's core event combinators\elide{(communication and choice)}.
+But where CML is aimed squarely at capturing synchronous communication
+protocols, reagents are designed for writing and tailoring fine-grained
+concurrent data structures and synchronization primitives.  This difference in
+motivation led us to include a number of additional combinators, including
+those dealing directly with shared state.
 
@@
 use lock-free bags for increased parallelism~(\secref{implementation}).  We
-also show how to incorporate choice resolution with simultaneous shared-state
-updates and our conjunction combinators.
+also show how to incorporate choice resolution with shared-state updates and
+our conjunction combinators.
 
@@
 \subsection{Software transactional memory}
+\label{sec:stm}
 
@@
 consistency guarantees~\cite{?} or only providing obstruction-freedom rather
-than lock-freemdom~\cite{?}.  %Work in the area is
+than lock-freemdom~\cite{Herlihy2003}.  %Work in the area is
 
@@
 hand-written algorithm would, so they introduce no overhead on shared-memory
-operations\elide{; we know of no STM implementation that can provide similar
-  guarantees or a similar cost model}.  Yet, unlike hand-written algorithms,
-reagents can be tailored and composed using choice, adding blocking behavior,
-or combining sequentially into larger atomic blocks. %TODO
-
-Haskell's STM~\cite{Harris2005a}
-
-left-biased choice, but differently from \lstinline{orElse} in
-Haskell's STM.
+operations; by recoding an algorithm use reagents, you lose nothing.\elide{;
+  we know of no STM implementation that can provide similar guarantees or a
+  similar cost model} Yet unlike hand-written algorithms, reagents can be
+composed using choice, tailored with new blocking behavior, or combined
+into larger atomic blocks. %TODO
+
+Haskell's STM~\cite{Harris2005a} was inspirational in showing that
+transactions can be represented via monads~\cite{PeytonJones1993}, explicitly
+composed, and combined with blocking and choice operators.  Reagents also form
+a monad, but we have chosen an interface closer to
+\emph{arrows}~\cite{Hughes2000}, to encourage static reagent layout wherever
+possible~(\secref{dynamic}).  Like \lstinline{orElse} in Haskell's STM, our
+choice operator is left-biased.  But unlike \lstinline{orElse}, our choice
+operator will attempt the right-hand side even when the left-hand side has
+only failed transiently (rather than permanently).  While the distinction may
+appear somewhat technical, it is crucial for supporting examples like the
+elimination backoff stack~(\secref{choice}).
 
@@
 Our mixture of message-passing and shared-state combinators most closely
-resembles the work on Communicating Transactions.   %TODO
+resembles the work on Communicating Transactions.  Of course, the most
+important difference is in the way we deal with shared state (and the
+guarantees provided thereby), which we discussed in~\secref{stm}.  Aside from
+that, we believe that \emph{synchronous} communication is a better choice for
+expressing patterns like elimination~(\secref{choice}), since the pattern
+relies on both participants being mutually-aware.  %TODO: say more here
 
@@
 
-\subsection{Composing lock-free data structures}
-
-Most of the literature on fine-grained concurrent data structures studies only
-``within-object'' atomicity, for example developing algorithms for inserting
-or removing elements into a collection atomically.  However, there has been
-some nice work studying the problem of transferring data atomically between
-such fine-grained data structures~\cite{Cederman2010}.  The basic approach
-relies on a \emph{k}CAS operation in much the same way that reagent sequencing
-does.  The transfer methods must be defined manually, in advance, and with
-access to the internals of the relevant data structures, whereas reagents
-allow arbitrary new compositions, without manual definition, and without
-access to the code or internals of the involved data structures.
-
-Atomic transfers between concurrent data structures
-
-elastic transactions~\cite{Felber2009} and company,
-
-composing lock-free data structure operations~\cite{Cederman2010}
+\subsection{Composing fine-grained concurrent data structures}
+
+Most of the literature on fine-grained concurrent data structures is focused
+on ``within-object'' atomicity, for example developing algorithms for
+inserting or removing elements into a collection atomically.  However, there
+has been some recent work studying the problem of transferring data atomically
+between such fine-grained data structures~\cite{Cederman2010}.  The basic
+approach relies on a \emph{k}CAS operation in much the same way that reagent
+sequencing does.  The transfer methods must be defined manually, in advance,
+and with access to the internals of the relevant data structures, whereas
+reagents allow arbitrary new compositions, without manual definition, and
+without access to the code or internals of the involved data structures.
+Nevertheless, the implementation of reagent composition yields an algorithm
+very similar to the manually-written transfer methods.
+
+It is also possible to go in the other direction: start from STM, which
+provides composition, and add an ``escape hatch'' for writing arbitrary
+fine-grained concurrent algorithms within the scope of a transaction.  The
+escape hatch can be provided through unlogged reads and/or writes to memory
+locations being used by transactions, as in early release~\cite{Herlihy2003}
+or elastic transactions~\cite{Felber2009}.  As we discussed
+above~(\secref{stm}), we favor an approach where the focus is foremost on
+writing fine-grained algorithms, with \emph{guarantees} about the performance
+and shared-memory semantics of those algorithms.  Providing such guarantees
+via an escape hatch mechanism may be difficult or impossible, depending on the
+details of the STM implementation.  As we showed in~\secref{reagents}, it is
+also very useful to have combinators for choice, message-passing, and
+blocking, if one wishes to capture the full range of fine-grained concurrent
+algorithms.
 

### 2011-11-02 src/main/scala/bench/Bench.scala

@@ -251,7 +251,7 @@ object Bench extends App {
 //    List(100, 250, 500)
 //    List(0) ++ (for (i <- 0 to 15) yield pow(10, 1+i.toDouble * 0.25).toInt)
 //    (0 to 600 by 50)
-    List(100, 1000)
+    List(50, 500, 5000)
 //    (50 to 1000 by 50).toList
   else config.workList = config.workList.reverse
 

### 2011-11-02 src/main/scala/bench/benchmarks/EnqDeq.scala

@@ -31,6 +31,20 @@ object EnqDeq extends Benchmark {
       }
     }
   } 
+  private object lock extends Entry {
+    def name = "lock"
+    type S = LockQueue[AnyRef]
+    def setup = new LockQueue()
+    def run(q: S, work: Int, iters: Int) {
+      val r = new Random
+      for (_ <- 1 to iters) {
+	q.enq(SomeData)
+	Util.noop(r.fuzz(work))
+	untilSome {q.tryDeq}
+	Util.noop(r.fuzz(work))
+      }
+    }
+  } 
   private object stm extends Entry {
     def name = "stm"
     type S = STMQueue[AnyRef]
@@ -101,6 +115,7 @@ object EnqDeq extends Benchmark {
       }
     }
   }
+/*
   private object lock extends Entry {
     def name = "lock"
     type S = (java.util.LinkedList[AnyRef], ReentrantLock)
@@ -124,6 +139,7 @@ object EnqDeq extends Benchmark {
       }
     }
   }
+*/
   private object fc extends Entry {
     def name = "fc"
     type S = FCQueue
@@ -173,7 +189,7 @@ object EnqDeq extends Benchmark {
   }
 */
 
-  def entries: List[Entry] = List(stm) //, simple, reagent, hand, juc)
+  def entries: List[Entry] = List(reagent, lock, hand, stm, simple, juc)
   // fc livelocks
   // basket is not licensed for distribution
 }

### 2011-11-02 src/main/scala/bench/benchmarks/PushPop.scala

@@ -112,6 +112,13 @@ object PushPop extends Benchmark {
     def tryPop(s: S): Option[AnyRef]  = s.tryPop
   }
 */
+  private object lock extends Entry with Generic{
+    def name = "lock"
+    type S = LockStack[AnyRef]
+    def setup = new LockStack()
+    def push(s: S, x: AnyRef) = s.push(x)
+    def tryPop(s: S): Option[AnyRef]  = s.tryPop
+  } 
   private object hand extends Entry with Generic{
     def name = "hand"
     type S = HandStack[AnyRef]
@@ -191,5 +198,5 @@ private object rTreiberFast extends Entry with Generic {
 //  def entries = List(rTreiber, rElim, hand)
 //  def entries = List(rElim, rTreiber, handElim, hand, handPool)
 //  def entries: List[Entry] = List(rElim, handPool, handElim)
-  def entries: List[Entry] = List(stm) //List(rTreiber, stm, hand)
+  def entries: List[Entry] = List(rTreiber, lock, hand, stm, rElim)
 }

### 2011-11-02 src/main/scala/bench/competition/LockQueue.scala

@@ -1,0 +1,24 @@
+package chemistry.bench.competition
+
+import java.util.concurrent.locks._
+import scala.annotation.tailrec
+import scala.collection.mutable._
+
+// Queue w/ coarse-grained locking
+class LockQueue[A >: Null] {
+  private val queue = new Queue[A]()
+  private val lock  = new ReentrantLock()
+
+  def enq(x: A) {
+    lock.lock()
+    queue.enqueue(x)
+    lock.unlock()
+  }
+
+  def tryDeq(): Option[A] = {
+    lock.lock()
+    val ret = if (queue.isEmpty) None else Some(queue.dequeue())      
+    lock.unlock()
+    ret
+  } 
+}

### 2011-11-02 src/main/scala/bench/competition/LockStack.scala

@@ -1,0 +1,24 @@
+package chemistry.bench.competition
+
+import java.util.concurrent.locks._
+import scala.annotation.tailrec
+import scala.collection.mutable._
+
+// Stack w/ coarse-grained locking
+class LockStack[A >: Null] {
+  private val stack = new Stack[A]()
+  private val lock  = new ReentrantLock()
+
+  def push(x: A) {
+    lock.lock()
+    stack.push(x)
+    lock.unlock()
+  }
+
+  def tryPop(): Option[A] = {
+    lock.lock()
+    val ret = if (stack.isEmpty) None else Some(stack.pop())      
+    lock.unlock()
+    ret
+  } 
+}

### 2011-11-04 src/main/scala/KCASRef.scala
transfer benchmarks, kcas and offer fixes

@@ -110,6 +110,8 @@ private final class KCAS(val casList: List[CAS[_]]) {
 
 private object KCAS {
   def tryCommit(casList: List[CAS[_]]): Boolean = {
+//    println("KCAS")
+
     // attempt to place KCAS record in each CASed reference.  returns null
     // if successful, and otherwise the point in the CAS list to rollback
     // from.
@@ -129,7 +131,8 @@ private object KCAS {
       if (start != end) start match {
 	case Nil => throw Util.Impossible // somehow went past the `end`
 	case CAS(ref, ov, nv) :: rest => {
-	  ref.data.compareAndSet(null, ov) // roll back to old value
+	  ref.data.lazySet(ov) // roll back to old value
+	  //ref.data.compareAndSet(null, ov) 
 	  rollBackBetween(rest, end)
 	}
       }

### 2011-11-04 src/main/scala/MSQueue.scala

@@ -64,10 +64,11 @@ final class MSQueue[A >: Null] {
     }
     search
   }
-  // val tryDeq: Reagent[Unit, Option[A]] = head.upd[Option[A]] {
-  //   case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
-  //   case emp => (emp, None)
-  // }
+
+  private val tryDeqForComp: Reagent[Unit, Option[A]] = head.upd[Option[A]] {
+    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
+    case emp => (emp, None)
+  }
   object tryDeq extends Reagent[Unit,Option[A]] {
     @inline def tryReact(u:Unit, rx: Reaction, offer: Offer[Option[A]]): Any = 
       head.data.get match {
@@ -76,14 +77,27 @@ final class MSQueue[A >: Null] {
 	  if (head.data.compareAndSet(old,n)) Some(x) else Retry
 	case _ => None
       }
-    def composeI[B](next: Reagent[Option[A],B]) = throw Util.Impossible
+    def composeI[B](next: Reagent[Option[A],B]) = tryDeqForComp >=> next
     def maySync = false
     def alwaysCommits = false
     def snoop(a: Unit) = false
   }
 
-  val deq: Reagent[Unit, A] = head.upd[A] {
+  private val deqForComp: Reagent[Unit, A] = head.upd[A] {
     case Node(_, Ref(n@Node(x, _))) => (n, x)
+  }
+  object deq extends Reagent[Unit,A] {
+    @inline def tryReact(u:Unit, rx: Reaction, offer: Offer[A]): Any = 
+      head.data.get match {
+	case null => Retry
+	case old@Node(_, Ref(n@Node(x, _))) =>
+	  if (head.data.compareAndSet(old,n)) x else Retry
+	case _ => Retry //nonblocking version
+      }
+    def composeI[B](next: Reagent[A,B]) = deqForComp >=> next
+    def maySync = false
+    def alwaysCommits = false
+    def snoop(a: Unit) = false
   }
 }
 

### 2011-11-04 src/main/scala/Offer.scala

@@ -5,6 +5,7 @@
 
 package chemistry
 
+import scala.annotation.tailrec
 import java.util.concurrent.locks._
 import java.util.concurrent.atomic._
 
@@ -65,23 +66,24 @@ private final class Waiter[-A](val blocking: Boolean)
   }
   
   @inline def isActive: Boolean = status.data.get == Waiting
-
-  // Poll current waiter value:
-  //   - None if Waiting or Aborted
-  //   - Some(ans) if completed with ans
-  // sadly, have to use `Any` to work around variance problems
-  @inline def poll: Option[Any] = status.data.get match {
-    case null => None
-    case (_:WaiterStatus) => None
+  
+  // Attempt to abort, returning 
+  //   - None if abort succeeded
+  //   - Some(ans) if waiter already completed with ans
+  @tailrec def tryAbort: Option[Any] = status.data.get match {
+    case null => tryAbort
+    case Aborted => None 
+    case Waiting => 
+      if (status.data.compareAndSet(Waiting, Aborted)) None
+      else tryAbort
     case ans => Some(ans)
   }
-  
-  // Attempt to abort, returning true iff successful
-  @inline def tryAbort = status.data.compareAndSet(Waiting, Aborted)
+
+
   @inline def rxWithAbort(rx: Reaction): Reaction =
     rx.withCAS(status, Waiting, Aborted)
 
-  def abortAndWake = if (tryAbort) wake()
+  def abortAndWake = if (tryAbort eq None) wake()
 
   @inline def tryComplete(a: A) = 
     status.data.compareAndSet(Waiting, a.asInstanceOf[AnyRef])

### 2011-11-04 src/main/scala/Reagent.scala

@@ -48,8 +48,8 @@ abstract class Reagent[-A, +B] {
 	tryReact(a, Reaction.inert, waiter) match {
 	  case (bc: BacktrackCommand) if wait => {
 	    bc.bottom(waiter, backoff, snoop(a))
-	    waiter.tryAbort	// rescind waiter,
-	    waiter.poll match { // but check if already completed
+	    waiter.tryAbort match { // rescind waiter, but check if already
+				    // completed
 	      case Some(ans) => ans.asInstanceOf[B] 
 	      case None      => retryLoop(bc.isBlock)
 	    }
@@ -144,8 +144,13 @@ private final case class Commit[A]() extends Reagent[A,A] {
   def tryReact(a: A, rx: Reaction, offer: Offer[A]): Any = {
     offer match {
       case null => if (rx.tryCommit) a else Retry
-      case (w: Waiter[_]) =>
-	if (w.rxWithAbort(rx).tryCommit) a else Retry
+//      case (w: Waiter[_]) => if (w.rxWithAbort(rx).tryCommit) a else Retry
+      case (w: Waiter[_]) => {
+	w.tryAbort match { // rescind waiter, but check if already completed
+	  case Some(ans) => ans
+	  case None      => if (rx.tryCommit) a else Retry
+	}
+      }
       case (_: Catalyst[_]) => {
 	rx.tryCommit
 	Block

### 2011-11-04 src/main/scala/TreiberStack.scala

@@ -97,6 +97,19 @@ final class TreiberStack[A >: Null] {
 
   val head = new Ref[List[A]](Nil)
 
+  private val pushForComp: Reagent[A,Unit] = head.upd[A,Unit] { 
+    case (xs,x) => (x::xs, ())
+  }
+
+  private val tryPopForComp: Reagent[Unit,Option[A]] = head.upd[Option[A]] {
+    case x::xs => (xs,  Some(x))
+    case Nil   => (Nil, None)
+  }
+
+  private val popForComp: Reagent[Unit,A] = head.upd[A] {
+    case x::xs => (xs, x)
+  }
+
   object push extends Reagent[A,Unit] {
     def tryReact(x:A, rx: Reaction, offer: Offer[Unit]): Any = {
       val cur = head.data.get
@@ -110,7 +123,7 @@ final class TreiberStack[A >: Null] {
 	  Retry
       }
     }
-    def composeI[B](next: Reagent[Unit,B]) = throw Util.Impossible
+    def composeI[B](next: Reagent[Unit,B]) = pushForComp >=> next
     def maySync = false
     def alwaysCommits = false
     def snoop(a: A) = false
@@ -128,13 +141,29 @@ final class TreiberStack[A >: Null] {
       else 
 	Retry 
     }
-    def composeI[B](next: Reagent[Option[A],B]) = throw Util.Impossible
+    def composeI[B](next: Reagent[Option[A],B]) = tryPopForComp >=> next
     def maySync = false
     def alwaysCommits = false
     def snoop(a: Unit) = false
   }
 
-  def pop: Reagent[Unit,A] = throw Util.Impossible
+  object pop extends Reagent[Unit,A] {
+    @inline def tryReact(u:Unit, rx: Reaction, offer: Offer[A]): Any = {
+      val cur = head.data.get
+      if (cur eq null) 
+	Retry
+      else if (cur.isEmpty)
+	Retry //nonblocking version
+      else if (head.data.compareAndSet(cur,cur.tail)) 
+	cur.head
+      else 
+	Retry 
+    }
+    def composeI[B](next: Reagent[A,B]) = popForComp >=> next
+    def maySync = false
+    def alwaysCommits = false
+    def snoop(a: Unit) = false
+  }
 
 /*
   object pop extends Reagent[Unit,A] {

### 2011-11-04 src/main/scala/bench/Bench.scala

@@ -251,7 +251,7 @@ object Bench extends App {
 //    List(100, 250, 500)
 //    List(0) ++ (for (i <- 0 to 15) yield pow(10, 1+i.toDouble * 0.25).toInt)
 //    (0 to 600 by 50)
-    List(50, 500, 5000)
+    List(30, 300, 3000)
 //    (50 to 1000 by 50).toList
   else config.workList = config.workList.reverse
 
@@ -264,7 +264,7 @@ object Bench extends App {
   } yield (b, 0, 1, 1)
   private val concBenches = for {
 //    b <- List(PushPop, EnqDeq, IncDec)
-    b <- List(PushPop, EnqDeq)
+    b <- List(Transfer, PushPop, EnqDeq)
     w <- config.workList
   } yield (b, w, config.minCores, config.maxCores)
 

### 2011-11-04 src/main/scala/bench/benchmarks/PushPop.scala

@@ -198,5 +198,6 @@ private object rTreiberFast extends Entry with Generic {
 //  def entries = List(rTreiber, rElim, hand)
 //  def entries = List(rElim, rTreiber, handElim, hand, handPool)
 //  def entries: List[Entry] = List(rElim, handPool, handElim)
-  def entries: List[Entry] = List(rTreiber, lock, hand, stm, rElim)
+//  def entries: List[Entry] = List(rTreiber, lock, hand, stm, rElim)
+  def entries: List[Entry] = List(rElim, rTreiber)
 }

### 2011-11-04 src/main/scala/bench/benchmarks/StackTransfer.scala

@@ -1,0 +1,126 @@
+package chemistry.bench.benchmarks
+
+import chemistry._
+import chemistry.bench._
+import chemistry.bench.competition._
+import chemistry.Util._
+
+object Transfer extends Benchmark {
+  def pureWork(work: Int, iters: Int) = {
+    val r = new Random
+    for (_ <-1 to iters) {
+      Util.noop(r.fuzz(work * 3))
+    }
+  }
+
+  private object reagent extends Entry {
+    def name = "reagent"
+    type S = (TreiberStack[AnyRef], TreiberStack[AnyRef])
+    def setup = (new TreiberStack, new TreiberStack)
+    def run(p: S, work: Int, iters: Int) {
+      val r = new Random
+      val (s1, s2) = p
+      val moveRight = s1.pop >=> s2.push
+      val moveLeft  = s2.pop >=> s1.push
+
+      for (_ <- 1 to iters) {
+	val (sFrom, sTo, move) = 
+	  if (r.next(2) == 0)
+	    (s1, s2, moveRight)
+	  else
+	    (s2, s1, moveLeft)
+
+	sFrom.push ! SomeData
+	Util.noop(r.fuzz(work))
+	move ! ()
+	Util.noop(r.fuzz(work))
+	untilSome {sTo.tryPop ! ()}
+	Util.noop(r.fuzz(work))
+      }
+    }
+  }
+
+  private object rElim extends Entry {
+    def name = "rElim"
+    type S = (EliminationStack[AnyRef], EliminationStack[AnyRef])
+    def setup = (new EliminationStack, new EliminationStack)
+    def run(p: S, work: Int, iters: Int) {
+      val r = new Random
+      val (s1, s2) = p
+      val moveRight = s1.pop >=> s2.push
+      val moveLeft  = s2.pop >=> s1.push
+
+      for (_ <- 1 to iters) {
+	val (sFrom, sTo, move) = 
+	  if (r.next(2) == 0)
+	    (s1, s2, moveRight)
+	  else
+	    (s2, s1, moveLeft)
+
+	sFrom.push ! SomeData
+	Util.noop(r.fuzz(work))
+	move ! ()
+	Util.noop(r.fuzz(work))
+	untilSome {sTo.tryPop ! ()}
+	Util.noop(r.fuzz(work))
+      }
+    }
+  }
+
+  private object lock extends Entry {
[…]

### 2011-11-04 src/main/scala/bench/benchmarks/Transfer.scala

@@ -1,22 +1,0 @@
-package chemistry.bench.benchmarks
-
-import chemistry._
-import chemistry.bench._
-import chemistry.bench.competition._
-import chemistry.Util._
-/*
-object Transfer extends Benchmark {
-  def pureWork(work: Int, iters: Int) = {
-    val r = new Random
-    for (_ <-1 to iters) {
-      Util.noop(r.fuzz(work * 20))
-    }
-  }
-
-  private object reagent extends Entry {
-    def name = "reagent"
-    type S = 
-  }
-}
-
-*/

### 2011-11-04 src/main/scala/bench/competition/LockQueue.scala

@@ -7,7 +7,7 @@ import scala.collection.mutable._
 // Queue w/ coarse-grained locking
 class LockQueue[A >: Null] {
   private val queue = new Queue[A]()
-  private val lock  = new ReentrantLock()
+  val lock  = new ReentrantLock()
 
   def enq(x: A) {
     lock.lock()

### 2011-11-04 src/main/scala/bench/competition/LockStack.scala

@@ -7,7 +7,7 @@ import scala.collection.mutable._
 // Stack w/ coarse-grained locking
 class LockStack[A >: Null] {
   private val stack = new Stack[A]()
-  private val lock  = new ReentrantLock()
+  val lock  = new ReentrantLock()
 
   def push(x: A) {
     lock.lock()
@@ -18,6 +18,13 @@ class LockStack[A >: Null] {
   def tryPop(): Option[A] = {
     lock.lock()
     val ret = if (stack.isEmpty) None else Some(stack.pop())      
+    lock.unlock()
+    ret
+  } 
+
+  def pop(): A = {
+    lock.lock()
+    val ret = stack.pop()
     lock.unlock()
     ret
   } 

### 2011-11-04 src/main/scala/bench/competition/STMStack.scala

@@ -17,4 +17,9 @@ class STMStack[A] {
       Some(cur.head)
     }
   }
+  def pop(): A = atomic {
+    val cur = state.get
+    state.set(cur.tail)
+    cur.head
+  }
 }

### 2011-11-04 src/main/scala/Channels.scala
more benchmarking work

@@ -10,14 +10,13 @@ final private class Message[A,B,C](
   payload: A,		 // the actual content of the message
   senderRx: Reaction,	 // the checkpointed reaction of the sender
   senderK: Reagent[B,C], // the sender's reagent continuation
-  offer: Offer[C]	 // the sender's offer
+  val offer: Offer[C]	 // the sender's offer
 ) extends DeletionFlag {
   private case class CompleteExchange[D](receiverK: Reagent[A,D]) 
 	       extends Reagent[C,D] {
-    def tryReact(c: C, rx: Reaction, enclosingOffer: Offer[D]): Any =
+    def tryReact(c: C, rx: Reaction, enclosingOffer: Offer[D]): Any =      
       offer.consumeAndContinue(
-	c, payload, rx ++ senderRx, receiverK, enclosingOffer
-      )
+	c, payload, rx ++ senderRx, receiverK, enclosingOffer)
     def composeI[E](next: Reagent[D,E]): Reagent[C,E] =
       CompleteExchange(receiverK >=> next)
     def maySync = receiverK.maySync
@@ -39,7 +38,9 @@ private final case class Endpoint[A,B,C](
     @tailrec def tryFrom(n: incoming.Node, retry: Boolean): Any = 
       if (n == null) {
 	if (retry) Retry else Block
-      } else n.data.exchange.compose(k).tryReact(a, rx, offer) match {
+      } else if ((n.data.offer eq offer) || rx.hasOffer(n.data.offer)) {
+	tryFrom(n.next, retry)
+      } else n.data.exchange.compose(k).tryReact(a, rx.withOffer(n.data.offer), offer) match {
 	case Retry => tryFrom(n.next, true)
 	case Block => tryFrom(n.next, retry)
 	case ans   => ans

### 2011-11-04 src/main/scala/Reaction.scala

@@ -3,10 +3,12 @@
 
 package chemistry
 
+import scala.collection.immutable.HashSet
+
 private sealed class Reaction private (
   val casList: List[CAS[_]],		// k-cas built up so far
-  val pcList: List[Unit => Unit]	// post-commit actions
-//  val msgSet: HashSet[Message[_,_,_]]   // messages intended for consumption
+  val pcList: List[Unit => Unit],	// post-commit actions
+  val offerSet: HashSet[Offer[_]]       // offers intended for consumption
 ) {
   import Reaction._
   
@@ -22,12 +24,16 @@ private sealed class Reaction private (
     })
 
   def withPostCommit(postCommit: Unit => Unit): Reaction =
-    new Reaction(casList, postCommit +: pcList)
+    new Reaction(casList, postCommit +: pcList, offerSet)
   def withCAS[A <: AnyRef](ref: Ref[A], ov: A, nv: A): Reaction =
-    new Reaction(CAS(ref, ov, nv) +: casList, pcList)
+    new Reaction(CAS(ref, ov, nv) +: casList, pcList, offerSet)
+  def withOffer(offer: Offer[_]): Reaction = 
+    new Reaction(casList, pcList, offerSet + offer)
 
   def ++(rx: Reaction): Reaction = 
-    new Reaction(casList ++ rx.casList, pcList ++ rx.pcList)
+    new Reaction(casList ++ rx.casList, pcList ++ rx.pcList, offerSet union rx.offerSet)
+
+  def hasOffer(offer: Offer[_]): Boolean = offerSet.contains(offer)
 
   def tryCommit: Boolean = {
     val success: Boolean = casCount match {
@@ -41,5 +47,5 @@ private sealed class Reaction private (
   }
 }
 private object Reaction {
-  val inert = new Reaction(Nil, Nil)
+  val inert = new Reaction(Nil, Nil, HashSet.empty)
 }

### 2011-11-04 src/main/scala/SimpleQueue.scala

@@ -4,6 +4,25 @@ import scala.collection.immutable._
 
 final class SimpleQueue[A] {
   private val head = Ref(Queue.empty[A])
+
+  private val enqForComp: Reagent[A,Unit] = head.upd[A,Unit] { 
+    case (xs,x) => (xs.enqueue(x), ())
+  }
+
+  private val tryDeqForComp: Reagent[Unit,Option[A]] = head.upd[Option[A]] {
+    case xs if (xs.isEmpty) => (xs, None) 
+    case xs => xs.dequeue match { 
+      case (x, rest) => (rest, Some(x)) 
+    }
+  }
+
+  private val deqForComp: Reagent[Unit,A] = head.upd[A](
+    new PartialFunction[Queue[A], (Queue[A], A)] {
+      def apply(xs: Queue[A]) = xs.dequeue match { case (x, rest) => (rest, x) }
+      def isDefinedAt(xs: Queue[A]) = !xs.isEmpty
+    }
+  )
+
   object enq extends Reagent[A,Unit] {
     def tryReact(x:A, rx: Reaction, offer: Offer[Unit]): Any = {
       val cur = head.data.get
@@ -15,7 +34,7 @@ final class SimpleQueue[A] {
 	else Retry
       }
     }
-    def composeI[B](next: Reagent[Unit,B]) = throw Util.Impossible
+    def composeI[B](next: Reagent[Unit,B]) = enqForComp >=> next
     def maySync = false
     def alwaysCommits = false
     def snoop(a: A) = false
@@ -33,7 +52,25 @@ final class SimpleQueue[A] {
 	if (head.data.compareAndSet(cur,rest)) Some(x) else Retry
       }
     }
-    def composeI[B](next: Reagent[Option[A],B]) = throw Util.Impossible
+    def composeI[B](next: Reagent[Option[A],B]) = tryDeqForComp >=> next
+    def maySync = false
+    def alwaysCommits = false
+    def snoop(a: Unit) = false
+  }
+
+  object deq extends Reagent[Unit,A] {
+    @inline def tryReact(u:Unit, rx: Reaction, offer: Offer[A]): Any = {
+      val cur = head.data.get
+      if (cur eq null)
+	Retry
+      else if (cur.isEmpty)
+	Retry //nonblocking version
+      else {
+	val (x, rest) = cur.dequeue
+	if (head.data.compareAndSet(cur,rest)) x else Retry
+      }
+    }
[…]

### 2011-11-04 src/main/scala/bench/Bench.scala

@@ -264,7 +264,7 @@ object Bench extends App {
   } yield (b, 0, 1, 1)
   private val concBenches = for {
 //    b <- List(PushPop, EnqDeq, IncDec)
-    b <- List(Transfer, PushPop, EnqDeq)
+    b <- List(StackTransfer, PushPop, QueueTransfer, EnqDeq)
     w <- config.workList
   } yield (b, w, config.minCores, config.maxCores)
 

### 2011-11-04 src/main/scala/bench/Results.scala

@@ -14,7 +14,7 @@ case class Measurement(
   def format = " %4.1f |".format(throughput)
   def formatR = " %4.1f |".format(rawThroughput)
   def formatN(compTo: Double) = 
-    " %8.2f |".format(rawThroughput / compTo)
+    " %4.1f |".format(rawThroughput / compTo)
 }
 
 case class EntryResult(name: String, ms: Seq[Measurement]) {

### 2011-11-04 src/main/scala/bench/benchmarks/PushPop.scala

@@ -198,6 +198,6 @@ private object rTreiberFast extends Entry with Generic {
 //  def entries = List(rTreiber, rElim, hand)
 //  def entries = List(rElim, rTreiber, handElim, hand, handPool)
 //  def entries: List[Entry] = List(rElim, handPool, handElim)
-//  def entries: List[Entry] = List(rTreiber, lock, hand, stm, rElim)
-  def entries: List[Entry] = List(rElim, rTreiber)
+  def entries: List[Entry] = List(rTreiber, lock, hand, stm, rElim)
+//  def entries: List[Entry] = List(rElim, rTreiber)
 }

### 2011-11-04 src/main/scala/bench/benchmarks/StackTransfer.scala

@@ -5,7 +5,7 @@ import chemistry.bench._
 import chemistry.bench.competition._
 import chemistry.Util._
 
-object Transfer extends Benchmark {
+object StackTransfer extends Benchmark {
   def pureWork(work: Int, iters: Int) = {
     val r = new Random
     for (_ <-1 to iters) {
@@ -121,6 +121,7 @@ object Transfer extends Benchmark {
     }
   }
 
-  def entries: List[Entry] = List(rElim, reagent) //List(stm,rElim,lock,reagent)
+//  def entries: List[Entry] = List(rElim, reagent) //List(stm,rElim,lock,reagent)
+  def entries: List[Entry] = List(stm,lock,reagent) //rElim
 }
 

### 2011-11-04 src/main/scala/bench/competition/LockQueue.scala

@@ -21,4 +21,11 @@ class LockQueue[A >: Null] {
     lock.unlock()
     ret
   } 
+
+  def deq(): A = {
+    lock.lock()
+    val ret = queue.dequeue()
+    lock.unlock()
+    ret
+  } 
 }

### 2011-11-04 src/main/scala/bench/competition/STMQueue.scala

@@ -18,4 +18,10 @@ class STMQueue[A] {
       Some(x)
     }
   }
+  def deq() = atomic {
+    val cur = state.get
+    val (x, xs) = cur.dequeue
+    state.set(xs)
+    x
+  }
 }

### 2011-11-04 src/main/scala/bench/benchmarks/QueueTransfer.scala
added queue transfer benchmark

@@ -1,0 +1,126 @@
+package chemistry.bench.benchmarks
+
+import chemistry._
+import chemistry.bench._
+import chemistry.bench.competition._
+import chemistry.Util._
+
+object QueueTransfer extends Benchmark {
+  def pureWork(work: Int, iters: Int) = {
+    val r = new Random
+    for (_ <-1 to iters) {
+      Util.noop(r.fuzz(work * 3))
+    }
+  }
+
+  private object reagent extends Entry {
+    def name = "reagent"
+    type S = (MSQueue[AnyRef], MSQueue[AnyRef])
+    def setup = (new MSQueue, new MSQueue)
+    def run(p: S, work: Int, iters: Int) {
+      val r = new Random
+      val (s1, s2) = p
+      val moveRight = s1.deq >=> s2.enq
+      val moveLeft  = s2.deq >=> s1.enq
+
+      for (_ <- 1 to iters) {
+	val (sFrom, sTo, move) = 
+	  if (r.next(2) == 0)
+	    (s1, s2, moveRight)
+	  else
+	    (s2, s1, moveLeft)
+
+	sFrom.enq ! SomeData
+	Util.noop(r.fuzz(work))
+	move ! ()
+	Util.noop(r.fuzz(work))
+	untilSome {sTo.tryDeq ! ()}
+	Util.noop(r.fuzz(work))
+      }
+    }
+  }
+
+  private object simple extends Entry {
+    def name = "simple"
+    type S = (SimpleQueue[AnyRef], SimpleQueue[AnyRef])
+    def setup = (new SimpleQueue, new SimpleQueue)
+    def run(p: S, work: Int, iters: Int) {
+      val r = new Random
+      val (s1, s2) = p
+      val moveRight = s1.deq >=> s2.enq
+      val moveLeft  = s2.deq >=> s1.enq
+
+      for (_ <- 1 to iters) {
+	val (sFrom, sTo, move) = 
+	  if (r.next(2) == 0)
+	    (s1, s2, moveRight)
+	  else
+	    (s2, s1, moveLeft)
+
+	sFrom.enq ! SomeData
+	Util.noop(r.fuzz(work))
+	move ! ()
+	Util.noop(r.fuzz(work))
+	untilSome {sTo.tryDeq ! ()}
+	Util.noop(r.fuzz(work))
+      }
+    }
+  }
+
[…]

### 2011-11-04 src/main/scala/Channels.scala
benchmark work, incorporating graphs

@@ -40,7 +40,9 @@ private final case class Endpoint[A,B,C](
 	if (retry) Retry else Block
       } else if ((n.data.offer eq offer) || rx.hasOffer(n.data.offer)) {
 	tryFrom(n.next, retry)
-      } else n.data.exchange.compose(k).tryReact(a, rx.withOffer(n.data.offer), offer) match {
+      } else n.data.exchange.compose(k).tryReact(a, 
+						 rx.withOffer(n.data.offer), 
+						 offer) match {
 	case Retry => tryFrom(n.next, true)
 	case Block => tryFrom(n.next, retry)
 	case ans   => ans
@@ -49,7 +51,9 @@ private final case class Endpoint[A,B,C](
     // send message if so requested.  note that we send the message
     // *before* attempting to react with existing messages in the
     // other direction.
-    if (offer != null) outgoing.put(new Message(a, rx, k, offer))
+    // TODO: should combine !maysync with blocking check
+    if (offer != null && !k.maySync)  
+      outgoing.put(new Message(a, rx, k, offer))
 
     // now attempt an immediate reaction
     tryFrom(incoming.cursor, false) match {

### 2011-11-04 src/main/scala/Pool.scala

@@ -134,7 +134,7 @@ final class CircularPool[A <: DeletionFlag] extends Pool[A] {
     }
   }
 
-  private def size = math.max(1,Chemistry.procs)
+  private def size = 8//math.max(1,Chemistry.procs)
   private val cursors = new Array[LinkNode](size)
   for (i <- 0 to size-1) cursors(i) = new LinkNode
   for (i <- 0 to size-1) cursors(i).nextRef.set(cursors((i+1) % size))

### 2011-11-04 src/main/scala/Reagent.scala

@@ -19,7 +19,7 @@ private[chemistry] case object Block extends BacktrackCommand {
 }
 private[chemistry] case object Retry extends BacktrackCommand {
   def bottom[A](waiter: Waiter[A], backoff: Backoff, snoop: => Boolean): Unit =
-    backoff.once(waiter.isActive && !snoop, 2)
+    backoff.once(waiter.isActive && !snoop, 1)
   def isBlock: Boolean = false
 }
 
@@ -59,6 +59,7 @@ abstract class Reagent[-A, +B] {
 	  case ans => ans.asInstanceOf[B]
 	}
       }
+      backoff.once
       retryLoop(false)
     }
     case ans => ans.asInstanceOf[B]
@@ -224,7 +225,14 @@ object choice {
 	case ans => ans
       }
     def composeI[C](next: Reagent[B,C]) = 
-      Choice(r1.compose(next), r2.compose(next))
+      next match {
+	case Choice(next1, next2) =>
+	  Choice(r1 >=> next1,
+		 Choice(r1 >=> next2,
+			Choice(r2 >=> next1,
+			       r2 >=> next2)))
+	case _ => Choice(r1.compose(next), r2.compose(next))
+      }
     def alwaysCommits = r1.alwaysCommits && r2.alwaysCommits
     def maySync = r1.maySync || r2.maySync
     def snoop(a: A) = r2.snoop(a) || r1.snoop(a) 

### 2011-11-04 src/main/scala/TreiberStack.scala

@@ -117,7 +117,7 @@ final class TreiberStack[A >: Null] {
 	Retry
       else {
 	val upd = x +: cur
-	if (head.data.compareAndSet(cur,upd)) 
+	if (head.data.compareAndSet(cur,upd))
 	  () 
 	else 
 	  Retry

### 2011-11-04 src/main/scala/bench/Bench.scala

@@ -29,7 +29,7 @@ private object config {
   var doTP: Boolean = false
   var minCores = 1
   var maxCores = min(Runtime.getRuntime.availableProcessors, 8)
-  var warmupMillis = 2000
+  var warmupMillis = 1000
   var benchMillis = 1000
   var verbose = true
 }
@@ -118,7 +118,7 @@ abstract class Entry {
     // "communication" will be very fast relative to spin-work
     val trialIters: Int = (totalTP * (benchMillis + (100 * i)) * 
 //      (if (i == 1) 5 else 1) *     			   
-      (if (work > 0) 0.1 * scala.math.log(work) * scala.math.log(work)	 
+      (if (work > 0) 0.1 * scala.math.log(work) //* scala.math.log(work)	 
        else 1)
     ).toInt
     
@@ -264,7 +264,7 @@ object Bench extends App {
   } yield (b, 0, 1, 1)
   private val concBenches = for {
 //    b <- List(PushPop, EnqDeq, IncDec)
-    b <- List(StackTransfer, PushPop, QueueTransfer, EnqDeq)
+    b <- List(PushPop, StackTransfer, EnqDeq, QueueTransfer)
     w <- config.workList
   } yield (b, w, config.minCores, config.maxCores)
 

### 2011-11-04 src/main/scala/bench/benchmarks/PushPop.scala

@@ -198,6 +198,6 @@ private object rTreiberFast extends Entry with Generic {
 //  def entries = List(rTreiber, rElim, hand)
 //  def entries = List(rElim, rTreiber, handElim, hand, handPool)
 //  def entries: List[Entry] = List(rElim, handPool, handElim)
-  def entries: List[Entry] = List(rTreiber, lock, hand, stm, rElim)
+  def entries: List[Entry] = List(rTreiber, rElim, hand, lock, stm)
 //  def entries: List[Entry] = List(rElim, rTreiber)
 }

### 2011-11-04 src/main/scala/bench/benchmarks/StackTransfer.scala

@@ -121,7 +121,7 @@ object StackTransfer extends Benchmark {
     }
   }
 
-//  def entries: List[Entry] = List(rElim, reagent) //List(stm,rElim,lock,reagent)
-  def entries: List[Entry] = List(stm,lock,reagent) //rElim
+//  def entries: List[Entry] = List(rElim, reagent) 
+  def entries: List[Entry] = List(rElim,reagent,stm,lock)
 }
 

### 2011-11-06 paper/paper.tex
benchmark work, incorporating graphs
writing

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass[preprint,nocopyrightspace]{sigplanconf}
 
@@
 Another answer is: communicate efficiently.  The past two
-decades\footnote{Herlihy/Wing, Mellor-Crummey/Scott} have produced a sizable
-collection of algorithms for synchronization, communication, and shared
-storage which minimize the use of memory bandwidth and avoid unnecessary
-waiting.  This research effort has led to industrial-strength
-libraries---\texttt{java.util.concurrent} (JUC) the most prominent---offering
-a wide range of concurrency primitives appropriate for fine-grained
-parallelism.
+decades\footnote{Herlihy/Wing~\cite{?}, Mellor-Crummey/Scott~\cite{?}} have
+produced a sizable collection of algorithms for synchronization, communication,
+and shared storage which minimize the use of memory bandwidth and avoid
+unnecessary waiting.  This research effort has led to industrial-strength
+libraries---\texttt{java.util.concurrent} (JUC) the most prominent---offering a
+wide range of concurrency primitives appropriate for fine-grained parallelism.
 
@@
 
-\paragraph{Expressive}  .
-
-\paragraph{Composable} .
-
-We see this work as serving the needs of two distinct groups: concurrency
-experts and concurrency users.  Using reagents, experts can write algorithms
-more easily, because common patterns are expressible as abstractions and many
-are built-in.  Users can compose, extend, and tailor the library without
-detailed knowledge of the algorithms involved.
-
-blend shared-state and message-passing concurrency
-
-implemented a Scala library
-
-CLEAR COST MODEL
+\paragraph{Expressive}  
+Reagents provide a basic set of building blocks for writing concurrent data
+structures and synchronizers.  The building blocks include isolated atomic
+updates to shared state, and interactive synchronous communication through
+message passing.  This blend of isolation and interaction is essential for
+expressing the full range of fine-grained concurrent
+algorithms~(\secref{choice}).  The building blocks also bake in many common
+concurrency patterns, like optimistic retry loops, backoff schemes, and blocking
+and signalling.  Using reagents, it is possible to express sophisticated
+concurrency algorithms at a higher level of abstraction, while retaining the
+performance and scalability of direct, hand-written implementations.
+
+\paragraph{Composable}
+Reagents also provide a set of composition operators, including choice,
+sequencing and pairing.  Using these operators, clients can extend and tailor
+concurrency primitives without knowledge of the underlying algorithms.  For
+example, if a reagent provides only a nonblocking version of an operation like
+\lstinline{dequeue}, a user can easily tailor it to a version that blocks if the
+queue is empty; this extension will work regardless of how \lstinline{dequeue}
+is defined, and will continue to work even if the \lstinline{dequeue}
+implementation is changed.  Similarly, clients of reagent libraries can sequence
+operations together to provide atomic move operations between arbitrary
+concurrent collections---again, without access to or knowledge of the
+implementations.  Compositionality is also useful to algorithm designers, since
+many sophisticated algorithms can be understood as compositions of simpler
+ones~(\secref{choice}, \secref{conjunction}).
+
+\vskip 6pt
+\noindent
+We begin in~\secref{background} by illustrating, through examples, some of the
+common patterns, tradeoffs and concerns in designing fine-grained concurrent
+algorithms.  The two subsequent sections contain our core technical
+contributions:
+\begin{itemize}
+\item The \textbf{design} of the reagent combinators is given
+  in~\secref{reagents}.  Each combinator is motivated by specific needs and
+  patterns in concurrent algorithms; the section shows in
+  particular how to write all of the algorithms described in~\secref{background}
+  concisely and at a higher-than-usual level of abstraction.
+\item The \textbf{implementation} of reagents is detailed
+  in~\secref{implementation}, via both high-level discussion and excerpts from
+  our code in Scala.  It reveals the extent to which reagents turn patterns of
+  fine-grained concurrency into a general algorithmic framework.  It also shows
+  how certain choices in the design of reagents enable important optimizations
+  in their implementation.
+\end{itemize}
+Reagents have a clear cost model that provides an important guarantee: they do
+not impose extra overhead on the atomic updates performed by individual
+concurrent algorithms.  That is, a reagent-based algorithm manipulates shared
+memory in exactly the same way as its hand-written counterpart, even if it reads
+and writes many disparate locations.  This is a guarantee not generally provided
+by concurrency abstractions like software transactional memory (STM), which
+employ redo or undo logs or other mechanisms to support composable atomicity.
+Reagents only impose overheads when \emph{multiple} algorithms are sequenced
+into a large atomic block, \emph{e.g.}  in an atomic transfer between
+collections.
+
+\emph{In principle, then, reagents offer a strictly better situation than with
+  current libraries}: when used to express the algorithms provided by current
+libraries, reagents provide a higher level of abstraction yet impose negligible
+overhead; nothing is lost.  But unlike with current libraries, the algorithms
+can then be extended, tailored, and combined.  Extra costs are only paid when
+the new atomic compositions are used.
+
+We test this principle empirically in~\secref{performance} by comparing multiple
+reagent-based collections to their hand-written counterparts, as well as to
+lock-based and STM-based implementations.  The benchmarks include both single
+operations used in isolation (where little overhead for reagentes is expected)
+and combined atomic transfers.  We compare at both high and low levels of
+contention.  Reagents perform universally better than the lock- and STM-based
+implementations, and are indeed competetive with hand-written lock-free
+implementations.
+
+Finally, it is worth noting in passing that reagents generalize both Concurrent
+ML~\cite{?} and Transactional Events~\cite{?}, yielding the first lock-free
+implementation for both.  We give a thorough discussion of this and other
+related work---especially STM---in~\secref{related}.
+
+%CML and TE?
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Background}
 \label{sec:background}
-
-\cite{Shavit1997a} %TODO: cite pure elimination
 
@@
 \lstinline{tryPop} effectively cancel each other out.  This insight leads to a
-scheme called \emph{elimination backoff}~\cite{?}.  Operations first try the
-usual \lstinline{cas}-based code.  If the \lstinline{cas} fails, rather than
-busy-waiting, the operations advertise their presence on a side-channel,
-reducing the contention for the head pointer.  If a \lstinline{push} and
-\lstinline{tryPop} detect their mutual presence, the \lstinline{push} can pass
-its argument directly to \lstinline{tryPop} through the side-channel, and no
-change to the head pointer is necessary.  Atomicity is not violated, because had
-the \lstinline{push} and \lstinline{pop} executed in sequence, the head pointer
-would have been returned to its original value anyway.
+scheme called \emph{elimination backoff}~\cite{Shavit1997a,Hendler2004}.
+Operations first try the usual \lstinline{cas}-based code.  If the
+\lstinline{cas} fails, rather than busy-waiting, the operations advertise their
+presence on a side-channel, reducing the contention for the head pointer.  If a
+\lstinline{push} and \lstinline{tryPop} detect their mutual presence, the
+\lstinline{push} can pass its argument directly to \lstinline{tryPop} through
+the side-channel, and no change to the head pointer is necessary.  Atomicity is
+not violated, because had the \lstinline{push} and \lstinline{pop} executed in
+sequence, the head pointer would have been returned to its original value
+anyway.
 
@@
 }
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Being reagents, \lstinline{push} and \lstinline{tryPop} are inert values.
-They can be invoked using the \lstinline{!} method provided by reagents; for a
-\lstinline{Reagent[A,B]} the \lstinline{!} method takes an \lstinline{A} and
-returns a \lstinline{B}.  Scala permits infix notation for methods, so we can
-use a \lstinline{TreiberStack} as follows:
+Being reagents, \lstinline{push} and \lstinline{tryPop} are inert values.  They
+can be invoked using the \lstinline{!} method provided by reagents, which is
+read ``react''.  For a \lstinline{Reagent[A,B]} the \lstinline{!} method takes
+an \lstinline{A} and returns a \lstinline{B}.  Scala permits infix notation for
+methods, so we can use a \lstinline{TreiberStack} as follows:
 \begin{lstlisting}
@@
 \subsection{Synchronization: how reagents react}
+\label{sec:sync}
 
@@
 Reagents can be composed in unanticpated ways.
+
+% TODO: ``remarkable'' the emergent behavior.  explain in more detail what
+% emerges -- the potential for transfer eliminations
 
@@
 
+%TODO: note tail recursion
+
 \elide{
@@
 
+Talk about more sophisticated algorithms, e.g. skiplists
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Implementation}
@@
 
+%TODO: talk about dissolve
+
+\elide{
+High level concepts:
+  Data abstraction/internal normalized syntactic representation
+  Continuations/Commit
+  Reactions
+  Retry versus Block
+  Offers
+
+Specific combinators:
+  Choice
+  CAS - esp. when 1CAS vs kCAS
+  Upd?
+  Channels - including bag implementation
     bags allow greatest degree of freedom and parallelism
-
-kcas implementations: \cite{Attiya2008,Fraser2007,Luchangco2003}
-
+  Computed?
+  Sequencing (works by continuation plumbing)
+
+Low-level details:
+  kcas implementation
+  Inlining
+}
+
+Having seen the design and application of reagents, we now turn to their
+implementation.  We begin with a high-level overview~(\secref{impl-approach})
+introducing the key techniques and data structures we use.  We then delve into
+the core code of our Scala implementation, beginning with the primary entry
+point (the \lstinline{!} method,~\secref{impl-reaction}) and continuing with the
+key combinators.
+
+\subsection{The basic approach}
+\label{sec:impl-approach}
+
+When invoked, reagents attempt to \emph{react}, which is \emph{conceptually} a
+two phase process: first, the desired reaction is built up; second, the reaction
+is atomically committed.  We emphasize ``conceptually'' because, as discussed in
+the introduction, reagents are designed to avoid this kind of overhead in the
+common case.  We first discuss the general case (which imposes overhead) but
+return momentarily to the common (no overhead) case.
+
+An attempt to react can fail during either phase.  A failure during the first
+phase, \emph{i.e.} a failure to build up the desired reaction, is always a
+permanent failure~(\secref{choice}).  Permanent failures indicate that the
+reagent cannot proceed given current conditions, and should therefore block
+until another thread intervenes and causes conditions to change.  On the other
+hand, a failure during the second phase, \emph{i.e.} a failure to commit, is
+always a transient failure~(\secref{choice}).  Transient failures indicate that
+the reagent should retry, since the reaction was halted due to active
+interference from another thread.  In slogan form, we have
+\begin{quotation}\noindent
+\begin{tabular}{l@{\ }l}
+Permanent failure:&wait for intervention\\
+Transient failure:&retry due to interference
+\end{tabular}
+\end{quotation}
+In general, a reaction encompasses three pieces of data:
+\begin{enumerate}
+\item a list of CASes to be performed,
+\item a list of messages to be consumed, and
+\item a list of actions to be performed after committing.
+\end{enumerate}
+It thus resembles the redo log used in some STM implementations~\cite{?}.
+
+In the common case that a reagent only performs one visible~(\secref{dynamic})
+CAS or message swap, those components of the reaction are not necessary and
+hence are not used.  Instead, the CAS or swap is performed immediately,
+compressing the two phases of reaction.  Aside from avoiding extra allocations,
+this key optimization means that in the common case a \lstinline{cas} or
+\lstinline{upd} in a reagents leads to exactly one executed CAS during reaction,
+with no extra overhead.  When a reaction encompasses multiple visible CASes, a
+costlier \emph{k}CAS protocol must be used to ensure atomicity.  We discuss the
+\emph{k}CAS protocol in~\secref{impl-commit}, and the common case single CAS
+in~\secref{impl-state}.
+
+In the implementation, \lstinline{Reagent[A,B]} is an abstract class all of
+whose subclasses are hidden.  The subclasses roughly correspond to the
+combinator functions (which are responsible for instantiating them), and
+instances of the subclasses store the arguments given to their corresponding
+combinator.  Each subclass provides an implementation of the
+\lstinline{tryReact} method, which is an abstract method of
+\lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
+  programming afficianados should regard \lstinline{tryReact} as a kind of
+  Kleisli arrow for reagents.}
 \begin{lstlisting}
-def !(a: A): B = tryReact(a, Reaction.inert, null) match {
+  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any
+\end{lstlisting}
+The \lstinline{Any} type in Scala lies at the top of the subtyping hierarchy,
+akin to \lstinline{Object} in Java.  Here we are using \lstinline{Any} to
+represent a union of the type \lstinline{B} with the type
+\lstinline{BacktrackCommand}, where the latter has just two singleton instances,
+\lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
+transient failures.
+
+In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to the
+reagent and the reaction built up so far, and either completes the reaction and
+returns its result (type \lstinline{B}) or one of the failure singletons
+(\lstinline{Block} or \lstinline{Retry}).
+
+\subsection{The entry point: reacting}
+\label{sec:impl-reaction}
+
+shown in \figref{reaction}
+
+\begin{figure}
+\begin{lstlisting}
+def !(a: A): B = tryReact(a, empty, null) match {
   case (_: BacktrackCommand) => {
@@
     val maySync = this.maySync // cache
-    @tailrec def retryLoop(shouldBlock: Boolean): B = {
+    def retryLoop(shouldBlock: Boolean): B = {
       val wait = maySync || shouldBlock
@@
 
-      tryReact(a, Reaction.inert, waiter) match {
+      tryReact(a, empty, waiter) match {
         case (bc: BacktrackCommand) if wait => {
@@
 \end{lstlisting}
+\nocaptionrule
+\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
+\label{fig:reaction}
+\end{figure}
+%TODO: note tail recursion
+
+\subsection{The exit point: committing}
+\label{sec:impl-commit}
+
+\subsection{Shared state}
+\label{sec:impl-state}
+
+kcas implementations: \cite{Fraser2007,Luchangco2003,Attiya2008}
 
 \begin{lstlisting}
-def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
+def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
+  if (!rx.hasCAS && !k.hasCAS) {
+    if (ref.cas(oldV, newV))
+      k.tryReact((), rx, offer)
+    else Retry
+  } else {
+    k.tryReact((), rx.withCAS(ref, oldV, newV), offer)
+  }
+\end{lstlisting}
+
+\subsection{Choice}
+
+\begin{lstlisting}
+def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
   r1.tryReact(a, rx, offer) match {
@@
         case Retry => Retry
-        case Block => Retry // must retry r1
+        case Block => Retry // must still retry r1
         case ans   => ans
@@
 
-\begin{lstlisting}
-  def tryReact(u: Unit, rx: Reaction, offer: Offer[B]): Any = 
-    if (rx.canCASImmediate(k, offer)) {
-      if (casI(expect, update))
-        k.tryReact((), rx, offer)
-      else Retry
-    } else k.tryReact((), rx.withCAS(Ref.this, expect, update), offer)
-\end{lstlisting}
+\subsection{Computed reagents}
+
+\subsection{Channels}
+
+Message passing is by far the most complicated aspect of the reagent
+implementation.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\newpage
+\begin{figure*}
+\newcommand{\gscale}{0.72}
+\hspace{-30pt}
+\begin{tabular}[t]{l@{\hspace{2mm}}c}
+\begin{sideways}
+\hspace{0.5in}
+\textbf{Throughput}: \quad
+  iterations/$\mu$s \quad
+  (bigger is better)
+\end{sideways} &
+\begin{tabular}[b]{l}
+\includegraphics[scale=\gscale]{graphs/PushPop-100.pdf}
+\hspace{1mm}
+\includegraphics[scale=\gscale]{graphs/PushPop-1000.pdf}
+\hspace{4mm}
+\includegraphics[scale=\gscale]{graphs/StackTransfer-100.pdf}
+\hspace{1mm}
+\includegraphics[scale=\gscale]{graphs/StackTransfer-1000.pdf}
+\\
+\\
+\includegraphics[scale=\gscale]{graphs/EnqDeq-100.pdf}
+\hspace{1mm}
+\includegraphics[scale=\gscale]{graphs/EnqDeq-1000.pdf}
+\hspace{4mm}
+\includegraphics[scale=\gscale]{graphs/QueueTransfer-100.pdf}
+\hspace{1mm}
+\includegraphics[scale=\gscale]{graphs/QueueTransfer-1000.pdf}
+\end{tabular}
+\\
+\\
+& \textbf{Threads}\quad (on 16-way machine)
+\end{tabular}
+\end{figure*}
+
+\subsection{Methodology and experimental setup}
+
+Fine-grained concurrent data structures are usually evaluated by targetted
+microbenchmarking, with focus on contention effects and fine-grained parallel
+speedup~\cite{?}.  In addition to those basic aims, we wish to evaluate (1) the
+extent to which reagent-based algorithms can compete with their hand-built
+counterparts and (2) whether reagent composition is a plausible approach for
+scalable atomic transfers.
+
+To this end, we designed a series of benchmarks
+
+0.25$\mu$s, 2.5$\mu$s
+
+Our benchmarks were run on a 3.46Ghz Intel Xeon X5677 (Westmere) with 32GB RAM
+and 12MB of shared L3 cache.  The machine has two physical processors with
+four hyperthreaded cores each, for a total of 16 hardware threads.  L1 and L2
+caches are per-core.  The software environment includes Ubuntu 10.04.3 and
+the Hotspot JVM 6u27.
+
+\subsection{Results}
+
+Performance measurements from our initial implementation show that reagents are
+competetive with hand-built concurrent data structures, yet also provide
+scalable composed operations that are rarely provided for such data structures.
+
+\subsection{Potential improvements}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Related work}
@@
 
+We see this work as serving the needs of two distinct groups: concurrency
+experts and concurrency users.  Using reagents, experts can write libraries more
+easily, because common patterns are expressible as abstractions and many are
+built-in.  Users can then extend, tailor and compose the resulting library
+without detailed knowledge of the algorithms involved.
+
 \bibliographystyle{abbrvnat}
@@
 1.5 Semantics 
-1.5 Implementation/algorithms           2.0
+1.5 Implementation/algorithms           2.2
 0.5 Performance results                 1.0
@@
 0.3 Conclusion                          0.3
-0.8 Bib                                 0.7
+0.8 Bib                                 0.5
 

### 2011-11-06 paper/paper.tex
writing

@@
 High level concepts:
-  Data abstraction/internal normalized syntactic representation
+x Reactions
+x Data abstraction/internal normalized syntactic representation
+x Retry versus Block
   Continuations/Commit
-  Reactions
-  Retry versus Block
   Offers
@@
 represent a union of the type \lstinline{B} with the type
-\lstinline{BacktrackCommand}, where the latter has just two singleton instances,
+\lstinline{Failure}, where the latter has just two singleton instances,
 \lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
@@
 In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to the
-reagent and the reaction built up so far, and either completes the reaction and
-returns its result (type \lstinline{B}) or one of the failure singletons
-(\lstinline{Block} or \lstinline{Retry}).
+reagent and the reaction built up so far, and either completes the reaction,
+returning a result (type \listinline{B}), or fails, returning one of the
+failure singletons (\lstinline{Block} or \lstinline{Retry}).
 
@@
 
+\elide{
+def !(a: A): B = tryReact(a, empty, null) match {
+  case (_: Failure) => {
+    val backoff = new Backoff
+    def retryLoop(shouldBlock: Boolean): B = {
+      val offer = if (maySync || shouldBlock) 
+                    new Offer[B]
+                  else null
+      tryReact(a, empty, offer) match {
+        case (f: Failure) => {
+          if (f == Block && offer != null) 
+            LockSupport.park 
+          else backoff.once
+
+          // rescind offer, but check for completion
+          if (offer != null && offer.tryAbort)
+            offer.answer
+          else
+            retryLoop(f == Block)
+        }
+        case ans   => ans.asInstanceOf[B]
+      }
+    }
+    backoff.once
+    retryLoop(false)
+  }
+  case ans => ans.asInstanceOf[B]
+}
+}
+
 \begin{figure}
 \begin{lstlisting}
-def !(a: A): B = tryReact(a, empty, null) match {
-  case (_: BacktrackCommand) => {
+def !(a: A): B = {
   val backoff = new Backoff
-    val maySync = this.maySync // cache
   def retryLoop(shouldBlock: Boolean): B = {
-      val wait = maySync || shouldBlock
-      val waiter = if (wait) new Waiter[B](shouldBlock) 
-                   else null
-
-      tryReact(a, empty, waiter) match {
-        case (bc: BacktrackCommand) if wait => {
-          bc.bottom(waiter, backoff, snoop(a))
-          waiter.tryAbort      // rescind waiter,
-          waiter.poll match { // but check if completed
-            case Some(ans) => ans.asInstanceOf[B] 
-            case None       => retryLoop(bc.isBlock)
-          }
-        }
-        case Retry => backoff.once; retryLoop(false)
+    if (maySync || shouldBlock) {
+      val offer = new Offer[B]
+      tryReact(a, empty, offer) match {
+        case Block => 
+          LockSupport.park
+          if (offer.rescind) retryLoop(true)
+          else offer.answer
+        case Retry =>
+          backoff.once
+          if (offer.rescind) retryLoop(maySync)
+          else offer.answer
+      }
+    } else {
+      tryReact(a, empty, null) match {
         case Block => retryLoop(true)
+        case Retry => backoff.once; retryLoop(maySync)
         case ans   => ans.asInstanceOf[B]
@@
   retryLoop(false)
-  }
-  case ans => ans.asInstanceOf[B]
 }
@@
 \subsection{Computed reagents}
+
+\begin{lstlisting}
+  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
+    c(a).tryReact((), rx, offer)
+\end{lstlisting}
 

### 2011-11-06 paper/paper.tex
work on implementation writing

@@
 lock-based and STM-based implementations.  The benchmarks include both single
-operations used in isolation (where little overhead for reagentes is expected)
+operations used in isolation (where little overhead for reagents is expected)
 and combined atomic transfers.  We compare at both high and low levels of
@@
 proceed.  When faced with a permanent failure, a reagent should block
-until signalled that the underlying state has changed.
+until signaled that the underlying state has changed.
 
@@
 require the stack user to know anything about the implementation.
-Reagents can be composed in unanticpated ways.
+Reagents can be composed in unanticipated ways.
 
@@
 
-In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to the
-reagent and the reaction built up so far, and either completes the reaction,
-returning a result (type \listinline{B}), or fails, returning one of the
-failure singletons (\lstinline{Block} or \lstinline{Retry}).
+In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to
+the reagent and the reaction built up so far, and either completes the
+reaction, returning a result (type \listinline{B}), or fails, returning one of
+the failure singletons (\lstinline{Block} or \lstinline{Retry}).  The
+remaining argument, \lstinline{offer}, is used for synchronization and
+communication between reagents, which we explain next.
+
+\subsection{Offers, communication and continuations}
+\label{sec:offers}
 
@@
 \label{sec:impl-reaction}
-
-shown in \figref{reaction}
 
@@
   val backoff = new Backoff
-  def retryLoop(shouldBlock: Boolean): B = {
-    if (maySync || shouldBlock) {
+  def withoutOffer(): B =
+    tryReact(a, empty, null) match {
+      case Block => withOffer()
+      case Retry => 
+        backoff.once()
+        if (maySync) withOffer() else withoutOffer()
+      case ans   => ans.asInstanceOf[B]
+    }
+  def withOffer(): B = {
     val offer = new Offer[B]
     tryReact(a, empty, offer) match {
-        case Block => 
-          LockSupport.park
-          if (offer.rescind) retryLoop(true)
-          else offer.answer
-        case Retry =>
-          backoff.once
-          if (offer.rescind) retryLoop(maySync)
-          else offer.answer
-      }
-    } else {
-      tryReact(a, empty, null) match {
-        case Block => retryLoop(true)
-        case Retry => backoff.once; retryLoop(maySync)
+      case (f: Failure) => 
+        if (f == Block) park() else backoff.once()
+        if (offer.rescind) withOffer() else offer.answer
       case ans => ans.asInstanceOf[B]
@@
   }
-  retryLoop(false)
+  withoutOffer()
 }
@@
 
-\subsection{The exit point: committing}
-\label{sec:impl-commit}
-
-\subsection{Shared state}
-\label{sec:impl-state}
-
-kcas implementations: \cite{Fraser2007,Luchangco2003,Attiya2008}
-
-\begin{lstlisting}
+The code for performing a reaction is given in the \lstinline{!} method
+definition for \lstinline{Reagent[A,B]}, shown in~\figref{reaction}.  This
+method provides two generalized versions of the optimistic retry loops we
+described in~\secref{background}.  The retry loops are written as a local,
+tail-recursive functions, which Scala compiles down to loops.
+
+The first retry loop, \lstinline{withoutOffer}, attempts to perform the
+reaction without making visible offers to other reagents.  It may, however,
+find and consume offers from other reagents as necessary for message passing.
+To initiate the reaction, \lstinline{withoutOffer} calls the abstract
+\lstinline{tryReact} method with the input \lstinline{a}, an empty reaction to
+start with, and no offer.  If the reaction fails in the first phase (a
+permanent failure, represented by \lstinline{Block}), the next attempt must be
+made with an offer, to set up the blocking/signaling protocol.  If the
+reaction fails in the second phase (a transient failure, represented by
+\lstinline{Retry}), there is likely contention over shared data.  To reduce
+the contention, it performs one cycle of exponential backoff before retrying.
+If the reagent includes communication attempts, the retry is performed with an
+offer, since doing so increases chances of elimination~(\secref{choice})
+without further contention.  Finally, if both phases of the reaction succeed,
+the final answer is returned.
+
+The second retry loop, \lstinline{withOffer}, is similar, but begins by
+allocating an \lstinline{Offer} object to make visible to other reagents.
+%TODO: finish this
+
+Initially, the reaction is attempted using \lstinline{withoutOffer},
+representing optimism that the reaction can be completed without making a
+visible offer.
+
+\begin{figure}
+\begin{lstlisting}[frame=single]
+class Commit[A] extends Reagent[A,A] {
+  def tryReact(a: A, rx: Reaction, offer: Offer[A]) = 
+    if (offer != null && !offer.rescind) offer.answer
+    else if (rx.commit) a 
+    else Retry
+}
+\end{lstlisting}
+\vskip -12pt
+\begin{lstlisting}[frame=single]
+class CAS[A](ref: Ref[A], oldV: A, newV: A) 
+  extends Reagent[Unit,Unit] {
   def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
-  if (!rx.hasCAS && !k.hasCAS) {
+    if (!rx.hasCAS && !k.hasCAS)
       if (ref.cas(oldV, newV))
@@
       else Retry
-  } else {
+    else
       k.tryReact((), rx.withCAS(ref, oldV, newV), offer)
@@
 \end{lstlisting}
-
-\subsection{Choice}
-
-\begin{lstlisting}
+\vskip -12pt
+\begin{lstlisting}[frame=single]
+class Choice[A,B](r1: Reagent[A,B], r2: Reagent[A,B]) 
+  extends Reagent[A,B] {
   def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
@@
     }
+}
 \end{lstlisting}
+\vskip -12pt
+\begin{lstlisting}[frame=single]
+class Computed[A,B](c: A => Reagent[Unit,B]) 
+  extends Reagent[A,B] {
+  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
+    c(a).tryReact((), rx, offer)
+}
+\end{lstlisting}
+\nocaptionrule
+\caption{Excerpts from core reagent subclasses}
+\end{figure}
+
+\subsection{The exit point: committing}
+\label{sec:impl-commit}
+
+As mentioned in~\secref{offers}, the initial continuation for reagents is the
+\lstinline{Commit} continuation:
+
+If the reagent has made an offer, but has also completed the first phase of
+reaction, the offer must be rescinded before the commit phase is
+attempted---otherwise, the reaction could complete twice.  As with the
+\lstinline{!} method, the attempt to rescind the offer is in a race with other
+reagents that may be completing the offer.  If \lstinline{Commit} loses the
+race, it returns the answer provided by the offer.  Otherwise, it attempts to
+commit the reaction, and if successful simply returns its input, which is the
+final answer for the reaction.
+
+Committing a reaction requires a $k$CAS operation: $k$ compare and sets must
+be performed atomically.  This operation, which forms the basis of STM, is in
+general expensive and not available through hardware.  There are several
+software implementations that provide various progress
+guarantees~\cite{Fraser2007,Luchangco2003,Attiya2008}, but they tend to
+
+\subsection{Shared state}
+\label{sec:impl-state}
+
+\subsection{Choice}
 
 \subsection{Computed reagents}
-
-\begin{lstlisting}
-  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
-    c(a).tryReact((), rx, offer)
-\end{lstlisting}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Performance}
-\label{sec:performance}
 
@@
 
+\section{Performance}
+\label{sec:performance}
+
 \subsection{Methodology and experimental setup}

### 2011-11-06 paper/paper.tex
writing on implementation and dynamic reagents

@@
 
+%TODO: cite art of multiprocessor programming
+
 Given our target of cache-coherent, shared-memory architectures, the
@@
 \begin{figure}
-\begin{lstlisting}
+\begin{lstlisting}[frame=single]
 class TreiberStack[A] {
@@
 In \secref{background}, we saw a range of strategies for communication and
-synchronizion at a fine grain:
+synchronization at a fine grain:
 \begin{itemize}
@@
 >> : Reagent[A,B] * Reagent[B,C] => Reagent[A,C]
-|||  : Reagent[A,B] * Reagent[A,C] => Reageng[A,(B,C)]
+|||  : Reagent[A,B] * Reagent[A,C] => Reagent[A,(B,C)]
 
@@
 \end{lstlisting}
+\nocaptionrule
+\caption{The Reagent API}
 \label{fig:combinators}
@@
 analyses are exhaustive, so the update functions are in fact total.
+\lstinline{Unit} is akin to \lstinline{void}: it is a type with a single
+member, written \lstinline{()}.
 
@@
 introduces it as a catalyst.\footnote{For simplicity, we have not given a way
-  to \emph{cancel} catalysts after they have been introduced, but cancellation
-  is easy to add.}  Operationally, in this example, that just means sending
+  to cancel catalysts after they have been introduced, but cancellation is
+  easy to add.}  Operationally, in this example, that just means sending
 messages along \lstinline{in1} and \listinline{in2} that are marked as
@@
 \begin{figure}
-\begin{lstlisting}
+\begin{lstlisting}[frame=single]
 class MSQueue[A] {
   private case class Node(data: A, next: Ref[Node])
-  private val head = new Ref(Node(null))
-  private val tail = new Ref(head.read ! ())
+  private val head = new Ref(new Node(null))  // sentinel
+  private val tail = new Ref(read(head) ! ()) // sentinel
   val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
@@
   val enq: Reagent[A, Unit] = computed { 
-    (x: A) => findAndEnq(Node(x, new Ref(null)))
+    (x: A) => findAndEnq(new Node(x, new Ref(null)))
   }
@@
 An example of this phenomenon is the classic Michael-Scott lock-free
-queue~\cite{?}.
+queue~\cite{Michael1996}.  Unlike a stack, in which all activity focuses on
+the head, queues have two loci of updates.  That means, in particular, that
+the \lstinline{Ref}s used by its reagents may vary depending on the current
+state of the queue.
+
+Here is a brief overview of the Michael-Scott queue.  The queue is represented
+as a \emph{mutable} linked list, with a sentinel node at the head (front) of
+the queue.  The head pointer always points to the current sentinel node; nodes
+are dequeues by a CAS to this pointer, just like Treiber stacks (but lagged by
+one node).  The tail of the queue is the unique node, reachable from the head
+pointer, with a null \lstinline{next} pointer; thanks to the sentinel, such a
+node is guaranteed to exist.  If the queue is empty, the same node will be the
+head (sentinel) and tail.  Finally, as an optimization for enqueing, a tail
+\emph{pointer} is maintained with the invariant that the real tail is always
+reachable from it.  The tail pointer may lag behind, however, which allows the
+algorithm to work using only single-word CAS instructions.
+
+Our reagent-based implementation of the Michael-Scott queue is shown
+in~\figref{msqueue}.  The node representation is given as an inner \emph{case}
+class.  In Scala, case classes provide two features we take advantage of.
+First, the parameters to their constructors (here \lstinline{data} and
+\lstinline{next}) are automatically added as final fields to the class, which
+are initialized to the constructor argument values.  Second, they extend
+pattern matching through \lstinline{case} so that instances can be
+\emph{deconstructed}.  A pattern like \lstinline{case Node(d, n)} matches any
+instance of the node class, binding \lstinline{d} to its \lstinline{data}
+field and \lstinline{n} to its \lstinline{next} field.
+
+The \lstinline{head} and \lstinline{tail} references of the queue are
+initialized to the same sentinel node.  Here we use a new reagent,
+\lstinline{read}~(\figref{combinators}), to extract the sentinel value from
+the head when constructing the tail.  The read reagent in \lstinline{tail}'s
+initializing expression is immediately executed during construction of an
+\lstinline{MSQueue} instance.
+
+The \lstinline{tryDeq} reagent is very similar to the \lstinline{tryPop}
+reagent in \lstinline{TreiberStack}, modulo the sentinel node.  The reagent
+pattern matches on the sentinel node, ignoring its \lstinline{data} field by
+using \lstinline{_}, the wildcard.  The \lstinline{next} field is then matched
+to a nested pattern, \lstinline{Ref(n@Node(x, _))}.  This pattern immediately
+reads the current value of the reference stored in \lstinline{next}, binds
+that value to \lstinline{n}, and then matches the
+pattern~\lstinline{Node(x,_)} against \lstinline{n}.  If the pattern
+matches---which it will any time the \lstinline{next} field of the sentinel is
+non-null---the node \lstinline{n} becomes the new \lstinline{head} (and hence
+the new sentinel).  
+%TODO: picture?
+
+Since the location of the tail node is determined dynamically by the data in
+the queue, the \lstinline{enq} reagent must itself be determined dynamically.
+
+The \lstinline{enq} reagent, on the other hand, must work with the shifting
+position of the tail, which requires new combinators.
+
+The \lstinline{cas} combinator~(\figref{combinators}) takes a
+\lstinline{Ref[A]} and two \lstinline{A} arguments, giving the expected and
+updated values, respectively.  Unlike its counterpart for
+\lstinline{AtomicReference}, a \lstinline{cas} reagent does \emph{not} yield a
+boolean result.  Instead, the reagent library builds in handling of failed
+CASes, including an optimistic retry loop and exponential backoff.  A failure
+to CAS is, as discussed in \secref{choice}, a transient failure that causes
+the reagent to retry.
+
+%TODO: treat !? as shorthand?
 
@@
 
-\elide{
-The \lstinline{read} combinator has a straightforward signature: if
+\elide{ The \lstinline{read} combinator has a straightforward signature: if
   \lstinline{r} has type \lstinline{Ref[A]}, then \lstinline{read(r)} is a
-\emph{value} of type \lstinline{Reagent[Unit, A]} which, when invoked, takes a
-unit\footnote{\lstinline{Unit} in Scala is akin to \lstinline{void}: it is a
-  type with a single member, written \lstinline{()}.} argument and yields an
-\lstinline{A} result.
-
-The \lstinline{cas} combinator takes two additional \lstinline{A} arguments,
-giving the expected and updated values, respectively.  Unlike its counterpart
-for \lstinline{AtomicReference}, a \lstinline{cas} reagent does \emph{not}
-yield a boolean result.  Instead, the reagent library builds in handling of
-failed CASes, including an optimistic retry loop and exponential backoff.  The
-failure mechanics depend on which other combinators are involved, so we delay
-a complete discussion until \secref{failure}.
-}
-
-% tail recursive
+  \emph{value} of type \lstinline{Reagent[Unit, A]} which, when invoked, takes
+  a unit argument and yields an \lstinline{A} result.
+}
 
@@
 
-\paragraph{A caveat for composition}
-
 Talk about more sophisticated algorithms, e.g. skiplists
+
+\subsection{Caveats for composition}
 
@@
 \label{sec:implementation}
+
+\elide{
+def !(a: A): B = tryReact(a, empty, null) match {
+  case (_: Failure) => {
+    val backoff = new Backoff
+    def retryLoop(shouldBlock: Boolean): B = {
+      val offer = if (maySync || shouldBlock) 
+                    new Offer[B]
+                  else null
+      tryReact(a, empty, offer) match {
+        case (f: Failure) => {
+          if (f == Block && offer != null) 
+            LockSupport.park 
+          else backoff.once
+
+          // rescind offer, but check for completion
+          if (offer != null && offer.tryAbort)
+            offer.answer
+          else
+            retryLoop(f == Block)
+        }
+        case ans   => ans.asInstanceOf[B]
+      }
+    }
+    backoff.once
+    retryLoop(false)
+  }
+  case ans => ans.asInstanceOf[B]
+}
+}
+
+\begin{figure}
+\begin{lstlisting}[frame=single]
+def !(a: A): B = {
+  val backoff = new Backoff
+  def withoutOffer(): B =
+    tryReact(a, empty, null) match {
+      case Block => withOffer()
+      case Retry => 
+        backoff.once()
+        if (maySync) withOffer() else withoutOffer()
+      case ans   => ans.asInstanceOf[B]
+    }
+  def withOffer(): B = {
+    val offer = new Offer[B]
+    tryReact(a, empty, offer) match {
+      case (f: Failure) => 
+        if (f == Block) park() else backoff.once()
+        if (offer.rescind) withOffer() else offer.answer
+      case ans => ans.asInstanceOf[B]
+    }
+  }
+  withoutOffer()
+}
+\end{lstlisting}
+\nocaptionrule
+\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
+\label{fig:reaction}
+\end{figure}
 
@@
 \lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
-  programming afficianados should regard \lstinline{tryReact} as a kind of
+  programming aficionados should regard \lstinline{tryReact} as a kind of
   Kleisli arrow for reagents.}
@@
 \label{sec:impl-reaction}
-
-\elide{
-def !(a: A): B = tryReact(a, empty, null) match {
-  case (_: Failure) => {
-    val backoff = new Backoff
-    def retryLoop(shouldBlock: Boolean): B = {
-      val offer = if (maySync || shouldBlock) 
-                    new Offer[B]
-                  else null
-      tryReact(a, empty, offer) match {
-        case (f: Failure) => {
-          if (f == Block && offer != null) 
-            LockSupport.park 
-          else backoff.once
-
-          // rescind offer, but check for completion
-          if (offer != null && offer.tryAbort)
-            offer.answer
-          else
-            retryLoop(f == Block)
-        }
-        case ans   => ans.asInstanceOf[B]
-      }
-    }
-    backoff.once
-    retryLoop(false)
-  }
-  case ans => ans.asInstanceOf[B]
-}
-}
-
-\begin{figure}
-\begin{lstlisting}
-def !(a: A): B = {
-  val backoff = new Backoff
-  def withoutOffer(): B =
-    tryReact(a, empty, null) match {
-      case Block => withOffer()
-      case Retry => 
-        backoff.once()
-        if (maySync) withOffer() else withoutOffer()
-      case ans   => ans.asInstanceOf[B]
-    }
-  def withOffer(): B = {
-    val offer = new Offer[B]
-    tryReact(a, empty, offer) match {
-      case (f: Failure) => 
-        if (f == Block) park() else backoff.once()
-        if (offer.rescind) withOffer() else offer.answer
-      case ans => ans.asInstanceOf[B]
-    }
-  }
-  withoutOffer()
-}
-\end{lstlisting}
-\nocaptionrule
-\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
-\label{fig:reaction}
-\end{figure}
 %TODO: note tail recursion
@@
 \begin{lstlisting}[frame=single]
-class CAS[A](ref: Ref[A], oldV: A, newV: A) 
-  extends Reagent[Unit,Unit] {
+class CAS[A,B](ref: Ref[A], ov: A, nv: A, k: Reagent[A,B]) 
+  extends Reagent[Unit,B] {
   def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
     if (!rx.hasCAS && !k.hasCAS)
-      if (ref.cas(oldV, newV))
+      if (ref.cas(ov, nv))
         k.tryReact((), rx, offer)
@@
     else
-      k.tryReact((), rx.withCAS(ref, oldV, newV), offer)
+      k.tryReact((), rx.withCAS(ref, ov, nv), offer)
 }
@@
 implementation.
+
+\subsection{Catalysis}
 

### 2011-11-07 paper/paper.tex
nearly done

@@
 Ahmdahl's law tells us that sequential bottlenecks fundamentally limit our
-profit from parallelism~\cite{?}.  In practice, the effect is amplified by
+profit from parallelism\elide{~\cite{?}}.  In practice, the effect is amplified by
 another factor: interprocessor communication, often in the form of cache
@@
 Another answer is: communicate efficiently.  The past two
-decades\footnote{Herlihy/Wing~\cite{?}, Mellor-Crummey/Scott~\cite{?}} have
+decades\elide{\footnote{Herlihy/Wing~\cite{?}, Mellor-Crummey/Scott~\cite{?}}} have
 produced a sizable collection of algorithms for synchronization, communication,
@@
 libraries---\texttt{java.util.concurrent} (JUC) the most prominent---offering a
-wide range of concurrency primitives appropriate for fine-grained parallelism.
+wide range of primitives appropriate for fine-grained parallelism.
 
@@
 Finally, it is worth noting in passing that reagents generalize both Concurrent
-ML~\cite{?} and Transactional Events~\cite{?}, yielding the first lock-free
-implementation for both.  We give a thorough discussion of this and other
-related work---especially STM---in~\secref{related}.
+ML~\cite{Reppy1991} and Transactional Events~\cite{Donnelly2006}, yielding the
+first lock-free implementation for both.  We give a thorough discussion of this
+and other related work---especially STM---in~\secref{related}.
 
@@
 \figref{classic-treiber} gives a classic example of a fine-grained concurrent
-data structure: Treiber's lock-free stack~\cite{?}.  The stack is represented as
-an \emph{immutable} linked list.  Mutation occurs solely through the
-\lstinline{head} pointer, represented here as an \lstinline{AtomicReference}.
-The \lstinline{head} is mutated using \lstinline{compareAndSet} (here
-abbreviated as \lstinline{cas}), which takes an expected value and a new value,
-and atomically updates the reference if it has the expected value.  It returns
-\lstinline{true} iff the update was successful.
+data structure: Treiber's lock-free stack~\cite{Treiber1986}.  The stack is
+represented as an \emph{immutable} linked list.  Mutation occurs solely through
+the \lstinline{head} pointer, represented here as an
+\lstinline{AtomicReference}.  The \lstinline{head} is mutated using
+\lstinline{compareAndSet} (here abbreviated as \lstinline{cas}), which takes an
+expected value and a new value, and atomically updates the reference if it has
+the expected value.  It returns \lstinline{true} iff the update was successful.
 
@@
 unlimited number of other operations on the stack are succeeding.  In other
-words, \lstinline{push} and \lstinline{tryPop} are formally \emph{lock
-  free}~\cite{?}.
+words, \lstinline{push} and \lstinline{tryPop} are formally lock
+  free\elide{~\cite{?}}.
 
 To successfully \lstinline{cas}, a core must acquire the relevant cache line in
-\emph{exclusive} mode~\cite{?}, or something conceptually equivalent.  When
+exclusive mode\elide{~\cite{?}}, or something conceptually equivalent.  When
 several cores attempt to \lstinline{cas} a common reference, they must
@@
 possible to cleverly treat signalling in essentially the same way as atomic
-updates~\cite{?}.
+updates~\cite{Scherer2004}.
 
 Synchronization and signalling is also subject to cache and memory bandwidth
-concerns.  For example, when a lock is released, how much memory traffic is
+concerns, but it would take us too far afield to discuss these in depth.
+Mellor-Crummey and Scott pioneered the now-common techniques for grappling with
+these concerns~\cite{Mellor-Crummey1991}, and we apply their techniques in
+implementing the blocking and signaling protocols for reagents.
+
+\elide{
+For example, when a lock is released, how much memory traffic is
 generated?  If the lock is implemented as a single bit which threads attempt to
@@
 should  % TODO: room for this??
-
-%talk about elimination backoff and blocking versions but don't show code.
-
-%discuss the insight behind MCS locks
-
-The goal of reagents is to systematize these techniques and others into a
+}
+
+\elide{
+The goal of reagents is to systematize the above techniques and others into a
 uniform set of combinators that make algorithms easier to express, and allows
@@
 see how to write all of the above examples, concisely and composably.
+}
 
@@
 
-\begin{figure}
+\begin{figure}\elide{
 \begin{lstlisting}[mathescape=true,frame=single]
@@
 postCommit: (A => Unit) => Reagent[A,A]
+\end{lstlisting}}
+\begin{lstlisting}[mathescape=true,frame=single]
+upd:  Ref[A] => PartialFunction[(A,B), (A,C)] 
+             => Reagent[B,C]
+read: Ref[A] => Reagent[Unit, A]
+cas:  Ref[A] * A * A => Reagent[Unit, Unit]
+swap: Endpoint[A,B] => Reagent[A,B]
++:    Reagent[A,B] * Reagent[A,B] => Reagent[A,B]
+>>:    Reagent[A,B] * Reagent[B,C] => Reagent[A,C]
+|||:    Reagent[A,B] * Reagent[A,C] => Reagent[A,(B,C)]  
+ret:  A => Reagent[Unit, A]
+computed: 
+      PartialFunction[A,Reagent[Unit,B]] => Reagent[A,B]
+postCommit: (A => Unit) => Reagent[A,A]
 \end{lstlisting}
 \nocaptionrule
-\caption{The Reagent API}
+\caption{The reagent combinators}
 \label{fig:combinators}
 \end{figure}
+
+\elide{
+// reagent invocation
+r ! a       // as a reactant, r: Reagent[A,B], a: A
+dissolve(r) // as a catalyst, r: Reagent[Unit, Unit]
+}
 
@@
 through mutation, synchronization, or both.
-
 Unlike $\lambda$-abstractions, which encompass arbitrary code and treat that
@@
 by applying additional combinators, without knowledge of the internal
-implementation.  At any point a user can invoke a reagent, either as a
+implementation.
+
+In this section, we introduce the reagent combinators and use them to build a
+series of increasingly complex fine-grained concurrent algorithms.  By the end
+of the section, we will have seen how to implement all of the algorithms
+described in~\secref{background}, and several more besides.
+
+\elide{
+At any point a user can invoke a reagent, either as a
 \emph{reactant} or a \emph{catalyst}---a distinction explained in
@@
 clear semantics (\secref{semantics}), ... %TODO
+}
 
@@
 Memory is shared between reagents using the type \lstinline{Ref[A]} of
-atomically-updatable references.\footnote{ These references are only
+atomically-updatable references.\elide{\footnote{ These references are only
   accessible through reagent combinators, which allows them to provide
   functionality beyond the built-in \lstinline{AtomicReference[A]} type, such
-  as the blocking logic we will see shortly.  }
+  as the blocking logic we will see shortly.  }}
 The \lstinline{upd} combinator~(\figref{combinators}) represents simple atomic
@@
 an \lstinline{A} and returns a \lstinline{B}.  Scala permits infix notation for
-methods, so we can use a \lstinline{TreiberStack} as follows:
-\begin{lstlisting}
-  var s = new TreiberStack[Int]; s.push ! 42
-\end{lstlisting}
-The key point is that when we invoke these reagents, we are executing
-\emph{exactly} the same algorithms written by hand in \secref{background},
-including the retry loop with exponential backoff; reagents systematize and
-internalize common patterns of fine-grained concurrency.  But by exposing
-\lstinline{push} and \lstinline{tryPop} as reagents, rather than methods, we
-leave the door open for further composition and tailoring by a user of the
-data structure (\secref{choice}, \secref{conjunction}).
+methods, so we can use a \lstinline{TreiberStack} \lstinline{s} by writing
+\lstinline{s.push ! 42}.  The key point is that when we invoke these reagents,
+we are executing \emph{exactly} the same algorithms written by hand in
+\secref{background}, including the retry loop with exponential backoff; reagents
+systematize and internalize common patterns of fine-grained concurrency.  By
+exposing \lstinline{push} and \lstinline{tryPop} as reagents rather than
+methods, we allow further composition and tailoring by a user of the data
+structure (\secref{choice}, \secref{conjunction}).
 
@@
 block the calling thread unless or until the stack is nonempty.  Blocking and
-signalling are entirely handled by the reagent implementation, using standard
-techniques~(\secref{implementation}); there is therefore no risk of a lost
-wakeup.  %TODO: add more here?
+signalling are entirely handled by the reagent implementation; there is
+therefore no risk of a lost wakeup.  %TODO: add more here?
 
@@
 
-Reagents interact through \emph{synchronous swap channels}, which
-consist of two complementary \emph{endpoints}:
-\begin{lstlisting}
-  mkChan[A,B]: (Endpoint[A,B], Endpoint[B,A])
-\end{lstlisting}
+Reagents interact through \emph{synchronous swap channels}, which consist of two
+complementary \emph{endpoints}.  The method \lstinline{mkChan[A,B]} returns a pair
+of type \lstinline{(Endpoint[A,B], Endpoint[B,A])}.
 The combinator for communication is \lstinline{swap} (see
@@
 
-If \lstinline{r} and \lstinline{s} are two reagents of the same type,
-their \emph{choice} \lstinline{r + s} will behave like one of them,
-nondeterministically, when invoked.  The most straightforward use of
-choice is in the spirit of Unix's \lstinline{select}~\cite{?}: waiting
-on several signals simultaneously, while consuming only one of them.
-For example, if \lstinline{c} and \lstinline{d} are endpoints of the
-same type, \lstinline{swap(c) + swap(d)} is a reagent that will accept
-exactly one message, either from \lstinline{c} or from \lstinline{d}.
-If neither endpoint has a message available, the reagent will block
-until one of them does.  As usual, the blocking protocol is handled by
-the reagent library.
-
-A more novel use of choice is adding backoff strategies (of the kind
-described in \secref{?}).  For example, we can build an elimination
-backoff stack as follows:
+If \lstinline{r} and \lstinline{s} are two reagents of the same type, their
+\emph{choice} \lstinline{r + s} will behave like one of them,
+nondeterministically, when invoked.  The most straightforward use of choice is
+in the spirit of Unix's \lstinline{select}\elide{~\cite{?}}: waiting on several
+signals simultaneously, while consuming only one of them.  For example, if
+\lstinline{c} and \lstinline{d} are endpoints of the same type,
+\lstinline{swap(c) + swap(d)} is a reagent that will accept exactly one message,
+either from \lstinline{c} or from \lstinline{d}.  If neither endpoint has a
+message available, the reagent will block until one of them does.  As usual, the
+blocking protocol is handled by the reagent library.
+
+A more novel use of choice is adding backoff strategies (of the kind described
+in \secref{background}).  For example, we can build an elimination backoff stack
+as follows:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Choice is left-biased,\footnote{The left bias makes choice more
+Choice is left-biased,\elide{\footnote{The left bias makes choice more
   similar to \lstinline{orElse} in Haskell's STM than to CML's choice,
-  but there are differences from both.  See \secref{related}.  } so
+  but there are differences from both.  See \secref{related}.  }} so
 when \lstinline{push} is invoked, it will first attempt to push onto
@@
 
-The double-pop reagent implicitly gives a solution to the Dining
-Philosophers problem~\cite{?}: two resources are atomically acquired.
-Internally, the reagent library ensures dead- and live-lock freedom
-for conjoined reagents.
+The double-pop reagent implicitly gives a solution to the Dining Philosophers
+problem\elide{~\cite{?}}: the two resources are atomically acquired.
+%% Internally, the reagent library ensures dead- and live-lock freedom
+%% for conjoined reagents.
 
@@
 
-In chemistry, a reagent is a participant in a reaction, and reagents are
-further subdivided into \emph{reactants}, which are consumed during reaction,
-and \emph{catalysts}, which enable reactions but are not consumed by them.
+In chemistry, a reagent is a participant in a reaction; reagents are
+subdivided into \emph{reactants}, which are consumed during reaction, and
+\emph{catalysts}, which enable reactions but are not consumed by them.
 
@@
 \begin{lstlisting}
-def zip(in1: Endpoint[Unit,  A], 
-        in2: Endpoint[Unit,  B], 
+def zip(in1: Endpoint[Unit,  A], in2: Endpoint[Unit,  B], 
         out: Endpoint[(A,B), Unit]) = 
@@
   easy to add.}  Operationally, in this example, that just means sending
-messages along \lstinline{in1} and \listinline{in2} that are marked as
+messages along \lstinline{in1} and \lstinline{in2} that are marked as
 catalyzing messages, and hence are not consumed during reaction.  The upshot
@@
 }
-\subsection{Dynamic reagents and \lstinline{cas}}
+
+\subsection{Post-commit actions}
+\label{sec:postcommit}
+
+It is often desirable for a reagent to include ``post commit actions'': code to
+be run after a reaction has successfully taken place, \emph{e.g.} for signaling
+or spawning another thread after an operation completes.  The
+\lstinline{postCommit} combinator~(\figref{combinators}) takes a function from
+\lstinline{A} to \lstinline{Unit} and produces a \lstinline{Reagent[A,A]}.  The
+idea is that the post-commit action will be recorded along with the input of
+type \lstinline{A}, which is passed along unchanged.  Once the reaction
+completes, the action will be invoked on the stored input.  The combinator is
+meant to be used in sequence with other combinators that will produce the
+appropriate input.  For example, the reagent
+\lstinline{pop >> postCommit(println)} will print the popped element from a stack,
+as soon as the pop has completed.
+
+\subsection{Low-level operations on shared state: \lstinline{read} and \lstinline{cas}}
+\label{sec:lowlevel}
+
+Although the \lstinline{upd} combinator is convenient, it is sometimes necessary
+to work with shared state with a greater degree of control.  To this end, we
+include two combinators, \lstinline{read} and
+\lstinline{cas}~(see~\figref{combinators}), for working directly on
+\lstinline{Ref} values.  Together with the \lstinline{computed} combinator
+described in~\secref{dynamic}, \lstinline{read} and \lstinline{cas} suffice to
+\emph{build} \lstinline{upd}.
+
+The \lstinline{read} combinator is straightforward: if \lstinline{r}
+has type \lstinline{Ref[A]}, then \lstinline{read(r)} has type
+\lstinline{Reagent[Unit, A]} and, when invoked, returns a snapshot of
+\lstinline{r}.
+%
+The \lstinline{cas} combinator takes a \lstinline{Ref[A]} and two \lstinline{A}
+arguments, giving the expected and updated values, respectively.  Unlike its
+counterpart for \lstinline{AtomicReference}, a \lstinline{cas} reagent does
+\emph{not} yield a boolean result.  Instead, the reagent library builds in
+handling of failed CASes, including an optimistic retry loop and exponential
+backoff.  A failure to CAS is, as discussed in \secref{choice}, a transient
+failure that causes the reagent to retry.
+
+\subsection{Constant and tentative reagents}
+\label{sec:failproof}
+
+% TODO: introduce these better
+
+The \lstinline{ret} combinator~(\figref{combinators}) always succeeds,
+immediately returning the given value.
+
+Becasue choice is left-biased, it can be used together with the remaining
+combinators to express \emph{tentative} reagents: if \lstinline{r} is a
+\lstinline{Reagent[A,B]} then \lstinline{r?} is a
+\lstinline{Reagent[A,Option[B]]} that first tries \lstinline{r} (wrapping its
+output with \lstinline{Some}) and then tries \lstinline{ret(None)}, which always
+succeeds.  This allows a reaction to be attempted, without retrying it when it
+fails.  The \lstinline{?!} method combines \lstinline{?} and \lstinline{!} into
+a single operation.
+
+\subsection{Dynamic reagents}
 \label{sec:dynamic}
@@
     read(tail) ! () match {
-      case ov@Node(_, r@Ref(null)) => 
+      case ov@Node(_, r@Ref(null)) => // found true tail
         cas(r, null, n) >>
-        postCommit { _ => cas(tail, ov, n) !? () }
-      case ov@Node(_, Ref(nv)) => 
-        cas(tail, ov, nv) !? (); 
+        postCommit { cas(tail, ov, n)? }
+      case ov@Node(_, Ref(nv)) => // not the true tail
+        cas(tail, ov, nv) ?! ();  // catch up tail ref
         findAndEnq(n)
@@
 \nocaptionrule
-\caption{The Michael-Scott queue, with reagents}
+\caption{The Michael-Scott queue, using reagents}
 \label{fig:msqueue}
@@
 Here is a brief overview of the Michael-Scott queue.  The queue is represented
-as a \emph{mutable} linked list, with a sentinel node at the head (front) of
-the queue.  The head pointer always points to the current sentinel node; nodes
-are dequeues by a CAS to this pointer, just like Treiber stacks (but lagged by
-one node).  The tail of the queue is the unique node, reachable from the head
+as a \emph{mutable} linked list, with a sentinel node at the head (front) of the
+queue.  The head pointer always points to the current sentinel node; nodes are
+dequeues by a CAS to this pointer, just like Treiber stacks (but lagged by one
+node).  The tail of the queue is the unique node, reachable from the head
 pointer, with a null \lstinline{next} pointer; thanks to the sentinel, such a
@@
 head (sentinel) and tail.  Finally, as an optimization for enqueing, a tail
-\emph{pointer} is maintained with the invariant that the real tail is always
-reachable from it.  The tail pointer may lag behind, however, which allows the
-algorithm to work using only single-word CAS instructions.
+pointer is maintained with the invariant that the real tail is always reachable
+from it.  The tail pointer may lag behind the true tail node, however, which
+allows the algorithm to work using only single-word CAS instructions.
 
@@
 The \lstinline{head} and \lstinline{tail} references of the queue are
-initialized to the same sentinel node.  Here we use a new reagent,
-\lstinline{read}~(\figref{combinators}), to extract the sentinel value from
-the head when constructing the tail.  The read reagent in \lstinline{tail}'s
-initializing expression is immediately executed during construction of an
-\lstinline{MSQueue} instance.
+initialized to the same sentinel node.  Here we use the \lstinline{read}
+combinator~(\secref{lowlevel}) to extract the sentinel value from the head
+when constructing the tail.  The read reagent in \lstinline{tail}'s initializing
+expression is immediately executed during construction of an \lstinline{MSQueue}
+instance.
 
@@
 
-Since the location of the tail node is determined dynamically by the data in
-the queue, the \lstinline{enq} reagent must itself be determined dynamically.
-
-The \lstinline{enq} reagent, on the other hand, must work with the shifting
-position of the tail, which requires new combinators.
-
-The \lstinline{cas} combinator~(\figref{combinators}) takes a
-\lstinline{Ref[A]} and two \lstinline{A} arguments, giving the expected and
-updated values, respectively.  Unlike its counterpart for
-\lstinline{AtomicReference}, a \lstinline{cas} reagent does \emph{not} yield a
-boolean result.  Instead, the reagent library builds in handling of failed
-CASes, including an optimistic retry loop and exponential backoff.  A failure
-to CAS is, as discussed in \secref{choice}, a transient failure that causes
-the reagent to retry.
-
-%TODO: treat !? as shorthand?
-
-%TODO: note tail recursion
-
-\elide{ The \lstinline{read} combinator has a straightforward signature: if
-  \lstinline{r} has type \lstinline{Ref[A]}, then \lstinline{read(r)} is a
-  \emph{value} of type \lstinline{Reagent[Unit, A]} which, when invoked, takes
-  a unit argument and yields an \lstinline{A} result.
-}
-
-This is where the differences from STM really start to show.
-
-post-commit action for catching up tail
-
-Talk about more sophisticated algorithms, e.g. skiplists
-
-\subsection{Caveats for composition}
+Since the location of the tail node is determined dynamically by the data in the
+queue, the \lstinline{enq} reagent must itself be determined dynamically.
+Dynamic reagents are created using the \lstinline{computed}
+combinator~(\figref{combinators}), which takes a partial function from
+\lstinline{A} to \lstinline{Reagent[Unit,B]} and yields a
+\lstinline{Reagent[A,B]}.  When the reagent \lstinline{computed(f)} is invoked,
+it is given an argument value of type \lstinline{A}, to which it applies the
+function \lstinline{f}.  If \lstinline{f} is not defined for that input, the
+reagent blocks, similarly to the \lstinline{upd} function.  Otherwise, the
+application of \lstinline{f} will yield another, dynamically-computed reagent,
+which is then invoked with (), the unit value.
+
+For \lstinline{enq}, we compute a dynamic reagent by first taking the given
+input \lstinline{x}, creating a node with that data, and then calling a private
+function \lstinline{findAndEnq} that will locate the tail of the queue and yield
+a reagent to update it to the new node.  Since \lstinline{findAndEnq} is
+private and tail-recursive, Scala will compile it to a loop.
+
+The \lstinline{findAndEnq} function searches for the \emph{true} tail node
+(whose \lstinline{next} field is \lstinline{null}) starting from the
+\lstinline{tail} pointer, which may lag.  To perform the search,
+\lstinline{findAndEnq} must read the \lstinline{tail} pointer, which it does
+using the \lstinline{read} combinator.  \emph{There is a subtle but important
+  point here: this read occurs while the final reagent is being computed.}  That
+means, in particular, that the read is \emph{not} part of the computed reagent;
+it is a side-effect of computing the reagent.  The distinction is important:
+such a read is effectively ``invisible'' to the outer reagent being computed,
+and thus is not guaranteed to happen atomically with it.  Invisible reads and
+writes are useful for avoiding compound atomic updates, but must be employed
+carefully to ensure that the computed reagent provides appropriate atomicity
+guarantees.  %TODO: strengthen this writing.  something about programming
+             %lock-free data structures, comparison to STM, etc
+
+Once the tail pointer has been read, its value is pattern matched to determine
+whether it points to the true tail.  If it does, \lstinline{findAndEnq} yields a
+\lstinline{cas} reagent~(\secref{lowlevel}) that will update the
+\lstinline{next} field of the tail node from \lstinline{null} to the new node.
+The attached post-commit action attempts to catch up the tail pointer through a
+\lstinline{cas}.  Since the \lstinline{cas} fails only if further nodes have
+been enqued by other concurrent threads, we perform it
+tentatively~(\secref{failproof}); it is not necessary or desireable to retry on
+failure.
+
+If, on the other hand, the \lstinline{tail} pointer is lagging,
+\lstinline{findAndEnq} performs an invisible \lstinline{cas} to update it.
+Since it may be racing with other enqueuers to catch up the tail, a failure to
+CAS is ignored here.  Regardless of the outcome of the \lstinline{cas}, the
+\lstinline{findAndEnq} function will restart from a freshly-read
+\lstinline{tail} pointer.  Notice that in this case, an entire iteration of
+\lstinline{findAndEnq} is executed with no visible impact or record on the final
+computed reagent---there is no extended redo log or compound atomic transaction.
+
+%This is where the differences from STM really start to show.
+
+\elide{ %TODO: room for this?
+Although the Michael-Scott queue is somewhat intricate, the reagent code is
+short and it offers, we hope, a fairly lucid formal description of the
+algorithm.  The combinators we employed to implement it readily scale to more
+complicated examples, such as concurrent skiplists~\cite{?} or the lazy,
+lock-free set algorithm~\cite{Heller2006}.  With any of these examples, we reap the usual benefits: a concise, composable and extensible exposition of the algorithm.
+}
+
+%TODO: can we say something brief here?
+%\subsection{Caveats for composition}
 
@@
       case (f: Failure) => 
-        if (f == Block) park() else backoff.once()
+        if (f == Block) park() else backoff.once(offer)
         if (offer.rescind) withOffer() else offer.answer
@@
 point (the \lstinline{!} method,~\secref{impl-reaction}) and continuing with the
-key combinators.
+key combinators.  For space reasons, we do not discuss the implementation of
+catalysts, which is fairly straightfoward.
 
@@
 the reagent should retry, since the reaction was halted due to active
-interference from another thread.  In slogan form, we have
+interference from another thread.  \elide{In slogan form, we have
 \begin{quotation}\noindent
@@
 \end{tabular}
-\end{quotation}
-In general, a reaction encompasses three pieces of data:
-\begin{enumerate}
-\item a list of CASes to be performed,
-\item a list of messages to be consumed, and
-\item a list of actions to be performed after committing.
-\end{enumerate}
-It thus resembles the redo log used in some STM implementations~\cite{?}.
+\end{quotation}}
+In general, a reaction encompasses three lists: the CASes to be performed, the
+messages to be consumed, and the actions to be performed after committing.  It
+thus resembles the redo log used in some STM implementations~\cite{Larus2006}.
 
@@
 the reagent and the reaction built up so far, and either completes the
-reaction, returning a result (type \listinline{B}), or fails, returning one of
+reaction, returning a result (type \lstinline{B}), or fails, returning one of
 the failure singletons (\lstinline{Block} or \lstinline{Retry}).  The
@@
 
-\subsection{Offers, communication and continuations}
+\subsection{Offers}
 \label{sec:offers}
+
+Message passing between reagents is synchronous, meaning that both reagents take
+part in a single, common reaction.  In the implementation, this works by one
+reagent placing an \emph{offer} to react in a location visible to the other.
+The reagent accepting the offer combines the accumulated reactions of both
+reagents, and attempts to commit them together.  The reagent making the offer
+either spinwaits or blocks until the offer is fulfilled; if it spinwaits, it may
+later decide to withdraw the offer.  Fulfilling the offer means, in particular,
+providing a final ``answer'' value that should be returned by the reagent that
+made the offer.  Each offer includes a status field, which is either
+\lstinline{Pending}, \lstinline{Rescinded}, or a final answer.  Hence, the
+\lstinline{Offer} class is parameterized by the answer type; a
+\lstinline{Reagent[A,B]} will use \lstinline{Offer[B]}.  When fulfilling an
+offer, a reagent CASes its status from \lstinline{Pending} to the desired final
+answer.
+
+In addition to providing a basic means of synchronization, the offer data
+structure is used to resolve external choices.  For example, the reagent
+\lstinline{swap(ep1) + swap(ep2)} may resolve its choices internally by
+fulfilling an existing offer on \lstinline{ep1} or \lstinline{ep2}; but if no
+offers are available, the reagent will post a \emph{single} offer to \emph{both}
+endpoints, allowing the choice to be resolved externally.  Reagents attempting
+to consume that offer will race to change a single, shared status field, thereby
+ensuring that such choices are resolved in a consistent way.
+
+Offers are made as part of the same \lstinline{tryReact} process that builds and
+commits reactions.  The \lstinline{offer} argument to \lstinline{tryReact} is
+non-null whenever an offer is to be made.
+
+%For reasons that will be explained in~\secref{impl-channels},
+
+\subsection{Continuations}
+
+For implementing backtracking choice and message passing, it is necessary for
+each reagent to know and have control over the reagents that are sequenced after
+it.  Thus we do not represent the reagent sequencing combinator \lstinline{>>}
+with its own class.  Instead, each reagent records its own \emph{continuation},
+which is another reagent.  Thus, for example, while the \lstinline{cas}
+combinator produces a reagent of type \lstinline{Reagent[Unit,Unit]}, the
+\lstinline{CAS} class has a parameter \lstinline{k} of type
+\lstinline{Reagent[Unit,R]}, and \lstinline{CAS} extends
+\lstinline{Reagent[Unit,R]} rather than \lstinline{Reagent[Unit,Unit]}.  The
+\lstinline{R} stands for (final) result.
+
+The combinator functions are responsible for mapping from the user-facing API,
+which does not use continuations, to the internal reagent subclasses, which do.
+Each reagent initially begins with the empty continuation, called
+\lstinline{Commit}, the behavior of which is explained in~\secref{impl-commit}.
+The sequencing combinator then merely plumbs together the continuation arguments
+of its parameters.
 
@@
 \label{sec:impl-reaction}
-%TODO: note tail recursion
 
@@
 The second retry loop, \lstinline{withOffer}, is similar, but begins by
-allocating an \lstinline{Offer} object to make visible to other reagents.
-%TODO: finish this
+allocating an \lstinline{Offer} object to make visible to other reagents.  Once
+the offer had been made, the reagent can actually block when faced with a
+permanent failure; the offer will ensure that the attempted reaction is visible
+to other reagents, which may complete it.  Blocking is performed by the
+\lstinline{park} method provided by Java's \lstinline{LockSupport} class.  On a
+transient failure, the reagent spinwaits, checking the offer's status.  In
+either case, once the reagent has finished waiting it attempts to rescind the
+offer, which will fail if another reagent has fulfilled the offer.  Because
+\lstinline{park} allows spurious wakeups, there is never a guarantee that the
+offer has been fulfilled.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{lstlisting}[frame=single]
@@
 \end{lstlisting}
-\vskip -12pt
+\vskip -21pt
 \begin{lstlisting}[frame=single]
-class CAS[A,B](ref: Ref[A], ov: A, nv: A, k: Reagent[A,B]) 
-  extends Reagent[Unit,B] {
-  def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
+class CAS[A,R](ref: Ref[A], ov: A, nv: A, k: Reagent[A,R]) 
+  extends Reagent[Unit,R] {
+  def tryReact(u: Unit, rx: Reaction, offer: Offer[R]) = 
     if (!rx.hasCAS && !k.hasCAS)
-      if (ref.cas(ov, nv))
-        k.tryReact((), rx, offer)
+      if (ref.cas(ov, nv)) k.tryReact((), rx, offer)
       else Retry
@@
 \end{lstlisting}
-\vskip -12pt
+\vskip -21pt
 \begin{lstlisting}[frame=single]
@@
     r1.tryReact(a, rx, offer) match {
-    case Retry => 
-        r2.tryReact(a, rx, offer) match {
-          case Retry => Retry
-          case Block => Retry // must still retry r1
+    case Retry => r2.tryReact(a, rx, offer) match {
+      case (_: Failutre) => Retry // must retry r1
       case ans           => ans
@@
 \end{lstlisting}
-\vskip -12pt
+\vskip -21pt
 \begin{lstlisting}[frame=single]
-class Computed[A,B](c: A => Reagent[Unit,B]) 
+class Computed[A,B](c: PartialFunction[A, Reagent[Unit,B]]) 
   extends Reagent[A,B] {
   def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
-    c(a).tryReact((), rx, offer)
+    if (c.isDefinedAt(a)) c(a).tryReact((), rx, offer)
+    else Block
+}
+\end{lstlisting}
+\vskip -21pt
+\begin{lstlisting}[frame=single]
+class Swap[A,B,R](ep: Endpoint[A,B], k: Reagent[B, R]) 
+extends Reagent[A,R] {
+  def tryReact(a: A, rx: Reaction, offer: Offer[R]) = {
+    if (offer != null) // send message if so requested
+      ep.put(new Message(a, rx, k, offer))
+    def tryFrom(cur: Cursor, failMode: Failure): Any = 
+      cur.getNext match {
+        case Some(msg, next) =>
+          msg.exchange(k).tryReact(a, rx, offer) match {
+            case Retry => tryFrom(next, Retry)
+            case Block => tryFrom(next, failMode)
+            case ans   => ans
+          }
+        case None => failMode
+      }    
+    tryFrom(ep.dual.cursor, Retry) // attempt reaction
   }
@@
 \caption{Excerpts from core reagent subclasses}
+\label{fig:impl}
 \end{figure}
@@
 As mentioned in~\secref{offers}, the initial continuation for reagents is the
-\lstinline{Commit} continuation:
+\lstinline{Commit} continuation, shown in~\figref{impl}.  The
+\lstinline{tryReact} method of \lstinline{Commit} makes the transition from
+building up a \lstinline{Reaction} object to actually committing it.
 
@@
 
-Committing a reaction requires a $k$CAS operation: $k$ compare and sets must
-be performed atomically.  This operation, which forms the basis of STM, is in
+Committing a reaction requires a $k$CAS operation: $k$ compare and sets must be
+performed atomically.  This operation, which forms the basis of STM, is in
 general expensive and not available through hardware.  There are several
-software implementations that provide various progress
-guarantees~\cite{Fraser2007,Luchangco2003,Attiya2008}, but they tend to
+software implementations that provide nonblocking progress
+guarantees~\cite{Fraser2007,Luchangco2003,Attiya2008}.  Reagents that perform a
+multiword CAS will inherit the progress properties of the chosen implementation.
+
+For our prototype implementation, we have opted to use an extremely simple,
+blocking implementation that replaces each location to be CASed with a sentinel
+value, essentially locking the location.  As the \lstinline{Reaction} object is
+assembled, locations are kept in a globally-consistent order, which avoids
+deadlock.  The advantage of this implementation, other than its simplicity, is
+that is has no impact on the performance of single-word CASes to references,
+which we expect to be the common case.  Our experimental results
+in~\secref{performance} indicate that the blocking implementation provides good
+performance---much better than STM or coarse-grained locking---but we expect
+that a more sophisticated $k$CAS would further improve the results.
 
@@
 
+The implementation of the \lstinline{cas} combinator is given by the
+\lstinline{CAS} class, shown in~\figref{impl}.  Its \lstinline{tryReact} method
+is fairly simple, but it illustrates a key optimization we have mentioned
+several times: if neither the reaction so far nor the continuation of the
+\lstinline{cas} are performing a CAS, then the entire reagent is performing a
+single CAS, and can thus attempt the CAS immediately.  This avoids the overhead
+of creating a new \lstinline{Reaction} object and employing the $k$CAS protocol,
+and it means that lock-free algorithms like \lstinline{TreiberStack} and
+\lstinline{MSQueue} behave just like their hand-written counterparts.
+%
+If, on the other hand, the reagent may perform a $k$CAS, then the current
+\lstinline{cas} is recorded into a new \lstinline{Reaction} object, which is
+passed to the continuation.
+%
+In either case, the continuation is invoked with the unit value as its argument.
+
 \subsection{Choice}
 
+The implementation of choice~(\figref{impl}) is pleasantly simple.  It attempts
+a reaction with either arm of the choice, going in left to right order.  As
+explained in~\secref{choice}, a permanent failure of choice can only result from
+a permanent failure of \emph{both} arms.  Also, note that the right arm is tried
+even if the left arm has only failed transiently.
+
 \subsection{Computed reagents}
 
+The implementation of computed reagents~(\figref{impl}) is exactly as described
+in~\secref{dynamic}: attempt to execute the stored computation \lstinline{c} on
+the argument \lstinline{a} to the reagent, and invoke the resulting reagent with
+a unit value.  If \lstinline{c} is not defined at \lstinline{a}, the computed
+reagent issues a permanent failure.  The implementation makes clear that the
+invisible reads and writes performed within the computation \lstinline{c} do not
+even have access to the \lstinline{Reaction} object, and so cannot enlarge the
+atomic update performed when it is committed.
+
 \subsection{Channels}
-
-Message passing is by far the most complicated aspect of the reagent
-implementation.
-
-\subsection{Catalysis}
+\label{sec:impl-channels}
+
+We represent each endpoint of a channel as a lock-free bag (which can itself be
+built using reagents).  The lock-freedom allows multiple reagents to interact
+with the bag in parallel; the fact that it is a bag rather than a queue trades a
+weaker ordering guarantee for increased parallelism, but any lock-free
+collection would suffice.
+
+The endpoing bags store messages, which wrap offers with additional data from
+the sender:
+\begin{lstlisting}
+case class Message[A,B,C](
+  payload: A,		 // sender's actual message data
+  senderRx: Reaction,	 // sender's checkpointed reaction
+  senderK: Reagent[B,C], // sender's reagent continuation
+  offer: Offer[C])	 // sender's offer
+\end{lstlisting}
+Each message is essentially a checkpoint of a reaction in progress, where the
+reaction is blocked until the payload (of type~\lstinline{A}) can be swapped for
+a dual payload (of type~\lstinline{B}).  Hence the stored sender continuation
+takes a \lstinline{B} for input; it returns a \lstinline{C}, which matches the
+final answer type of the sender's offer.
+
+The core implementation of \lstinline{swap} is shown in the \lstinline{Swap}
+class in~\figref{impl}.  If an offer is being made, it must be posted in a new
+message on the endpoint before any attempt is made to react with existing
+offers.  This ordering guarantees that there are no lost wakeups: each reagent
+is responsible only for those messages posted prior to it posting its own
+message.
+
+Once the offer (if any) is posted, \lstinline{tryReact} peruses messages on the
+dual endpoint using the tail-recursive loop, \lstinline{tryFrom}.  The loop
+navigates through the dual endpoint's bag using a simple cursor, which will
+reveal at least the messages present prior to the reagent's own message being
+posted to its endpoint.  If a message is found, \lstinline{tryFrom} attempts to
+complete the reaction; the \lstinline{exchange} method combines the reaction and
+continuation of the located message with those of the reagent executing it, and
+actually performs the payload swap.  If the reaction is successful, the final
+result is returned (and the result for the other reagent is separately written
+to its offer status).  If the reaction fails, \lstinline{tryFrom} continues to
+look for other messages.  If any reaction failed transiently, then the entire
+\lstinline{swap} fails transiently; this is necessary to avoid lost wakeups.
+
+%\subsection{Catalysis}
 
@@
 \\
+& \textbf{Threads}\quad (on 16-way machine) \\
+& 
+\includegraphics[scale=0.9]{graphs/reagent-based.pdf}
+\qquad\qquad
+\includegraphics[scale=0.9]{graphs/hand-built.pdf}
+\qquad\qquad
+\includegraphics[scale=0.9]{graphs/lock-based.pdf}
+\qquad\qquad
+\includegraphics[scale=0.9]{graphs/stm-based.pdf}
 \\
-& \textbf{Threads}\quad (on 16-way machine)
 \end{tabular}
+\nocaptionrule
+\caption{Benchmarking results}
+\label{fig:perf}
 \end{figure*}
@@
 
-\subsection{Methodology and experimental setup}
+%\subsection{Methodology and experimental setup}
 
@@
 microbenchmarking, with focus on contention effects and fine-grained parallel
-speedup~\cite{?}.  In addition to those basic aims, we wish to evaluate (1) the
-extent to which reagent-based algorithms can compete with their hand-built
-counterparts and (2) whether reagent composition is a plausible approach for
-scalable atomic transfers.
-
-To this end, we designed a series of benchmarks
-
-0.25$\mu$s, 2.5$\mu$s
-
-Our benchmarks were run on a 3.46Ghz Intel Xeon X5677 (Westmere) with 32GB RAM
-and 12MB of shared L3 cache.  The machine has two physical processors with
-four hyperthreaded cores each, for a total of 16 hardware threads.  L1 and L2
-caches are per-core.  The software environment includes Ubuntu 10.04.3 and
-the Hotspot JVM 6u27.
-
-\subsection{Results}
-
-Performance measurements from our initial implementation show that reagents are
-competetive with hand-built concurrent data structures, yet also provide
-scalable composed operations that are rarely provided for such data structures.
-
-\subsection{Potential improvements}
+speedup~\cite{Mellor-Crummey1991,Michael1996,Herlihy2003,Scherer2004,Hendler2004,Fraser2007,Cederman2010,Hendler2010}.
+In addition to those basic aims, we wish to evaluate (1) the extent to which
+reagent-based algorithms can compete with their hand-built counterparts and (2)
+whether reagent composition is a plausible approach for scalable atomic
+transfers.  To this end, we designed a series of benchmarks focusing on simple
+lock-free collections, where overhead from reagents is easy to gauge.  Each
+benchmark consists of $n$ threads running a loop, where in each iteration they
+apply one or more atomic operations on a shared data structure and then simulate
+a workload by spinning for a short time.  For a high contention simulation, the
+spinning lasts for 0.25$\mu$s on average, while for a low contention simulation,
+we spin for 2.5$\mu$s.
+
+In the ``PushPop'' benchmark, all of the threads alternate pushing and popping
+data to a single, shared stack.  In the ``StackTransfer'' benchmark, there are
+two shared stacks, and each thread pushes to one stack, atomically transfers an
+element from that stack to the other stack, and then pops an element from the
+second stack; the direction of movement is chosen randomly.  The ``EnqDeq'' and
+``QueueTransfer'' benchmarks are analogous, but work with queues instead.  The
+stack benchmarks compare our reagent-based \lstinline{TreiberStack} to (1) a
+hand-built Treiber stack, (2) a mutable stack protected by a single lock, and
+(3) a stack using STM.  The queue benchmarks compare our reagent-based
+\lstinline{MSQueue} to (1) a hand-built Michael-Scott queue, (2) a mutable queue
+protected by a lock, and (3) a queue using STM.  For the transfer benchmarks,
+the hand-built data structures are dropped, since they do not support atomic
+transfer; for the lock-based data structures, we acquire both locks in a fixed
+order before performing the transfer.
+
+We used the Multiverse STM, a sophisticated open-source implementation of
+Transaction Locking II~\cite{Dice2006a} which is distributed as part of the Akka
+package for Scala.  Our benchmarks were run on a 3.46Ghz Intel Xeon X5677
+(Westmere) with 32GB RAM and 12MB of shared L3 cache.  The machine has two
+physical processors with four hyperthreaded cores each, for a total of 16
+hardware threads.  L1 and L2 caches are per-core.  The software environment
+includes Ubuntu 10.04.3 and the Hotspot JVM 6u27.
+
+%\subsection{Results}
+
+The results are shown in~\figref{perf}; the x-axes show thread counts, while the
+y-axes show throughput (so larger numbers are better).  The reagent-based data
+structures perform universally better than the lock- or STM-based data
+structures.  The results show that reagents can plausibly competete with
+hand-built concurrent data structures, while providing scalable composed
+operations that are rarely provided for such data structures.
+
+%\subsection{Potential improvements}
 
@@
 Software transactional memory was originally intended ``to provide a general
-highly concurrent method for translating sequential object implementations
-into non-blocking ones''~\cite{Shavit1997}.  This ambitious goal has led to a
+highly concurrent method for translating sequential object implementations into
+non-blocking ones''~\cite{Shavit1997}.  This ambitious goal has led to a
 remarkable research literature, which has been summarized in textbook
-form~\cite{Larus2006}.  Much of the research is devoted to achieving
-scalability on multiprocessors or multicores, sometimes by relaxing
-consistency guarantees~\cite{?} or only providing obstruction-freedom rather
-than lock-freemdom~\cite{Herlihy2003}.  %Work in the area is
+form~\cite{Larus2006}.  Much of the research is devoted to achieving scalability
+on multiprocessors or multicores, sometimes by relaxing consistency
+guarantees\elide{~\cite{?}} or only providing obstruction-freedom rather than
+lock-freemdom~\cite{Herlihy2003}.  %Work in the area is
 
@@
 
+\elide{
 \section{Conclusion}
@@
 without detailed knowledge of the algorithms involved.
+}
 

### 2011-11-07 paper/paper.tex
submittable version

@@
 
+%\usepackage[compact]{titlesec}
+
 \begin{document}
@@
 \begin{abstract}
-blah blah\\
-blah blah\\
-blah blah\\
-blah blah\\
-blah blah\\
-blah blah
+Efficient communication and synchronization is crucial for fine-grained
+parallelism.  Libraries providing such features, while indispensable, are
+difficult to write, and often cannot be tailored or composed to meet the needs
+of specific users.  We introduce \emph{reagents}, a set of combinators for
+concisely expressing concurrency algorithms.  Reagents scale as well as their
+hand-coded counterparts, while providing the composability existing libraries
+lack.
 \end{abstract}
@@
 \hspace{\stretch{1}}\emph{Programs are what happens between cache
-  misses}.\footnote{Folklore; origin unknown.}
+  misses}. %{\small[origin unknown]}%\footnote{Folklore; origin unknown.}
 
@@
 \begin{lstlisting}
-val pop: Reagent[Unit, A] = upd(head) {
-  case (x::xs, ()) => (xs, x)
-}
+val pop: Reagent[Unit, A] = 
+  upd(head) { case (x::xs, ()) => (xs, x) }
 \end{lstlisting}
@@
 When \lstinline{push} or \lstinline{pop} retry, they follow an
-exponential backoff pattern with \emph{spinwaiting} for elimination
-offers, exactly as described in \secref{background}.  The backoff
-logic follows directly from the implementation of reagents; see
-\secref{implementation} for details.
+exponential backoff pattern with spinwaiting for elimination
+offers\elide{, exactly as described in \secref{background}}.  The backoff
+logic follows directly from the implementation of reagents~(\secref{implementation}).
 
@@
 
-\elide{
-def !(a: A): B = tryReact(a, empty, null) match {
-  case (_: Failure) => {
-    val backoff = new Backoff
-    def retryLoop(shouldBlock: Boolean): B = {
-      val offer = if (maySync || shouldBlock) 
-                    new Offer[B]
-                  else null
-      tryReact(a, empty, offer) match {
-        case (f: Failure) => {
-          if (f == Block && offer != null) 
-            LockSupport.park 
-          else backoff.once
-
-          // rescind offer, but check for completion
-          if (offer != null && offer.tryAbort)
-            offer.answer
-          else
-            retryLoop(f == Block)
-        }
-        case ans   => ans.asInstanceOf[B]
-      }
-    }
-    backoff.once
-    retryLoop(false)
-  }
-  case ans => ans.asInstanceOf[B]
-}
-}
-
-\begin{figure}
-\begin{lstlisting}[frame=single]
-def !(a: A): B = {
-  val backoff = new Backoff
-  def withoutOffer(): B =
-    tryReact(a, empty, null) match {
-      case Block => withOffer()
-      case Retry => 
-        backoff.once()
-        if (maySync) withOffer() else withoutOffer()
-      case ans   => ans.asInstanceOf[B]
-    }
-  def withOffer(): B = {
-    val offer = new Offer[B]
-    tryReact(a, empty, offer) match {
-      case (f: Failure) => 
-        if (f == Block) park() else backoff.once(offer)
-        if (offer.rescind) withOffer() else offer.answer
-      case ans => ans.asInstanceOf[B]
-    }
-  }
-  withoutOffer()
-}
-\end{lstlisting}
-\nocaptionrule
-\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
-\label{fig:reaction}
-\end{figure}
-
 %TODO: talk about dissolve
@@
 \lstinline{tryReact} method, which is an abstract method of
-\lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
+\lstinline{Reagent[A,B]} with the following signature:\elide{\footnote{ Functional
   programming aficionados should regard \lstinline{tryReact} as a kind of
-  Kleisli arrow for reagents.}
+  Kleisli arrow for reagents.}}
 \begin{lstlisting}
@@
 of its parameters.
+
+\begin{figure}
+\begin{lstlisting}[frame=single]
+def !(a: A): B = {
+  val backoff = new Backoff
+  def withoutOffer(): B =
+    tryReact(a, empty, null) match {
+      case Block => withOffer()
+      case Retry => 
+        backoff.once()
+        if (maySync) withOffer() else withoutOffer()
+      case ans   => ans.asInstanceOf[B]
+    }
+  def withOffer(): B = {
+    val offer = new Offer[B]
+    tryReact(a, empty, offer) match {
+      case (f: Failure) => 
+        if (f == Block) park() else backoff.once(offer)
+        if (offer.rescind) withOffer() else offer.answer
+      case ans => ans.asInstanceOf[B]
+    }
+  }
+  withoutOffer()
+}
+\end{lstlisting}
+\nocaptionrule
+\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
+\label{fig:reaction}
+\end{figure}
 
@@
 \end{lstlisting}
-\vskip -21pt
+\vskip -4pt
 \begin{lstlisting}[frame=single]
@@
 & \textbf{Threads}\quad (on 16-way machine) \\
+\\
 & 
@@
 
-Haskell's STM~\cite{Harris2005a} was inspirational in showing that
-transactions can be represented via monads~\cite{PeytonJones1993}, explicitly
-composed, and combined with blocking and choice operators.  Reagents also form
-a monad, but we have chosen an interface closer to
-\emph{arrows}~\cite{Hughes2000}, to encourage static reagent layout wherever
-possible~(\secref{dynamic}).  Like \lstinline{orElse} in Haskell's STM, our
-choice operator is left-biased.  But unlike \lstinline{orElse}, our choice
-operator will attempt the right-hand side even when the left-hand side has
-only failed transiently (rather than permanently).  While the distinction may
-appear somewhat technical, it is crucial for supporting examples like the
-elimination backoff stack~(\secref{choice}).
+Haskell's STM~\cite{Harris2005a} was inspirational in showing that transactions
+can be represented via monads~\cite{PeytonJones1993}, explicitly composed, and
+combined with blocking and choice operators.  Reagents also form a monad, but we
+have chosen an interface closer to \emph{arrows}~\cite{Hughes2000}, to encourage
+static reagent layout wherever possible~(\secref{dynamic}).  Like
+\lstinline{orElse} in Haskell's STM, our choice operator is left-biased.  But
+unlike \lstinline{orElse}, our choice operator will attempt the right-hand side
+even when the left-hand side has only failed transiently (rather than
+permanently).  While the distinction appears technical, it is crucial for
+supporting examples like the elimination backoff stack~(\secref{choice}).
 
@@
 Communicating Transactions~\cite{Lesani2011} and Transaction
-Communicators~\cite{Luchangco2011b} (the latter extends earlier work on
-Transaction Synchronizers~\cite{Luchangco2005}).  In both cases, a key
+Communicators~\cite{Luchangco2011b}\elide{ (the latter extends earlier work on
+Transaction Synchronizers~\cite{Luchangco2005})}.  In both cases, a key
 question is how the expected isolation of shared memory can safely coexist
@@
 Our mixture of message-passing and shared-state combinators most closely
-resembles the work on Communicating Transactions.  Of course, the most
-important difference is in the way we deal with shared state (and the
-guarantees provided thereby), which we discussed in~\secref{stm}.  Aside from
-that, we believe that \emph{synchronous} communication is a better choice for
-expressing patterns like elimination~(\secref{choice}), since the pattern
-relies on both participants being mutually-aware.  %TODO: say more here
+resembles the work on Communicating Transactions.  Of course, the most important
+difference is in the way we deal with shared state \elide{(and the guarantees
+  provided thereby), which we} discussed in~\secref{stm}.  We also
+believe that \emph{synchronous} communication is better for expressing
+patterns like elimination~(\secref{choice}), since they rely on 
+participants being mutually-aware.  %TODO: say more here
 

### 2011-11-07 paper/paper.tex
finalversion

@@
 concurrency patterns, like optimistic retry loops, backoff schemes, and blocking
-and signalling.  Using reagents, it is possible to express sophisticated
+and signaling.  Using reagents, it is possible to express sophisticated
 concurrency algorithms at a higher level of abstraction, while retaining the
@@
 contention.  Reagents perform universally better than the lock- and STM-based
-implementations, and are indeed competetive with hand-written lock-free
+implementations, and are indeed competitive with hand-written lock-free
 implementations.
@@
 blocked threads, depending on the changes being performed.  In some cases, it is
-possible to cleverly treat signalling in essentially the same way as atomic
+possible to cleverly treat signaling in essentially the same way as atomic
 updates~\cite{Scherer2004}.
 
-Synchronization and signalling is also subject to cache and memory bandwidth
+Synchronization and signaling is also subject to cache and memory bandwidth
 concerns, but it would take us too far afield to discuss these in depth.
@@
 block the calling thread unless or until the stack is nonempty.  Blocking and
-signalling are entirely handled by the reagent implementation; there is
+signaling are entirely handled by the reagent implementation; there is
 therefore no risk of a lost wakeup.  %TODO: add more here?
@@
 \lstinline{swap} will block until a reaction can take place.  Once
-again, the waiting and signalling protocol is managed internally by
+again, the waiting and signaling protocol is managed internally by
 the reagent implementation, so lost wakeups are impossible.
@@
 reagent placing an \emph{offer} to react in a location visible to the other.
-The reagent accepting the offer combines the accumulated reactions of both
-reagents, and attempts to commit them together.  The reagent making the offer
-either spinwaits or blocks until the offer is fulfilled; if it spinwaits, it may
-later decide to withdraw the offer.  Fulfilling the offer means, in particular,
+The reagent making the offer either spinwaits or blocks until the offer is
+fulfilled; if it spinwaits, it may later decide to withdraw the offer.  The
+reagent accepting the offer combines the accumulated reactions of both reagents,
+and attempts to commit them together. Fulfilling the offer means, in particular,
 providing a final ``answer'' value that should be returned by the reagent that
@@
 in~\secref{performance} indicate that the blocking implementation provides good
-performance---much better than STM or coarse-grained locking---but we expect
-that a more sophisticated $k$CAS would further improve the results.
+performance---much better than STM or coarse-grained locking---but a more
+sophisticated $k$CAS would likely do even better.
 
@@
 \lstinline{swap} fails transiently; this is necessary to avoid lost wakeups.
+
+The code sketch we have given for channels glosses over several details of the
+real implementation, including avoiding fulfilling one's own offer or a single
+offer multiple times, and allowing multiple interactions between two reagents
+within a single reaction.
 
